Evolutionary dynamics of insect odorant receptors reveal ecological tuning shaping olfactory perception

Version 1

Abstract

Insect olfaction is facilitated by a heterotetrameric odorant receptor–odorant receptor co-receptor (OR-Orco) complex, which is distinct from that of vertebrate ORs. However, extreme sequence divergence among insect ORs has hindered a unified understanding of their evolutionary history and ecological importance. In this study, we present a multiscale analysis of OR genes across 115 insect species. We overcome the limitations of traditional phylogenetic approaches by applying a protein similarity network-based strategy and introduce a "trunk–branch" framework to systematically describe the evolutionary trajectories of insect ORs across sequence, structural, and functional levels. Although they possess different sequences and structural communities, all the insect orders were found to contain fully functional OR repertoires. Notably, insects adapted to end-Permian mass extinction through shifts in their functional OR repertoires, and early- and late-diverging lineages exhibit distinct patterns of OR differentiation. The emergence of Orco represents a key evolutionary transition point, marking the shift from a homomeric to a heteromeric complex accompanied by specialization of the extracellular domain and binding pocket. Furthermore, we established robust associations between olfactory recognition breadth and ecological variables, including diet, circadian rhythm, and habitat. Our findings provide a comprehensive framework for the evolution of insect ORs, explaining the complex adaptive relationship between insect olfactory potential and diverse ecological environments.

Introduction

Olfactory systems play a critical role in animal survival strategies, governing critical behaviors from nutrient acquisition to mate selection (Bargmann 2006; Nei et al. 2008; Kaupp 2010). Within the kingdom Animalia, insects—the most speciose metazoan group—have evolved exceptionally sophisticated chemosensory systems to interpret complex volatile landscapes (Nei et al. 2008; Kaupp 2010; Mora et al. 2011; Robertson 2019). The extraordinary sensitivity of insect olfaction to hydrophobic volatile organic compounds (VOCs) is facilitated by a heterotetrameric channel complex consisting of odorant receptors (ORs) and odorant receptor co-receptors (Orco) (Clyne et al. 1999; Vosshall et al. 1999; Sato et al. 2008). Insect ORs represent a distinct protein superfamily that originated independently of vertebrate ORs, with no shared homology between the two (Sato et al. 2008; Hansson and Stensmyr 2011). Studies have shown that this ancient structural fold has a deeper phylogenetic origin (Benton and Himmel 2023; Himmel et al. 2023). While vertebrate ORs have been extensively studied (Niimura and Nei 2005; Hughes et al. 2018; Pacalon et al. 2023; Policarpo et al. 2024), the extremely low sequence similarity among insect ORs has made resolving their internal sequence, structural, and functional differences challenging (Robertson and Wanner 2006). Extensive studies have reported advances in the olfactory recognition mechanism and OR function (Guo et al. 2021; Chang et al. 2023; Depetris-Chauvin et al. 2023; Jiang et al. 2024), which have greatly facilitated a deeper understanding of insect olfactory perception. However, a comprehensive integrative analysis of OR sequences, structures, and functions across the entire class Insecta is still lacking, making it difficult to uncover the common principles underlying species-specific features and thereby limiting the exploration of the fundamental logic of insect olfaction.

Within the class Insecta, Orcos exhibit exceptional conservation (>400 Mya evolutionary stability) (Larsson et al. 2004; Brand et al. 2018; Robertson 2019; Wicher and Miazzi 2021), whereas lineage-specific ORs display unprecedented diversity (∼30% identity, Fig. 2B), far exceeding that of vertebrate ORs (∼80% identity, Fig. S1B and C). While recent cryo-electron microscopy breakthroughs have shed light on OR-Orco activation gating mechanisms (Wang et al. 2024; Zhao et al. 2024), the evolutionary interplay between receptor family diversification and macroecological adaptation remains a profound enigma in evolutionary biology. Traditional phylogeny-driven subfamily categorization, constrained by extreme cross-order sequence divergence, fails to capture the continuum of OR evolutionary innovation (Zhou et al. 2012; Zhou et al. 2015; Mitchell et al. 2020; Legan et al. 2021; Tian et al. 2022; Yin et al. 2022). Unlike vertebrate G protein-coupled ORs, insect ORs assemble with three nonligand-binding Orco subunits into ligand-gated ion channels, an architectural innovation accompanied by striking evolutionary contrasts (Jones et al. 2005; Nei et al. 2008; Kaupp 2010; Jones et al. 2011; Wang et al. 2024; Zhao et al. 2024). The mechanistic drivers of this functional divergence raise key evolutionary questions: What were the key factors that drove the emergence of Orco and the formation of the insect-specific OR–Orco olfactory perception model, and what advantages did it provide?

Identifying the binding profiles of ORs for VOCs is fundamental to understanding how insect olfactory systems perceive and adapt to their environments (McBride et al. 2014; Auer et al. 2020). To date, numerous studies have characterized the ligand-binding profiles of various insect ORs using heterologous expression systems (Hallem and Carlson 2006; Carey et al. 2010; de Fouchier et al. 2017; Guo et al. 2021; Chang et al. 2023; Jiang et al. 2024). However, when attempting to investigate the functional relevance of tens of thousands of ORs across the class Insecta at a macroevolutionary scale, experimental methods are no longer feasible due to their long timeframes and low throughput. Consequently, theoretical prediction has become the only viable approach. Machine learning approaches are limited by the size of available training datasets and exhibit poor scalability when applied across species on a large scale (Boyle et al. 2013; Chepurwar et al. 2019; Caballero-Vidal et al. 2020). Therefore, molecular docking remains the mainstream approach for ligand screening (Lyu et al. 2019; Comte et al. 2025). This strategy has already proven successful in mammals: by employing large-scale molecular docking and OR binding profile prediction, researchers have revealed differences in olfactory perception among bats with different diets, as well as dynamic trade-offs between vision and olfaction in primates (Zhang et al. 2024; Chi et al. 2025). These results suggest that molecular docking can facilitate the exploration of macroevolutionary patterns.

In this study, we reconstructed the evolutionary history of 450 million years of OR through a multilevel analysis of the OR superfamily of 115 insect species spanning 16 orders. Using a newly developed pipeline combining network analysis and protein structure and function prediction, we constructed a sequence similarity network (SSN), a structural similarity network (StSN), and a functional similarity network (FSN) for insect ORs and performed similarity community classification to investigate OR relationships from three distinct perspectives. The results revealed that holometabolous and nonholometabolous insects exhibit distinct OR differentiation patterns and that the composition of functional ORs in insects significantly differed before and after the end-Permian mass extinction (EPME) event. Moreover, the β-sheet structure of the extracellular loop 2 (EL2) region in Orco, along with its highly specialized ligand-binding pocket, plays a critical role in the loss of ligand-binding function and the evolutionary origin of Orco. In addition, we systematically revealed broad evidence of adaptive evolution linking the potential olfactory repertoire to different lifestyles at the scale of the class Insecta.

Results

Characterization of the insect odorant receptor gene repertoire

We compiled all published insect genomes (as of December 2022), selecting the highest-quality genome assembly (contig N50 > 1 Mb) from each family as the representative species. Our study includes 115 species, featuring 8 representative species from different insect orders with previously annotated ORs, spanning 16 orders and 111 families. This approach ensured a comprehensive analysis across the class Insecta (Fig. 1A and Data S1). To increase annotation reliability, we implemented the most stringent protein superfamily member screening strategy to date (fig. S1D). A total of 8,912 intact ORs were successfully annotated and curated (Data S2). Our analysis revealed substantial variation in OR counts across insect species. Hymenoptera exhibited the highest number of OR, with an average of 174.5 ± 110.7 (mean ± SD, N = 23) and accounted for 8 of the top 10 species with the largest OR repertoires. In contrast, Odonata had the fewest intact ORs, with a mean of 4 (N = 3). Hymenoptera also showed the widest range in OR counts (min: 38, max: 473, N = 23), followed by Coleoptera (min: 9, max: 271, N = 15) (Fig. 1B).

Evolutionary ecology of insect OR repertoires.

(A). Phylogeny of 115 insect species (see Methods and Data S4 for details). Genome size and the number of OR genes for each species are shown as bar plots. Ecological parameters, including larval and adult diets, habitat, and circadian rhythm, are displayed as labels with color coding corresponding to the legend below. Different geological periods are separated by shaded intervals. The tree was visualized using the iTOL (Letunic and Bork 2024). (B). Boxplot showing the number of OR genes for each insect order. (C). Relationship between genome size and OR gene number across 114 insect species. Each point represents one species. The gray shading indicates the 95% confidence interval. (D-E). Results of two-sided pGLS tests assessing the relationship between intact OR gene numbers and various ecological parameters: adult diet (D), larval diet (E), circadian rhythm (F), and habitat (G). Dots represent OR counts of each species. N represents the counts of species. Animal silhouettes were obtained from PhyloPic.org.

We found no significant correlation between the number of ORs and genome size across all the insect species analyzed (P = 0.0958, fig. S1E and Data S1). However, we observed that Locusta migratoria exhibited a sharp increase in genome size, primarily due to transposable element proliferation (Wang et al. 2014; Li et al. 2024). When L. migratoria was excluded from the analysis, the OR gene number and genome size were significantly negatively correlated (P < 0.01, Fig. 1C). These findings suggest that, except for the order Orthoptera, the evolutionary trajectories of genome size and OR gene repertoire size generally oppose those of insects, potentially driven by distinct environmental factors and/or life habit traits.

Previous studies have shown that in certain insect orders and specific lineages, OR repertoire characteristics are associated with various ecological behaviors (McBride 2007; Liu et al. 2021; Schrader et al. 2021; Jongepier et al. 2022; Singh et al. 2025). To determine whether these relationships extend to the class Insecta, we analyzed four ecological factors: adult diet, larval diet, habitat, and circadian rhythm (Data S3). Adult and larval diets represent stage-specific ecological traits, and our analysis of OR gene numbers at the genomic level aimed to reveal their relationship with long-term evolutionary pressures shaping OR gene family size. Our results revealed a significant correlation between insect OR counts and dietary preferences, both in adults (Fig. 1D, P = 0.61 × 10⁻⁶) and larvae (Fig. 1E, P = 0.02756). Habitat was also significantly associated with OR counts (P = 0.03984), with terrestrial insects exhibiting more ORs than aquatic insects do (Fig. 1F). However, there was no significant correlation between circadian rhythm and OR count (P = 0.3952) (Fig. 1G). Overall, our findings suggest that insect OR counts correlate with dietary specialization (in both larvae and adults) and habitat type but are not associated with circadian rhythm.

OR sequence similarity network

This study includes species representing the full phylogenetic diversity of the class Insecta, and their OR sequences exhibit substantial variation in amino acid similarity (min = 13.7%, max = 97.6%, and mean = 29.6%) (Fig. 2B). The traditional method of classifying subfamilies on the basis of bootstrap values of phylogenetic tree nodes suffers from reduced precision and accuracy. Therefore, we constructed an SSN on the basis of an "ALL vs. ALL" sequence alignment. In the SSN, each node represents a cluster of ORs sharing at least 60% sequence similarity, and edges are drawn between nodes if the average alignment E value < 1e-25 (fig. S2A). Nodes with a greater number of connections indicate closer evolutionary relationships.

Subfamily classification of ORs based on SSN.

(A). Schematic representation of the “trunk-branch” strategy pipeline for constructing the similarity network and classifying subfamilies. First, all ORs were subjected to greedy clustering based on sequence similarity to generate nodes. The average similarity between nodes was then calculated, and nodes were treated as units to identify core regions (trunks) using a strict similarity threshold. By gradually relaxing the threshold, new core regions and additional nodes (branches) emerged. After a final merging step based on defined criteria, the optimal community structure was determined using modularity as an index (see Methods). (B). Distribution of pairwise sequence similarity among all ORs. The red line indicates the mean similarity. (C). SSN of ORs constructed using the “trunk-branch” strategy. Each node represents ORs sharing >60% sequence identity. Edges represent node pairs with an average sequence alignment E-value of ≥ 1 × 10⁻²⁵. Node colors indicate SeqC classifications. (D). Heatmap depicting the number of ORs within different SeqCs across various insect orders. Due to substantial differences in OR numbers among orders, the values are ln-transformed. The bar plot above the heatmap shows the number of SeqCs present in each order, while the bar plot on the right indicates the proportion of ORs within each SeqC generated by tandem duplication. Two ORs located within 5 kb of each other on the genome were defined as tandem duplicates. TD: Tandem Duplication.

Using the "trunk–branch" strategy (Fig. 2A), we identified 38 OR sequence communities (SeqCs), which together account for 94.9% of all ORs. Among them, 22 SeqCs demonstrated clear interaction relationships (Fig. 2C and Data S5). The sequence similarity within each SeqC was predominantly between 20% and 30% (fig. S2B). By mapping previously classified Acromyrmex echinatior sequences using phylogenetic methods (Zhou et al. 2015), we found that most subfamilies—such as 9-exon, Orco, F, H, B, and A—were independently delineated (fig. S2C) and exhibited well-defined boundaries (fig. S2E). Our classification did not disrupt any known subfamilies and instead merged some families with minor differences (fig. S2, C and D). We further evaluated the SSN using a curated set of pheromone receptors (PRs) of moths (Yang et al. 2022) and found that 81.2% clustered into SeqC6, suggesting it may represent a PR clade (fig. S2F). The remaining ∼19% were scattered across other SeqCs, supporting the hypothesis of multiple independent origins of PRs (Bastin-Heline et al. 2019). These results further support the applicability and robustness of SSN-based community partitioning.

SeqC21 includes the Orco group as well as all ORs from Thermobia domestica and Machilis hrabei (fig. S2G), suggesting that ORs in early wingless insects share high sequence similarity with Orco. In addition to SeqC21, most SeqCs are order specific, indicating substantial sequence divergence among ORs from different insect orders (Fig. 2D). Lepidoptera was found to contain the greatest number of SeqCs, reflecting the highest level of OR sequence divergence. Additionally, we identified SeqC25 and SeqC27 as two exceptionally large, Hymenoptera-specific groups, comprising 39.8% and 31.9% of all Hymenopteran ORs, respectively. SeqC27 corresponds to the previously identified Hymenopteran 9-exon family (fig. S2, C and D), which originated through tandem duplication events (Engsontia et al. 2015; Zhou et al. 2015; McKenzie et al. 2016; Legan et al. 2021). Interestingly, we found that SeqC25 exhibited an even higher rate of tandem duplication than did SeqC27, suggesting that SeqC25 also expanded through extensive tandem duplication. Notably, among SeqCs with tandem duplication rates exceeding 60%, 5 of 7 are exclusive to Hymenoptera, indicating that OR lineages in this order have undergone particularly intense tandem duplication expansion compared with other insect orders.

Structural group diversity within the ORs

Insect ORs fall within the "twilight zone" (10%–40% sequence identity) of sequence alignment. However, nearly all ORs remain within the "safe zone" in structural alignment (Fig. 3A).

StSN and structural community diversity of ORs.

(A). Sequence similarity versus structural similarity of insect ORs, assessed using pairwise comparisons with Dali. Each dot represents comparison between a pair of ORs, with color indicating point density. (B). StSN of ORs constructed using the “trunk- branch” strategy. Nodes represent ORs with pairwise Dali Z-scores >50. Edges connect nodes with an average pairwise Z-score of at least 42.9. Node colors denote StrCs classifications. (C). Structural comparison of ORs and Orco. The structures were aligned using US-align, with multiple structural alignments performed for annotated Orco and OR structures (visualized in PyMol). Colors represent pLDDT scores provided by AlphaFold2, indicating the confidence of structural modeling. (D). Visualization of differences among StrCs. The heatmap shows the number of ORs within different StrC across insect orders, with OR counts ln-transformed due to large inter-order variations. The bar plot above the heatmap indicates the number of StrC in each order. On the right, a ridge plot illustrates the flexibility rates of different structural regions across StrCs. Flexibility rate = 1 - conservation rate. Light gray represents transmembrane regions, white denotes loop regions, and dark gray marks the anchor domain of ORs. Colors correspond to StrCs, with gray indicating StrCs without interactions. (E). Two types of ORs in StrC17 — long IL3 ORs and general ORs — with the IL3 regions highlighted in color. (F). Phylogenetic analysis of StrC17 ORs based on amino acid sequences. Red circles indicate bootstrap values, with most ORs collapsed based on bootstrap support. Gray branches represent the clade of long IL3 ORs, and the dN/dS values are shown in blue on the corresponding evolutionary branches.

To investigate the structural diversity of insect ORs, we constructed a StSN, where each node represents ORs with Z score > 50 (from Dali (Holm 2022)). Edges were drawn between nodes if their average Z score was > 42.9 (fig. S3A). Similar to the SSN, we classified 28 structural communities (StrCs), which together accounted for 87.0% of all ORs. Among these, 13 StrCs exhibited distinct interaction relationships (Fig. 3B and Data S5). The boundaries between StrCs were well defined (fig. S3C).

Most StrCs were also order specific (Fig. 3D). StrC1 represents the Orco group, which includes 8 T. domestica ORs. These ORs were previously identified as Orco-like genes through sequence analysis (Brand et al. 2018). Compared to ORs, Orco is characterized by two distinct structural features: a β-sheet in EL2 and a long loop in intracellular loop 2 (IL2) (Fig. 3C and fig. S3B). Lepidoptera presented the highest diversity of StrC types (Fig. 3D). We found that the transmembrane regions in different StrCs presented low variability, whereas the loop regions presented distinct variable patterns (Fig. 3D). In both ORs and Orcos, the IL2 region was identified as the most variable region and the only variable segment within the anchor domain. This high degree of variability may be associated with its role in stabilizing and regulating the anchoring domain (Bahk and Jones 2016; Jain et al. 2021; Wang et al. 2024; Zhao et al. 2024).

Notably, StrC17 exhibited a high degree of flexibility in the IL3 region. We observed that a subset of ORs within this group possessed unusually long IL3 regions (Fig. 3E). These ORs were distributed across 15 families from two suborders of Diptera, but they were absent from the earliest-diverging families in the Nematocera suborder, including Culicidae, Chironomidae, and Tipulidae (fig. S3, E and F). After manual inspection, annotation errors were ruled out. The construction of a maximum likelihood tree for StrC17 ORs revealed that ORs with long IL3 regions formed a distinct clade, with one subclade identified as orthologous (fig. S3D). To investigate the evolutionary pressures acting on these ORs, we calculated the dN/dS ratios for different branches (Fig. 3F and Data S6). The results revealed that all StrC17 ORs were subject to strong purifying selection (dN/dS = 0.17). The amino acid substitution rate of ORs with long IL3 regions was two orders of magnitude lower than that of other ORs (dN/dS = 0.001–0.004), indicating intensified purifying selection. Further analysis of functional group classifications (described later) revealed that nearly all ORs with long IL3 regions (18/19) were either narrowly tuned or had unknown functions, suggesting that these ORs may play highly conserved olfactory roles.

Origin and driving factors of Orco

Orco originated in the ancestor of Zygentoma and evolved from ORs (Brand et al. 2018; Thoma et al. 2019). The evolutionary drivers underlying Orco specialization remain unresolved. Our phylogenetic analyses revealed that early-diverging insect ORs lack direct orthologs, precluding the reconstruction of ancestral ORs (fig. S4A). The T. domestica ORs TdomOR1-8 are classified within Orco communities (SeqC21 and StrC1) and occupy a transitional phylogenetic position between ORs and characterized Orcos (fig. S4B), representing the closest ORs related to Orcos. Therefore, we propose that TdomOR1-8 were candidate receptors during the recruitment and expansion of the co-receptors in the ancestral lineage of Zygentoma. We found that TdomOrcoFJ has the smallest, least polar, and most hydrophobic binding pocket among TdomOR1-8 (Fig. 4A). Molecular docking analysis revealed that TdomOrcoFJ is less suited for VOC binding than other candidates are (fig. S4C). These findings suggest that the extreme characteristics of the binding pocket are key features that distinguish TdomOrcoFJ from other candidate Orcos, with a reduced VOC-binding capacity serving as the primary driving force in Orco determination.

Drivers promoting the early origin of Orco.

(A). Differences in EL2, IL2, and binding pocket properties between M. hrabei and T. domestica ORs. Only ORs with closed binding pockets were analyzed, and all scores were normalized. Circles and triangles represent the minimum and maximum scores after column normalization, respectively. The presence of an EL2 β-sheet is indicated with a check mark. (B). Differences in amino acid properties of the EL2 region between OR and Orco. Hydrophilic amino acids are shown in blue, neutral amino acids in green, and hydrophobic amino acids in black (visualized by WebLogo (Crooks et al. 2004)). The x-axis numbers represent amino acid positions from the multiple sequence alignment, with highly gapped positions removed. (C). MD simulation-based analysis of the movement trajectory of VOC (geranyl acetate) near Orco and OR. The red protein represents ApisOR5, and the blue protein represents ApisOrco. The green area indicates the probability of VOC remaining above the binding pocket, while the pink represents its probability of occupying other regions of the protein (see Methods). Statistical significance was evaluated using a Chi-Squared Test. The red pentagram indicates the initial position of geranyl acetate in the simulation. (D). Changes in relative calcium fluorescence intensity (ΔF/F0) in cells. Cells were stimulated with 100 μM of the OR47a agonist pentyl acetate (PA) at 50 s and 100 μM the Orco agonist VUAA1 at 180 s. el2.DmOR47a: OR47a mutant in which the EL2 region of OR47a was replaced with that of Orco. mut.DmOr47a: the EL2 region of wild-type DmOr47a contains an inserted short peptide. The total number of cells (n = 30) was derived from three independent replicate experiments. The shading indicates the 95% confidence interval. (E). Relative calcium fluorescence intensity in cells following stimulation with PA and VUAA1, respectively. n=30. Statistical significance was determined using Tukey’s HSD test. (F). Schematic illustration of the mechanism by which the β-sheet structure in the EL2 region of Orco influences VOC proximity to the binding pocket. (G). Comparative analysis of binding pocket parameters (volume, hydrophobicity score, and polarity score) between Orco and ORs across all studied insect species. Statistical significance was determined using a two-sided t-test. **** P < 0. 0001.n.s. no significant.

The reasons behind the loss of Orco function remain unclear, although its conserved binding pocket may act as a molecular sieve (Pacalon et al. 2023). We found that M. hrabei ORs lack the EL2 β-sheet but have a long IL2 loop region. Similarly, most T. domestica ORs also have a long IL2 loop region, whereas only TdomOR1-8 retains the EL2 β-sheet (Fig. 4A). This finding suggests that the long IL2 loop region was conserved from ancient ORs but gradually disappeared during subsequent evolution. In contrast, the EL2 β-sheet likely emerged during Orco evolution and was retained in all subsequent Orcos.

The formation of the EL2 β-sheet may be associated with the loss of Orco function. We observed that the EL2 region of Orco is distinctly divided into two parts: the β-sheet region and the transmembrane loop region (Fig. 4B). In the β-sheet region, the proportion of hydrophilic amino acids is 47.8%, whereas it is 14.2% in the transmembrane loop region and 18% in ORs. Overall, across all the insects, the β-sheet region of Orco EL2 presented a significantly greater proportion of hydrophilic amino acids than did the ORs (fig. S4D), facilitating its stable presence in the extracellular aqueous environment. The extracellular port is the main pathway through which ligands enter the binding pocket (Renthal and Chen 2022; Pacalon et al. 2023). Using molecular dynamics (MD) simulations to analyze VOC movement outside the membrane, we found that the presence of the β-sheet in EL2 region significantly reduced the likelihood of VOCs reaching the binding pocket (Fig. 4C, fig. S4E and Movie S1). Using Drosophila OR47a as an example, we demonstrated through calcium imaging experiments that replacing the EL2 region of OR47a with that of Orco resulted in a mutant OR47a (el2. DmOR47a) that exhibited a markedly reduced ability to bind pentyl acetate (PA) (Fig. 4, D and E and fig. S4F). Additionally, we inserted a non–β-sheet short peptide into the EL2 region of OR47a, and the results showed that its function remained unaffected (Fig. 4, D and E and fig. S4F). This finding suggests that the β-sheet acts as a spatial barrier, limiting VOC access to the binding pocket as a primary physical gate (Fig. 4F). Furthermore, we observed that Orco has a smaller and less hydrophobic binding pocket than ORs do (Fig. 4G). These characteristics are unfavorable for VOC entry into the binding pocket, representing the key chemical factors contributing to Orco’s loss of natural VOC-binding function.

Functional labels for ORs on the basis of VOC recognition similarity

Currently, molecular docking and virtual screening have been successfully applied to the high-throughput prediction of OR binding profiles in mammals, revealing the macroevolutionary trajectories of species’ olfactory recognition abilities (Zhang et al. 2024; Chi et al. 2025). To better understand the functions of all insect ORs and classify them accordingly, we collected 14,499 VOCs from 4 molecular libraries (including pheromones (Li et al. 2023), fragrances (the thegoodscentscompany.com), food odors (Chi et al. 2025), and theoretical odors (Mayhew et al. 2022)). These VOCs were categorized into 9 functional groups (Fig. 5A and Data S7). We used a comprehensive VOC library to systematically evaluate the theoretical functions of ORs, aiming to assess the olfactory potential encoded by the genome, even though some of these compounds may never be encountered by insects in nature.

Classification of insect OR functional communities and the relationship between ecological parameters and the VOC recognition capability of ORs.

(A). 2D odor space based on molecule physicochemical properties. Each dot represents an odor molecule, with different colors indicating different functional groups. (B). Relationship between hit rate and docking score, inferred using Bayesian statistics. The black curve shows the distribution of hit rates and docking scores across all OR-VOC docking results. The top plateau has a hit rate of 22%, while the bottom plateau has a hit rate of 0%. Cyan dots represent the mean hit rate ± s.e.m. for each docking score interval. The orange curve indicates the peak hit rate at a docking score of -8 kcal/mol. The gray curve represents posterior distributions from Bayesian inference, with n = 500. (C). FSN and FunC of ORs. Nodes represent ORs with pairwise PCCs > 0.6 based on docking scores. Edges represent node pairs with an average PCCs ≥ 0.48. (D). Functional labels for the recognition of different functional group small molecules by each FunC. Red indicates a tendency to recognize a higher number of odor molecules from a particular functional group, while blue indicates to recognize fewer. Circle size represents the degree of this tendency. (E). Correlation between four ecological parameters and species’ BBI for different VOC functional groups. pGLS tests were used for significance assessment. Red circles represent significant correlations, with size indicating the degree of significance. Gray triangles denote non-significant results. (F). Distribution of BBI values for different VOC functional groups across species with different larval diets. (G). BBI values for amine VOCs recognized by species with different larval diets. (H). 2D distribution of BBI values across species with different life habit combinations. The density distributions of different clusters along PCA1 and PCA2 are shown at the top and right. (I). Summed BBI values for all functional groups across three life habit combinations. One-way ANOVA followed by Bonferroni post-hoc tests was used for significance analysis. (J). Binary logistic regression coefficients for different life habit combinations. ****P < 0.0001

We performed molecular docking for all insect ORs using these VOCs to generate their recognition profiles (fig. S5A and Data S8). A long-standing challenge in molecular docking is how to predict binding likelihood on the basis of ranking scores. To address this issue, we modeled a theoretical "hit rate" curve using docking scores and experimental datasets from Drosophila (Hallem and Carlson 2006), mosquitoes (Carey et al. 2010), and moths (de Fouchier et al. 2017) (Fig. 5B). We used the prior probability distribution in Bayesian statistics and determined the effective docking score range for insect ORs. A threshold score of -8 kcal mol-1 was established to maximize data retention while maintaining a reliable hit rate (Fig. 5B).

We constructed an FSN for ORs (Fig. 5C). In this network, each node represents ORs with a recognition profile Pearson correlation coefficient (PCC) greater than 0.6, and nodes with an average PCC greater than 0.48 are connected by edges. We identified 6 functional communities (FunCs), representing 72.5% of the ORs (fig. S5B and Data S5). We subsequently quantified the number of VOCs with different functional groups recognized by each FunC (fig. S5C) and assigned theoretical functional labels on the basis of their significant recognition compared with the background (Fig. 5D). The labels represent the breadth of VOC functional group recognition by each FunC. To evaluate the accuracy of these functional labels, we calculated the area under the curve (AUC) by comparing the OR experimental data with the theoretical functional labels. For most functional group predictions, an AUC exceeding 0.75 was achieved, indicating strong agreement (fig. S5D).

FunC1 showed broad recognition of almost all functional group VOCs, whereas FunC6 had a limited recognition range, potentially indicating the existence of broad-tuned and narrow-tuned ORs (Fig. 5D). Additionally, FunC3 primarily recognized aromatics, whereas FunC5 specifically recognized nitriles. Overall, ORs had the smallest recognition range for acids and amines. Previous studies have suggested that ionotropic receptors (IRs) primarily recognize acids and amines, reducing the selective pressure on ORs to detect these VOCs (Silbering et al. 2011).

Unlike SeqC and StrC, almost all insect orders contain all FunCs without exhibiting specificity (fig. S5E). The broad-tuned FunC1 dominates in most orders, particularly in Psocodea, Diptera, and Lepidoptera (fig. S5F). Additionally, all the insect species presented a minimal proportion of FunC6, indicating that a small number of narrow-tuned ORs are present across all the insects. These ORs are likely specialized to recognize specific VOCs, increasing the ability of a species to adapt to unique environmental conditions (Hallem and Carlson 2006; Carey et al. 2010; Zhang et al. 2024; Chi et al. 2025) (fig. S5F).

Relationship between OR function and insect life habits

To quantify OR promiscuity in recognizing VOCs, we developed the binding breadth index (BBI), which is based on FunC labels. This metric assigns positive values to ORs that bind to a greater variety of VOCs (higher BBI = broader olfactory breadth) and negative values to those that bind to limited VOCs (lower BBI = narrower olfactory breadth). Using this index, we generated olfactory functional barcodes for different life habits in insect orders (fig. S6A), facilitating the comparison of the breadth of VOC recognition across various life habits. It should be noted that the olfactory breadth referred to in the subsequent analyses represents the species’ potential olfactory capacity, as reflected by the full repertoire of OR genes encoded in its genome, rather than the olfactory perception exhibited at a specific developmental stage of the insect.

Phylogenetic generalized least squares (pGLS) analysis revealed differential correlations between ecological traits and VOC class BBI patterns. Larval diet was correlated with the BBI (all except nitriles), whereas adult diet was correlated exclusively with aromatics (P= 0.038) (Fig. 5E). Habitat type was associated with nitrile/ketone perception (P= 0.0017/P= 0.003), whereas circadian rhythm was not significantly linked (Fig. 5E). Building on earlier results that the OR number was associated with insect diet (Fig. 1, D and E), suggesting that larval diet significantly influences both the number and the species’ potential olfactory capacity. In contrast, the circadian rhythm had no apparent effect on either OR number or potential olfactory capacity. Insects with saprophagous larvae ranked first in the BBI for most VOC functional groups, indicating a broader VOC recognition capability than that of insects with other larval diets. In contrast, in insects with hematophagous larvae, the BBIs for all VOCs except nitriles were negative, suggesting that the ORs of these insects tend to bind to fewer VOCs (Fig. 5F). Additionally, insects with saprophagous larvae were the only life habit type with a positive BBI for amines (Fig. 5F, g and fig. S6B). This divergence of potential olfactory capacity correlated with expanded broad-tuned ORs (FunC1) in insects with saprophagous larvae (fig. S6C). Although single-parameter ecological classification revealed broad correlations between niche variables and VOC-specific BBI patterns, it failed to resolve combinatorial trait interactions shaping species-level olfactory plasticity.

To resolve combinatorial ecological effects, we defined 3 categories of potential olfactory capacity based on life habit combinations: low (Cluster 0, n=5), moderate (Cluster 1, n=23), and high (Cluster 2, n=12) responsiveness (Fig. 5, H and I). Logistic regression identified 11 key types of life habits, including enhancers (terrestrial, xylophagous and saprovorous in larvae, omnivorous and nectarivorous in adults) and suppressors (nocturnal, hematophagic, larval herbivorous, adult predatory) of potential olfactory capacity (Fig. 5J). These results support the previously established close relationship between insect ORs function and ecological niches (McBride et al. 2014; Auer et al. 2020), and further provide a systematic and expanded explanation of the potential association between the olfactory breadth of insects and diverse ecological lifestyles.

Different evolutionary patterns of insect ORs across geological periods

Although the sequence evolution of the insect OR gene family and functional diversity among different orders and species have been reported multiple times (Brand and Ramirez 2017; McKenzie and Kronauer 2018; Legan et al. 2021; Chang et al. 2023), these studies reveal only a partial view rather than the overall evolutionary landscape of insect OR sequences, structures, and functions. To elucidate the macroevolutionary trajectories of insect ORs, we integrated SeqCs, StrCs, and FunCs. The results revealed strong congruence between SeqCs and StrCs, with both communities including nearly all FunCs (Fig. 6 and fig. S7). This finding indicates that the evolution of OR functions follows a fundamentally different pattern from that of OR sequences and structures.

Relationships between OR sequence structure and functional communities across insect orders.

The phylogenetic tree follows the structure from Fig.1. On the right, the mapping relationships between all insect OR SeqCs StrCs and FunCs are shown, considering only ORs classified into communities. Time nodes for different geological periods are based on previous study (Misof et al. 2014) and are separated by gray shading. The timeline beneath the tree marks the origin of various insect orders. Animal silhouettes were obtained from PhyloPic.org.

Additionally, we found that holometabolous and nonholometabolous insects presented distinct OR differentiation patterns (fig. S8). Among ORs with FunC labels, nonholometabolous insects have fewer SeqCs, whereas holometabolous insects display greater SeqC diversity. The number of SeqCs increases progressively with increasing degree of insect order divergence, suggesting a gradual accumulation of sequence variation during insect evolution (fig. S9A). Despite the limited SeqC diversity in nonholometabolous insects, all six types of FunC were retained, indicating that functional community types were established early in insect evolution.

The EPME was a significant geological event that resulted in the extinction of most species and caused drastic environmental changes (Close et al. 2020; Jouault et al. 2022). We detected significant differences in the number of insect OR FunCs before and after the EPME event (fig. S9B). Post-EPME insects presented a greater proportion of FunC1 (broad-tuned ORs) and a lower proportion of FunC2 (fig. S9C). This finding was confirmed by data showing that the proportion of broad-tuned ORs in Locusta is much smaller than that in Drosophila and Anopheles (Chang et al. 2023). These findings suggest that post-EPME insects are characterized by more broad-tuned ORs, which are predominantly transformed from other tuned ORs.

Discussion

In this study, we developed a unified sequence–structure–function classification system to discover insect OR evolutionary trajectories. Our results revealed the mechanism of insect olfactory perception underlying macroenvironmental influences. Next, we comprehensively described the evolutionary process through which Orco emerged within the insect “conserved chassis–diversified sensors” perception model and identified the key driving factors underlying its origin from ORs. Finally, we found broad evidence at the scale of the class Insecta that adaptive evolution links species’ lifestyles to the breadth of their potential olfactory capacity.

The drivers of genome size variation remain a topic of ongoing debate (Abdel-Haleem 2007; Canapa et al. 2015). Previous studies have shown that genome size is largely determined by non-coding DNA, and its relationship with the number of protein-coding genes and organismal complexity is limited (the C-value enigma) (Gregory 2005a; Gregory 2005b; Elliott and Gregory 2015). Therefore, from a broad perspective, genome size does not directly determine the number of OR genes, but instead reflects a complex evolutionary process. The significant negative correlation observed between genome size and OR number may result from selective pressures imposed by the environment on insect evolutionary strategies, in which the forces driving genome expansion (e.g., transposable element proliferation) and those driving OR gene expansion act in opposite directions across the insect clade.

Despite differences in OR sequences and structural communities, all maintain a complete set of functional clusters. The results demonstrate stronger evolutionary coordination between OR sequences and structural types than between these features and functional types. Specifically, structural variations closely reflect sequence divergence, whereas functional distinctions show limited correlation with structural configurations. This represents a distinctive evolutionary characteristic unique to insect odorant receptor families, which allows all insect orders to contain full functional OR repertoires, despite possessing different sequences and structural communities. Furthermore, major geological events influenced the composition of insect-tuned ORs. Insects that evolved after the EPME event had a greater proportion of broad-tuned ORs, likely influenced by multiple factors. During the Permian–Triassic transition, gymnosperms (such as cycads, lycophytes, and conifers) gradually replaced the previously dominant ferns (Grauvogel-Stamm and Ash 2005; Vajda et al. 2020; Lu et al. 2021). Plant diversification played a key role in driving insect diversification, and herbivorous insects held a central position in the Permo–Triassic interaction network (Jouault et al. 2022). The shift in plant composition led to significant changes in insect food sources and host plants. Additionally, EPME led to the extinction of 90–95% of species (Benton 2016) and was accompanied by continental reconfiguration, such as the breakup of Pangaea into Laurasia and Gondwana (Zaffos et al. 2017), resulting in numerous vacant ecological niches and driving insects to undergo adaptive radiation into new habitats. During the Triassic period, the plant community shifted (Benton 2016; Lu et al. 2021; Jouault et al. 2022), and the origin and rise of mammals (Luo 2007) and dinosaurs (Marsicano et al. 2016) led to a more complex and diverse environment. Increasing the proportion of broad-tuned ORs may increase the ability of insects to recognize a wide range of plant and animal VOCs on the basis of a limited number of OR types, thus providing a greater survival advantage for insects foraging and searching for habitats to achieve rapid occupation of new ecological niches.

Here, we systematically outline the evolutionary origin of the insect OR–Orco olfactory perception model. The ancestral insect recruited GRs to perform olfactory functions, giving rise to the earliest ORs (Robertson et al. 2003; Frank et al. 2024). These early ORs retained the homologous tetrameric pattern of GRs. However, the sensitivity requirements of olfaction and gustation differ (de Fouchier et al. 2017; Pask et al. 2017; Slone et al. 2017; Dweck and Carlson 2023; Wang et al. 2024; Zhao et al. 2024). This issue appears to have been first address in the ancestor of Zygentoma by reducing the number of functional subunits while maintaining the tetrameric structure, thereby increasing sensitivity (Brand et al. 2018; Thoma et al. 2019; Wang et al. 2024; Zhao et al. 2024). We propose that the transition from OR to Orco involved duplication and selection. The number of ORs in T. domestica is eight times greater than that in M. hrabei, with no direct orthologous relationships, suggesting that OR duplication occurred in the ancestor of T. domestica. During Orco selection, at least two essential features were preserved: the β-sheet structure in the EL2 region and the extreme-property binding pocket. The β-sheet physically prevents VOCs from accessing the binding pocket, whereas the chemically restrictive nature of the binding pocket prevents VOCs from penetrating deeply. Orco exhibits strong sequence conservation, ensuring that these chemical and physical barriers are effectively preserved. Sequence similarity is much lower in the insect OR family than in the GR family (fig. S1A and C), which may be related to the emergence of Orco. The conserved and stable basal structure of Orco (as a conserved chassis) reduces structural constraints during OR (as a diversified sensor) evolution, allowing ORs to evolve at a faster rate. This strategy represents a unique evolutionary mechanism in the olfactory perception model, where structural conservation of chassis on the basis of Orco enables OR diversification to adapt to complex environments. However, we also identified some ORs with β-sheets in the EL2 region, which may be remnants of early Orco candidates. As insects continued to evolve, the proportion of β-sheet-containing ORs within OR repertoires gradually decreased (fig. S4G and H), suggesting that OR evolution is a process of continuous renewal and elimination.

Insect olfaction has undergone evident adaptive evolution in response to lifestyle demands. We found that only larval feeding habits were generally correlated with the theoretical breadth of insect olfactory recognition (Fig. 5E), suggesting that, at a macroevolutionary scale, the classification of larval diets plays an important role in shaping the evolutionary trajectory of potential olfactory capacity encoded in the genome. However, this does not imply that the olfactory ability or biological importance of larvae is greater than that of adults. The actual olfactory function at any developmental stage depends on the specific types and expression levels of OR genes, which constitute only a subset of the species’ overall olfactory potential. This is consistent with common sense of biology, as a central objective of insect evolution is to ensure the successful survival of larvae. The larval stage constitutes a significant portion of an insect’s life cycle and is crucial for nutrient accumulation (Boggs 2009; Truman 2019). For example, insects with saprophagous larvae exhibit broader recognition of amines. Since the odor profile of humus is highly complex, broad-tuned ORs are required to detect a wide range of VOCs, including amines that are rarely found in other food sources (Hamana and Matsuzaki 1984; Hussain et al. 2013). This indicates that larval saprophagy may have driven the evolutionary shift of species-level olfactory potential toward the recognition of a broader range of VOCs. However, since behaviors such as host recognition and oviposition site selection are carried out by adults, the functional expression of this olfactory potential is likely to occur primarily at the adult stage. Regardless of the developmental stage, we found that hematophagy negatively impacts olfactory potential, likely due to highly specialized diets. Additionally, predatory behavior in adults negatively affects olfactory potential, whereas nectar feeding has a positive effect. Owing to the trade-off between vision and olfaction, predatory insects, such as mantises and dragonflies, have developed excellent vision at the cost of reduced olfactory potential (Lancer et al. 2020; Yuan et al. 2022; Kittelmann and McGregor 2024). These insects rely on strong dynamic tracking abilities, which olfaction alone cannot effectively achieve. In contrast, flowers are stationary and emit distinct, easily identifiable scent profiles, allowing nectar-feeding insects to locate their food sources precisely via olfaction.

In addition to diet, terrestriality has a positive effect on the olfactory breadth of insects. Aquatic insect larvae live in water, where they encounter far fewer VOCs than terrestrial larvae do. Moreover, many aquatic insects do not feed during their adult stage (Data S3), suggesting that their olfactory function may be more related to reproduction rather than foraging. It is widely believed that diurnal species rely primarily on visual cues, whereas nocturnal species prefer to invest in olfaction (Kaskan et al. 2005; Balkenius et al. 2006; Campi and Krubitzer 2010; Montgomery and Ott 2015). However, we found that nocturnality has a slight negative effect on the olfactory breadth of insects. This finding does not contradict previous conclusions but rather relates to the trade-off between sensory sensitivity (information collection) and resolution (information processing) (Stockl et al. 2016a). Although diurnal moths generally have fewer ommatidia than nocturnal moths do, they possess more visual neuropils, which enhances resolution at the cost of sensitivity (Stockl et al. 2016a; Stockl et al. 2016b). Similarly, compared with diurnal insects, nocturnal insects allocate a greater proportion of their brain to calyces for olfactory processing (Stockl et al. 2016a). As a corresponding trade-off, their olfactory breadth is slightly reduced. In addition, this phenomenon is related to the number of VOC types. High daytime temperatures cause VOCs to be volatile, and a wide range of VOCs needs to be recognized during olfaction, whereas the opposite is true at night (Menchaca-Torre et al. 2015; Monard et al. 2021). Overall, we systematically explored the relationship between the sequence structure and function of insect ORs, elucidated for the first time, to our knowledge, revealed how insect Orco originated from ORs and identified the key driving factors, and comprehensively explained its adaptive mechanisms to the macroenvironment and lifestyle. This study provides new insights into the ecological factors driving OR evolution at the class scale. With the improvement of genome assemblies, our dataset and similarity-based community classification strategy will serve as valuable resources for future insect OR studies.

Materials and methods

Genome data

We compiled data from all published insect genomes available in the NCBI public database as of December 2022. Genomes with a contig N50 > 1 Mb were considered reliable for OR annotation. When multiple high-quality genomes were available within a single insect family, we selected the genome with the highest assembly quality as the representative. After filtering, 107 insect genomes were included: Odonata (2), Plecoptera (2), Blattodea (1), Hemiptera (6), Psocodea (1), Hymenoptera (22), Megaloptera (1), Neuroptera (1), Coleoptera (14), Trichoptera (7), Lepidoptera (30), and Diptera (20). Detailed genome information is provided in the Data S1. To maximize phylogenetic diversity, we added eight representative species from different orders with annotated ORs (Machilis hrabei (Brand et al. 2018), Thermobia domestica (Brand et al. 2018), Ladona fulva (Brand et al. 2018), Ephemera danica (Brand et al. 2018), Locusta migratoria (Wang et al. 2015), Acyrthosiphon pisum (Robertson et al. 2019), Acromyrmex echinatior (Zhou et al. 2015), and Tribolium castaneum(RRID:SCR_010440) (Engsontia et al. 2008)). In total, our study includes 115 insect species, covering 16 orders and 111 families.

Species Phylogenies

To construct the phylogenetic tree used in this study, we merged existing phylogenies across different insect orders (Misof et al. 2014). For relationships among orders within Insecta, we referred to previously published phylogenetic frameworks. To ensure broad taxonomic representation, we included 16 insect orders in our analysis. Intra-order relationships were obtained from insectphylo.org and other published species trees covering various insect taxa (Chesters 2017). For species with available genomes that were not included in existing phylogenies, we inferred their phylogenetic placements based on family-level information. Divergence times among orders were obtained from TimeTree.org (Kumar et al. 2022) and calibrated using MCMCtree (Rannala and Yang 2007). The resulting phylogeny includes 115 insect species spanning 16 orders.

Ecological Trait Data of Insect Species

Given the vast diversity and ecological complexity of insects, and the lack of a comprehensive authoritative database, we manually retrieved ecological trait data from published literature for each species. We characterized insect ecology based on four aspects: larval diet, adult diet, habitat, and circadian rhythm. For dietary traits, we classified diets into 12 types: herbivorous, xylophagous, hematophagy, mycophagous, non-feeding, omnivorous, predatory, saprovorous, frugivorous, nectarivorous, detritivorous, and parasitic. For habitat, following previous classification criteria, species were assigned as either terrestrial or aquatic (Dijkstra et al. 2014). Insects that live entirely or have at least one life stage in water were considered aquatic; those that are strictly land-dwelling throughout their life cycle were considered terrestrial. Circadian rhythm was classified based on adult activity periods as either diurnal or nocturnal. The ecological classification and corresponding literature sources for each species are provided in the Data S3.

OR gene mining and structural filtering

All OR and Orco protein sequences were downloaded from NCBI, supplemented with ORs reported in the paper, yielding a total of 16,176 ORs. Redundancy and data cleaning were then performed using CD-HIT (Li and Godzik 2006; Fu et al. 2012) v4.8.1 to remove highly similar sequences with >90% identity, resulting in 6,029 non-redundant ORs used as queries. We used Exonerate (Slater and Birney 2005) v2.4.0 to search whole-genome data and identify potential exon regions. To improve annotation efficiency, genome contigs were split into 20 Mb segments, with 20 kb flanks added at breakpoints to avoid truncating ORs. Next, InsectOR (Karpe et al. 2021) was used for OR gene annotation.

Given that multi-exon gene annotation from genome can be error-prone, we applied the strictest structural filtering strategy to enhance annotation reliability (Fig.S1d). All annotated ORs were modeled using AlphaFold2 (Jumper et al. 2021). Structural alignment against the cryo-EM structure of MhOR5 (PDB: 7LIC) was performed using Dali (Holm 2022), and ORs with Z-scores <20 were excluded. We further manually removed sequences lacking at least seven transmembrane domains.

In total, 8,912 ORs with complete and credible structural models were retained for downstream analyses. This rigorous approach eliminates structurally compromised ORs that may appear intact at the sequence level, thereby increasing the likelihood of identifying ORs with functional physiological and biochemical properties.

Phylogenetic comparative analyses

The function pGLS of the R package caper was used to perform all phylogenetic generalized linear models presented in this study. Phylogenetic manipulations were conducted using the R packages phytools (RRID:SCR_015502) and ape. Phylogenetic trees were visualized by iTOL (Letunic and Bork 2024) (https://itol.embl.de/). Animal silhouettes used in this study were retrieved from http://phylopic.org/.

"Trunk-Branch" strategy

Most existing insect ORs have arisen through gene duplication events (McKenzie and Kronauer 2018; Legan et al. 2021). Initially, these duplicated OR sequences were highly similar, forming the “trunk” of core communities. Over time, mutations led ORs to diverge from their original sequences and functional features, forming the “branches” of the community (Fig. 2A). Based on this “trunk-branch” concept, we developed a strategy to construct similarity networks and define communities and the specific implementation process is as follows:

Communities with more than 20 nodes in the initial-threshold network are defined as core communities (trunks). The initial thresholds were set at e-value = 1e-68 for the SSN and Z-score = 45.9 for the StSN.

Thresholds were gradually relaxed to extend the branches of each core community. Newly appearing nodes were assigned to the community with which they shared the highest number of edges. The SSN threshold was relaxed in logarithmic steps (e.g., 1e-68, 1e-67, …), while the StSN threshold was relaxed in steps of 0.1 Z-score (e.g., 45.9, 45.8, …).

Communities were merged based on their inter-community connectivity. If the proportion of edges connecting two communities exceeded 50% for either community, they were merged. This process was repeated until all inter-community connectivities fell below the 50% threshold.

New communities with more than 20 unassigned nodes were identified and included.

Steps 2–4 was repeated until the network reached a stage of edge explosion and massive community merging.

Previously, final visualization thresholds for sequence similarity networks were often chosen based on known functional annotations. However, such information is lacking for most ORs. We found that under different thresholds, the modularity of community structures defined by the “trunk-branch” strategy showed a sharp inflection point when compared with theoretical modularity derived from the Louvain algorithm. Here, we propose a quantifiable approach for determining the final community display threshold by identifying the inflection point. Validation against known data confirmed that the resulting community partitions are highly reliable.

Sequence similarity network

First, we performed pairwise sequence alignment of all ORs using MMseqs2 (Steinegger and Soding 2017) to obtain sequence similarities and e-values. To improve network readability, we applied pre-clustering with CD-HIT (Li and Godzik 2006; Fu et al. 2012) v4.8.1 (‘-c 0.6 -n 4’) to reduce redundant edges between highly similar nodes, resulting in a total of 4,947 nodes. The edge weight between nodes was defined as the average e-value of all OR pairs they represent. We then constructed the sequence similarity network based on the “trunk-branch” strategy (see "Trunk-Branch Strategy"). An e-value threshold of 1e-25 was ultimately selected as the cutoff for defining the final sequence similarity network and OR subfamilies. At this threshold, 38 OR subfamilies were identified, comprising 8,457 ORs. The network was visualized using the organic layout algorithm in Cytoscape (Shannon et al. 2003).

Structure similarity network

We performed structural modeling of all ORs using AlphaFold2 (Jumper et al. 2021). Pairwise structural comparisons were then conducted using Dali (Holm 2022) to obtain Z-scores. Following an approach similar to that used in SSN, we applied a greedy clustering algorithm to pre-cluster structurally similar ORs at a Z-score threshold of 50, resulting in 3,854 nodes. The edge weight between nodes was defined as the average Z-score of all OR pairs they represent. We then constructed the structure similarity network using the “trunk-branch” strategy. A Z-score threshold of 42.9 was selected as the final cutoff for defining the StSN and OR structural subfamilies. At this threshold, 28 OR structural communities were identified, comprising 7,750 ORs. The network was visualized using the organic layout algorithm in Cytoscape (Shannon et al. 2003).

Phylogenetic tree and selection pressure analysis

To infer the amino acid phylogeny of ORs within StrC17, we performed multiple sequence alignment using MAFFT (Katoh and Standley 2013) v7.505 with the parameters ‘--globalpair --maxiterate 1000’. The alignments were trimmed using trimAl (Capella-Gutierrez et al. 2009) v1.4.rev22 with the parameter ‘-automated1’. A maximum likelihood tree was then constructed using IQ-TREE (Minh et al. 2013; Nguyen et al. 2015) v 2.2.0 with the parameters ‘-m MFP -B 1000 --bnnì. To estimate dN/dS ratios, we used the codeml program from the PAML package (RRID:SCR_014932) (Yang 1997; Yang 2007). The codeml analysis employed the branch model with the control file parameters set to ‘model = 2’, ‘NSsites = 0’. We used a custom script to extract log-likelihood values (lnL) from the output files and perform likelihood ratio tests (LRTs).

Evaluation of OR binding pocket properties

We used fpocket to identify OR binding pockets and assess their properties, with parameters set to ‘-m 3.1 -i 10’. To locate the centroid region of the binding pocket, we aligned each OR structure to MhOR5 using USalign (Zhang et al. 2022) v 20220924. Among all candidate pockets identified by fpocket (Le Guilloux et al. 2009) 4.0, the one closest to the centroid was designated as the binding pocket. Due to the stochastic nature of OR modeling, some ORs exhibited open binding pockets, making it difficult to define cavity-related parameters. To ensure consistency in the quantitative analysis, we only included ORs with closed binding pockets. Specifically, we retained only binding pockets with Number of Alpha Spheres < 140 and Volume < 800, excluding any that were open to the exterior.

Molecular dynamics simulation of Orco β-Sheet structural function

We used GROMACS (Abraham et al. 2015) to simulate the interaction between the EL2 regions of OR and Orco with VOCs. The proteins used were the cryo-EM structures of ApOrco and ApOR5 (Wang et al. 2024) (PDB: 8z9z), and the selected VOC was geranyl acetate (CID: 1549026). The membrane system was rapidly constructed using the QwikMD plugin in VMD (Humphrey et al. 1996). After default minimization and annealing steps, an optimized membrane-protein system was obtained. The Caver (Jurcik et al. 2018) tool was used to identify the pocket tunnels in the cryo-EM structures of ApisOR5 and ApisOrco. Surface residues were identified using FindSurfaceResidues(https://pymolwiki.org/index.php/FindSurfaceResidues), helping define the residues surrounding the binding pocket. Geranyl acetate was positioned above the pocket, and 5 ns MD simulations were performed to evaluate interactions.

A binding state was considered potential if the following criteria were met:

- The minimum distance between geranyl acetate atoms and pocket residue atoms was <12 Å (cutoff).

- The spatial distribution of the ligand satisfied the following exclusion conditions:

- Not located on the lateral side of the protein

- Not embedded in the phospholipid bilayer

- Not sterically blocked by the EL2 of ApisOrco

Each simulation was repeated 60 times for both ApOrco and ApOR5.

Heterologous expression and calcium imaging

pDmelOR-mRho.V5.mER.hOr47a (RRID:Addgene_126478), a high-copy number bidirectional expression vector by inserting D. melanogaster Or47a (DmOr47a) and Orco in the pCMV-BI vector (RRID:Addgene_126475) backbone (Miazzi et al. 2019), was a gift from Bill Hansson (Addgene plasmid # 126478), then V5 tag was replaced with Flag tag, named as pDmelOR-OR47a. The EL2 of OR47a (GHAEPELPFPCLFPWNIHI) in pDmelOR-OR47a was replaced by the EL2 of Orco (GDSVKMVVDHETNSSIPVEIPRLPIKSFYPWNASHG) to constructed the new plasmid of pDmelOR-el2.OR47a. The sequences were confirmed by Sanger sequencing (Sangon Biotech, China).

HEK293T cells (RRID:CVCL_0063) were purchased from Wuhan Procell Biotechnology and maintained in DMEM medium (BasalMedia, China) supplemented with 10% fetal bovine serum (ExPhos, China) and 1% penicillin−streptomycin mixture (Sangon Biotech, China) at 37 °C with 5% CO2 until reaching 70–80% confluency. Cells were transfected with plasmids using LipoFiter transfection reagent (Hanbio, China). 6-8h after transfection, transfected cells were seeded at a density of 6 × 105 cells/well into a 35 mm glass-bottom culture dish (NEST, China). The HEK293T(RRID:CVCL_0063) cell line used in this study is derived from a male human fetus. HEK293T (RRID:CVCL_0063) cells were authenticated by STR profiling (Wuhan Procell) and tested negative for mycoplasma contamination every 3 months.

Cells were harvested for calcium imaging approximately 36 hours post-transfection. The culture medium was removed, and cells were rinsed with HBSS solution (BasalMedia, China). Subsequently, cells were incubated with Fluo-8 AM (5 μM, AAT Bioquest, USA) diluted in HBSS + 0.04% pluronic acid (F127, Sangon Biotech, China) in a CO2 cell culture incubator for 30 minutes at 37°C. Following three washes with HBSS solution, the Ca2+ dependent fluorescence was recorded with a fluorescence microscope (DeltaVision Ultra, GE) with the emission wavelength set at 525 nm and the excitation wavelength at 490 nm. Experiments lasted 5 min with a sampling interval of 5 s. The stimuli—pentyl acetate (PA, 100 μM; Aladdin, China), an OR47a agonist, and VUAA1 (100 μM; Sigma-Aldrich, USA), an Orco agonist—were gently added to the culture dish at 50 s and 180 s, respectively, after the start of the imaging protocol. The recorded signals were analyzed by Fiji (RRID:SCR_002285) ImageJ v.1.54 (RRID:SCR_003070) (Schindelin et al. 2012). Levels of Ca2+ were expressed as ΔF/F0 ratios. The average fluorescence intensity of the cells during the first 30 seconds was recorded as the baseline (F0). Each assay was conducted in triplicate to ensure reproducibility. For calcium imaging experiments, data acquisition and initial AF/F0 ratio calculations were performed by researchers who were blinded to the treatment groups during the analysis phase. Sample sizes for calcium imaging (n = 30) were determined based on common standards in the field and prior pilot experiments to ensure sufficient statistical power for detecting changes in fluorescence intensity

Virtual docking data of ORs and VOCs

We constructed an insect-relevant odorant molecule library from four sources: pheromones (Li et al. 2023), food odors (Chi et al. 2025), fragrances (thegoodscentscompany.com), and theoretical odors (Mayhew et al. 2022). This approach ensured ecological relevance and chemical diversity. After deduplication, a total of 14,499 VOCs were retained for downstream analysis. We used Dragon (v.5.4) (Haddad et al. 2008) to calculate 32 representative molecular descriptors for each VOC, and the ChemmineR (Cao et al. 2008) R package to identify functional groups. VOCs were assigned functional group labels based on group priority rules. Dimensionality reduction of the 32 descriptors was performed using t-SNE, and the resulting two-dimensional odor space was visualized using Matplotlib (Hunter 2007). All VOC SMILES were converted to db2 format using the online resource tldr.docking.org. OR structures were aligned to the cryo-EM structure of MhOR5 (PDB:7LIC) to estimate the approximate position of the binding pocket. Virtual docking was then performed using DOCK 3.8 (RRID:SCR_000128) (Coleman et al. 2013), with batch processing handled via custom scripts ‘prepare_dock_file.sh’ and ‘run_dock_paraFly.sh’. In total, we obtained 130 million docking records between ORs and VOCs. This dataset serves as a foundational resource for predicting potential OR functions.

Hit rate curve prediction and maximum hit estimation

To determine the docking score range at which VOCs are likely to exhibit binding activity in electrophysiological assays, we applied Bayesian statistical modeling using prior probability distributions to define the top, midpoint, random hit rate, and slope of the hit rate curve. We assumed that the relationship between docking score (energy, ei) and hit rate follows a dose-response-like curve (fig. S10A) (Lyu et al. 2019). HitRate(ei)=top−bottom1+e−S⋅(ei− dock 50)+ This function is defined by four parameters: (1) top: the maximum hit rate; (2) dock50: the docking score at which the hit rate is half of the top value (in kcal mol⁻¹); (3) S: the slope of the curve, calculated as S = slope × 4 / top, where slope is the percentage change in hit rate at dock50 per kcal mol⁻¹; (4) bottom: the minimum hit rate, fixed at 0. To define prior distributions, we analyzed OR hit rates for Drosophila melanogaster (Hallem and Carlson 2006), Anopheles gambiae (Carey et al. 2010), and Spodoptera littoralis (de Fouchier et al. 2017) across eight docking energy intervals and used this empirical data to derive posterior distributions of the model parameters. We specified independent Bayesian priors for each parameter: P(top) ∼ beta (α=20, β= 80), P(dock50) ∼ normal (μ = -10, σ = 15), P(S) ∼ normal (μ = -0.6, σ = 0.1) (fig. S10B to D). Sampling from the prior distributions was performed using Hamiltonian Monte Carlo implemented in Stan, with parameters: ‘num_samples = 100000’, ‘num_chains = 4’, ‘num_warmup = 50000’, ‘max_depth = 12’.

Functional similarity network

We first calculated the Pearson correlation coefficients (PCCs) for docking energies between all OR pairs across the VOCs library. ORs were then ranked by the number of small molecules they bind, and a greedy clustering algorithm was applied to pre-cluster ORs with high functional similarity (PCCs ≥ 0.6), resulting in 6,299 nodes. The edge weight between nodes was defined as the average PCC of all OR pairs. The functional similarity network was then constructed using the “trunk-branch” strategy. Due to the complexity of OR functional divergence, our strategy yielded a modularity value that deviated by 0.1 from the theoretical Louvain modularity. To ensure accurate community assignment, we adopted the Louvain algorithm for final community detection. A PCC threshold of 0.48 was selected as the cutoff for defining the final functional similarity network and OR functional communities. The network was visualized using the organic layout algorithm in Cytoscape (Shannon et al. 2003).

OR binding breadth to VOCs

We assigned a VOC recognition breadth label to each functional community. First, we counted the number of VOCs with each functional group that were recognized by ORs in each functional community. These counts were then compared against the total number of ORs recognizing that VOCs across all ORs, using Bonferroni-corrected post hoc tests. If a community showed a significantly higher or lower recognition count, it was considered to have distinct recognition capacity for that functional group. The results were normalized by the average VOC count to derive a recognition breadth score for each subfamily.

Based on the functional community labels, we further proposed a metric called binding breadth index (BBI) to quantify the olfactory breadth of a group of ORs toward a particular class of VOCs. To correct for sampling bias caused by small community sizes, we applied the Wilson score interval. (Wk)=pk+z22n−zpk(1−pk)n+z24n21+z2n The definitions are as follows: (1) n: total number of ORs in the community; (2) pk: proportion of ORs in the kth community; (3) z: Z-score for the 95% confidence level (z = 1.96)

The BBI for the ith functional group of VOCs is defined as: BBIi=∑k=1n(Wk⋅Fki) Where: (1) Wk: Wilson score for the kth community; (2) Fki: the functional label matrix element representing the recognition score of kth community for ith functional group of VOCs; (3) n: the total number of functional communities (n=6)

This approach provides a statistically robust framework to characterize and compare the olfactory recognition breadth of ORs.

Association between life habits and VOC recognition breadth

We calculated the overall BBI for each species based on combinations of life habits. Using BBI scores across functional groups of VOCs as input features, we performed t-SNE dimensionality reduction, followed by K-means clustering, identifying three major life habit clusters corresponding to high, moderate, and low levels of small-molecule recognition breadth. To further explore the association, we performed binary logistic regression using the R package glmnet, focusing on species in the high and low breadth categories. All life habits were encoded using one-hot encoding, and the dataset was balanced using the ROSE (Random Over Sampling Examples) method to address class imbalance. We split the dataset into training (80%) and testing (20%) sets. The training data underwent 10-fold cross-validation to determine the optimal value of the regularization parameter lambda (fig. S11A and B). The model’s performance on the testing set was evaluated using the AUC. Finally, we fit the model using the testing data and the selected optimal lambda value, and extracted the coefficients of the predictor variables, identifying key life habits associated with VOC recognition breadth.

eLife Assessment

This large-scale comparative study of odorant receptor (OR) genes across more than 100 insect species, combining sequence- and structure-based approaches, aims to explore the evolution of this large gene family involved in the detection of odorant signals by olfactory neurons. This useful work uncovers a structural feature unique to the odorant receptor co-receptor Orco that reduces ligand binding affinity. However, the strength of evidence is incomplete: the pipeline for in silico identification of odorant receptor genes lacks validation through comparison with known odorant receptor repertoires from previously studied species, and claims regarding odor response spectra, evolutionary, and ecological interpretations are not fully supported by the analyses.

Reviewer #1 (Public review):

Objectives of the study and impact of the work:

The authors of this article primarily aim to reconstruct the evolutionary history of the insect odorant receptor (OR) family, which is responsible for the detection of odorant signals by olfactory neurons. Due to the lack of phylogenetic signal present in the sequences of this multigene family, which evolves very rapidly, phylogenetic analyses have so far never made it possible to precisely retrace how ORs diversified prior to the appearance of present-day insect orders, and what the drivers of this diversification were. For example, one may suspect that the adaptation of ORs to odors emitted by plants constituted a critical step in insect evolution during the "angiosperm terrestrial revolution," which occurred at the end of the Cretaceous, but nothing currently allows this to be asserted.

There are very nice examples, notably in Drosophilids, derived from comparisons between closely related species and documenting mechanisms of OR adaptation to certain signals. However, what the authors attempt to do in this work is to produce a macroevolutionary analysis at the scale of insects as a whole, based almost exclusively on bioinformatic analyses. To do this, they annotated OR genes in about one hundred insect species and developed pipelines for analyzing sequence similarity, structural similarity, and functional similarity, the latter being estimated through a molecular docking approach. An important feature in the evolution of insect ORs is the emergence of a unique co-receptor, called Orco, which appears to be an OR that has lost the ability to bind odorants. In addition to the large-scale bioinformatic analysis, the authors also aim to explore more specifically the factors that favored the emergence of Orco and the selective advantage conferred by the existence of OR-Orco complexes.

Given the importance of odorant receptors in insect biology and in their adaptation to different environments and lifestyles, retracing their evolutionary history is indeed a major question in evolutionary biology. In principle, this type of work therefore has the potential to become a reference in the field and to provide a basis for significant scientific advances.

Major strengths and weaknesses:

The sampling chosen for collecting OR sequences is very impressive, with more than 100 insect families represented, covering most of the major orders. This sampling appears appropriate for the question being addressed. The analysis pipeline used to collect the sequences makes sense, relying on homology-based annotation tools coupled with a structure-based filter. Nevertheless, one can note aberrant numbers of ORs for certain species (much lower than reality), which indicates that the pipeline probably did not function correctly for all genomes. In the absence of a validation step comparing the results with already known OR repertoires, it is difficult to estimate the overall quality of the data. The authors chose to apply a fairly stringent filter on sequence quality (based on predicted 3D structure), which reduces the number from 14,000 to 9,000. This choice seems logical given the subsequent use of these data, but it inevitably leads to data loss. The fact that some OR genes may be missing and that the total number may not be exact for each species is not prohibitive for studying the evolution of the family at a broad scale; however, it calls into question certain results that rely on this total number, such as the correlation between the number of ORs and genome size, lifestyle, and diet.

From the dataset collected, the authors attempted to categorize ORs in several ways, starting with the reconstruction of sequence similarity networks. The approach is interesting, but in the end, the results do not seem to be sufficiently exploited, and it is not obvious what the advantage of this approach is compared with the "classical" phylogenetic approach, which generally fails to reveal homology relationships between ORs from species belonging to different insect orders. Here again, the majority of the clusters identified are "order-specific," and when this is not the case, the authors did not attempt to exploit the results. For example, clusters SeqC26 or SeqC28, which appear to be shared by many insects, are potentially very interesting. It might have been relevant to combine this similarity-based clustering approach with phylogenetic reconstructions within each shared cluster.

The clustering based on structure also leads to the identification of a majority of "order-specific" clusters, but once again, the clusters shared by several orders are not truly exploited, which does not provide new insight into the evolution of ORs. However, the authors highlight a group of ORs in flies that appear to possess an unusual intracellular region. This is interesting, although it is a result more relevant to OR structure than to their evolution. The function of these ORs in Drosophila melanogaster, if it is known, is not discussed.

The analysis of structural diversity then leads the authors to focus on the Orco co-receptors, which are characterized by modifications of the binding pocket and the emergence of an extracellular loop that could explain the loss of the ability to bind odorant molecules. This part, which relies on in vitro experiments, is interesting and constitutes the most striking result of the study, which could in itself have been the subject of a separate manuscript. However, the molecular dynamics modelling does not add anything in the way it is conducted (5 ns is too short).

The rest of the manuscript is based on the prediction of OR response spectra using molecular docking. The work that has been carried out is extremely substantial, and the objective of linking clusters based on sequence similarity or 3D structural similarity with functional categories is entirely relevant. Nevertheless, I see two major problems with this in silico functional analysis:

(1) The docking score threshold used was chosen thoughtfully, which is very good, and according to the calculation performed, should ensure a true positive rate of more than 20%, which is excellent in such a docking analysis. But in the absence of functional validation, this 20% true positive rate is not sufficient to extrapolate OR function as the authors do in the remainder of the manuscript. The risk of error remains too high to compare in such detail the function of ORs from insects with different lifestyles or diets.

(2) The six functional clusters identified are only slightly different from one another, with similar detection of all chemical families except acids and amines (which was expected, given that these families are a priori detected by IRs rather than ORs). This shows that even though the approach is relevant and deserves to be tested, it cannot be used to establish a link between groups/lineages of ORs and response spectra at the scale of insects as a whole. This is reflected in the final analysis by the fact that there is no visible link between sequence or structural clusters and functional clusters. Given the uncertainty surrounding the docking results, the entire subsequent analysis of the relationship between the Binding Breadth Index and ecological variables is highly questionable.

Finally, the evolutionary analysis proposed to conclude that the work suffers from an incorrect interpretation: ORs of non-holometabolous insects cannot be considered equivalent to those of species that existed before the Permian-Triassic extinction. The fact that a locust or a cockroach has more narrowly tuned ORs than holometabolous insects does not mean that this was also the case for ancestral insects. To advance this type of conclusion, it would be necessary to conduct a phylogenetic analysis and reconstruct ancestral states, which is not the case here.

In summary, despite the large number of analyses performed, the authors do not succeed in achieving the stated objective of reconstructing the evolutionary history of insect ORs, and the results obtained do not sufficiently support the conclusions regarding the links between OR repertoires and environment or lifestyle.

Reviewer #2 (Public review):

The remarkable evolvability of the olfactory system enables animals to rapidly adapt to dynamic and chemically complex environments. Over the past two decades, substantial effort has been devoted to uncovering the evolutionary principles that drive the diversification of odorant receptors (ORs), yielding key insights into the forces shaping their striking variability in both vertebrates and insects. In this manuscript, Zhang and colleagues analyze the OR repertoires of over 100 insect species, leveraging sequence and structural similarity to infer patterns of gene family evolution within this diverse and ecologically important clade. By integrating sequence-based and structure-based comparisons, their study builds on a compelling and recently emerging line of research made possible by the advent of AlphaFold, which has previously clarified the phylogenetic relationship between insect Ors and the gustatory receptor gene family and revealed the unexpectedly deep evolutionary origins of this ancient structural fold.

Applying this approach to a large set of ORs derived from species throughout the insect phylogeny, the authors confirm many previously reported patterns of OR evolution. Unfortunately, the way these results are presented lacks clarity in what is already known from previous work in the field versus what is a novel finding based on the analysis of this dataset.

It is unclear how complete the odorant receptor sets are. I recommend benchmarking the pipeline by comparing its output to a gold standard and a frequently vetted complete OR set, such as that of Robertson and Wanner 2006 or similar.

Using their structural clustering approach, the authors identify a structural feature mostly unique to the OR co-receptor ORco, a beta-sheet in EL2, which they functionally show reduces odorant binding affinity - a key aspect of ORco, which does not bind ligands in the ancestral ligand-binding site. This is a particularly strong part of the manuscript, since the authors support their in silico-derived hypothesis with functional data.

Lastly, in an attempt to assess the relationship between sequence identity and structure on one hand and function on the other, the authors perform an in silico structure prediction and chemical docking analysis. As it stands, this part is on the more speculative side since the docking approach has not been verified with available functional datasets.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Objectives of the study and impact of the work:

The authors of this article primarily aim to reconstruct the evolutionary history of the insect odorant receptor (OR) family, which is responsible for the detection of odorant signals by olfactory neurons. Due to the lack of phylogenetic signal present in the sequences of this multigene family, which evolves very rapidly, phylogenetic analyses have so far never made it possible to precisely retrace how ORs diversified prior to the appearance of present-day insect orders, and what the drivers of this diversification were. For example, one may suspect that the adaptation of ORs to odors emitted by plants constituted a critical step in insect evolution during the "angiosperm terrestrial revolution," which occurred at the end of the Cretaceous, but nothing currently allows this to be asserted.

There are very nice examples, notably in Drosophilids, derived from comparisons between closely related species and documenting mechanisms of OR adaptation to certain signals. However, what the authors attempt to do in this work is to produce a macroevolutionary analysis at the scale of insects as a whole, based almost exclusively on bioinformatic analyses. To do this, they annotated OR genes in about one hundred insect species and developed pipelines for analyzing sequence similarity, structural similarity, and functional similarity, the latter being estimated through a molecular docking approach. An important feature in the evolution of insect ORs is the emergence of a unique co-receptor, called Orco, which appears to be an OR that has lost the ability to bind odorants. In addition to the largescale bioinformatic analysis, the authors also aim to explore more specifically the factors that favored the emergence of Orco and the selective advantage conferred by the existence of OR-Orco complexes.

Given the importance of odorant receptors in insect biology and in their adaptation to different environments and lifestyles, retracing their evolutionary history is indeed a major question in evolutionary biology. In principle, this type of work therefore has the potential to become a reference in the field and to provide a basis for significant scientific advances.

Major strengths and weaknesses:

The sampling chosen for collecting OR sequences is very impressive, with more than 100 insect families represented, covering most of the major orders. This sampling appears appropriate for the question being addressed. The analysis pipeline used to collect the sequences makes sense, relying on homology-based annotation tools coupled with a structure-based filter. Nevertheless, one can note aberrant numbers of ORs for certain species (much lower than reality), which indicates that the pipeline probably did not function correctly for all genomes. In the absence of a validation step comparing the results with already known OR repertoires, it is difficult to estimate the overall quality of the data. The authors chose to apply a fairly stringent filter on sequence quality (based on predicted 3D structure), which reduces the number from 14,000 to 9,000. This choice seems logical given the subsequent use of these data, but it inevitably leads to data loss. The fact that some OR genes may be missing and that the total number may not be exact for each species is not prohibitive for studying the evolution of the family at a broad scale; however, it calls into question certain results that rely on this total number, such as the correlation between the number of ORs and genome size, lifestyle, and diet.

We thank the reviewer for raising this important concern. To objectively evaluate how much our strict structural filtering may have reduced OR counts, we collected published OR annotations for species included in our study and compared those values with our structurally intact OR counts (Author response table 1). For most previously annotated species, the difference was within approximately ten ORs, indicating that our counts are generally comparable to published annotations after structural filtering. Diabrotica virgifera virgifera was a clear exception. Previous work, using transcriptomic evidence and manual annotation, reported 193 ORs, including 124 complete ORs and three pseudogenes, whereas our original pipeline retained only nine structurally intact ORs. We found that this discrepancy was mainly caused by insufficient query representation. The previous D. virgifera virgifera annotation was published in 2025, whereas our original annotation was performed in 2022, so our query set likely did not adequately represent this lineage. When we re-annotated this species using new queries, 78 complete ORs were retained after structural filtering. We have updated the D. virgifera virgifera annotation in the revised dataset.

We also considered genome size as a possible contributor to low automated recovery. Among the 115 species analyzed, D. virgifera virgifera has the third-largest genome, after Locusta migratoria and Thermobia domestica. Large genomes contain many repetitive regions, which can make automated OR annotation more difficult and may reduce the number of recovered intact gene models. The two larger-genome species in our dataset had previously been manually annotated with transcriptomic support, suggesting that automated annotation alone may be less reliable for large, repeat-rich genomes. At the same time, published OR annotations are not always free from error. For example, Propsilocerus akamusi was previously reported to have 17 ORs, whereas our pipeline retained ten complete ORs. After downloading and modeling the 17 published sequences, we found that PaOR1, PaOR5, PaOR6, PaOR9, and PaOR14 differed substantially from canonical OR structures or appeared to be incorrectly annotated (Author response image 1). Truncated genes such as PaOR4, which lacks part of the intramembrane region of TM2-4 but still clearly resembles an OR, would be retained by our strategy; sequences with more severe structural inconsistency would be excluded. Because many previous OR annotations have not been experimentally validated, it is difficult to treat all published OR counts as exact ground truth.

To test whether possible OR loss affected OR-count-based conclusions, we first removed all potential GR sequences from the dataset. For species with published OR counts, we then repeated the pGLS analyses using the larger of our structurally intact OR count and the published OR count (Fig. S1b-e). The main ecological associations were retained: OR count remained associated with larval diet, adult diet, and habitat, and was not associated with circadian rhythm. This indicates that the main OR count ecological analyses are reasonably robust to moderate underestimation caused by missed ORs or strict structural filtering.

In contrast, the association between OR count and genome size was no longer supported in this sensitivity analysis (Author response image 2a and b). We therefore removed the analysis and discussion of a negative relationship between genome size and OR count. The original pattern likely reflected the difficulty of annotating ORs in large, repeat-rich genomes rather than a reliable biological relationship.

The following revision has been added to the revised manuscript (lines 122-126): " We then repeated the phylogenetic generalized least squares (pGLS) analyses using the larger of our structurally intact OR count and the published OR count, and the main ecological associations were retained. This indicates that the main OR-count ecological analyses are reasonably robust to moderate underestimation caused by missed ORs or strict structural filtering."

These new benchmark and sensitivity analyses directly strengthen the evidence supporting our dataset and its ecological interpretations.

From the dataset collected, the authors attempted to categorize ORs in several ways, starting with the reconstruction of sequence similarity networks. The approach is interesting, but in the end, the results do not seem to be sufficiently exploited, and it is not obvious what the advantage of this approach is compared with the "classical" phylogenetic approach, which generally fails to reveal homology relationships between ORs from species belonging to different insect orders. Here again, the majority of the clusters identified are "order-specific," and when this is not the case, the authors did not attempt to exploit the results. For example, clusters SeqC26 or SeqC28, which appear to be shared by many insects, are potentially very interesting. It might have been relevant to combine this similarity-based clustering approach with phylogenetic reconstructions within each shared cluster.

We thank the reviewer for this insightful suggestion. We agree that the previous version did not sufficiently explore the evolutionary information contained in shared sequence communities such as SeqC26 and SeqC28. Following the suggestion, we extracted the sequences from these key shared communities and reconstructed within-community gene trees to search for orthologous groups.

Because shared SeqCs may contain conserved genes across insect orders, we used the gene trees of SeqC26 and SeqC28 to identify candidate cross-order orthologous groups. In SeqC26, we identified 56 orthologous groups, 11 of which included more than ten species. Among these 11 larger groups, 4 spanned multiple insect orders. One example, OGG54, includes sequences from Orthoptera, Blattodea, Trichoptera, and Hymenoptera; after excluding the possibility of GR contamination, this group includes locust LmigOR5 and LmigOR4. LmigOR5 has been reported to bind geranyl acetone and to be associated with avoidance behavior in the migratory locust (Chang et al. 2023). The functions of the corresponding receptors in other insect orders remain to be tested, and we now present these as candidate conserved modules rather than confirmed functional orthologs. In contrast, we did not identify clear cross-order orthologous groups in SeqC28. This suggests that the ORs in SeqC28 are more similar at the sequence-community level, but do not necessarily reflect traceable orthologous relationships.

The following revision has been added to the revised manuscript (lines 156-164): "SeqC26 and SeqC28 are large shared clusters that may contain cross-order orthologous relationships. In SeqC26, we identified 56 orthologous groups, four of which spanned multiple insect orders. OGG54 included ORs from Orthoptera, Blattodea, Trichoptera, and Hymenoptera. This group contains the locust receptors LmigOR5 and LmigOR4. LmigOR5 has been reported to bind geranyl acetone and to be associated with avoidance behavior in the migratory locust. The functions of the corresponding receptors in other insect orders remain to be tested, and we now present these as candidate conserved modules rather than confirmed functional orthologs. In contrast, we did not identify clear cross-order orthologous groups in SeqC28."

The clustering based on structure also leads to the identification of a majority of "orderspecific" clusters, but once again, the clusters shared by several orders are not truly exploited, which does not provide new insight into the evolution of ORs. However, the authors highlight a group of ORs in flies that appear to possess an unusual intracellular region. This is interesting, although it is a result more relevant to OR structure than to their evolution. The function of these ORs in Drosophila melanogaster, if it is known, is not discussed.

We thank the reviewer for this useful suggestion. Similar to the sequence communities, the structural communities also contain broadly shared groups. In the revised analysis, we focus in particular on StrC23, the only structural community present in 12 insect orders. StrC23 accounts for 46.3% of all structurally intact ORs in our dataset (4124 of 8905 ORs).

Given the extremely low sequence similarity among insect ORs, the existence of such a broad conserved structural community is important. We found that StrC23 has a significantly larger binding-pocket volume than other ORs and Orco (Fig. S3 d). From a structural perspective, this suggests that StrC23 receptors may have greater potential to accommodate diverse VOCs. We therefore interpret StrC23 not as sequence-level conservation, but as conservation at the level of OR structural evolution: many species appear to retain a large set of ORs with relatively large binding pockets, which may provide a structural basis for broad docking-derived VOC binding potential before lineage-specific OR diversification occurs.

Following the reviewer’s suggestion, we also examined available functional data for the long-IL3 receptors in StrC17. The long-IL3 orthologous group includes Drosophila melanogaster DmOr13a (droMel44 in our dataset) and Bactrocera dorsalis BdorOR13a (bacDor_18 in our dataset). Previous studies indicate that both receptors respond to 1-octen-3-ol (Kreher et al. 2008; Xu et al. 2023). Earlier work also identified a group of Dipteran fly OR homologs, including DmOr13a and BdorOR13a, as 1octen-3-ol-specific responsive receptors associated with oviposition behavior(Liu et al. 2023). When we modeled these homologous receptors, they also showed an extended IL3 region.

We therefore propose that the long-IL3 ORs identified here likely correspond to the previously reported fly homologs involved in 1-octen-3-ol responses. However, 1-octen-3-ol has different behavioral functions in different species: it acts as an attractant in blood-feeding mosquitoes and tsetse flies (Hall et al. 1984; Kline et al. 2007), contributes to plant-host localization in parasitoids(Morawo and Fadamiro 2016), functions as an aggregation cue in some beetles(Pierce et al. 1989), and is often associated with oviposition regulation in flies (Kreher et al. 2008; Liu et al. 2023). Mosquito receptors known to detect 1-octen-3-ol, such as AgOR8, AaOR8, TaOR8, and CquiOR118b(Lu et al. 2007; Xu et al. 2015; Dekel et al. 2016; Frunze et al. 2024), do not contain a long IL3 region (Fig. S3 e). This raises the possibility that the long IL3 structure in fly homologs may relate to the oviposition-associated role of 1-octen-3-ol in flies.

Because OR ligand binding is primarily mediated by the binding pocket, IL3 may not directly determine ligand binding. Instead, it may influence downstream regulation of olfactory responses. Mutational analysis of Orco suggests that IL3 can regulate channel activation and Orco-dependent olfactory responses, indicating that variation in intracellular loops may affect response efficiency or sensitivity(Turner et al. 2014; Bobkov et al. 2021). We now discuss this only as a plausible mechanism requiring future experimental validation.

The following revision has been added to the revised manuscript (lines 196-202): "StrC23 is the only structural community present in 12 insect orders and accounts for 46.3% of all structurally intact ORs in the dataset. We found that StrC23 has a much larger binding-pocket volume than other ORs and Orco. From a structural perspective, this suggests that StrC23 receptors may have greater potential to accommodate diverse VOCs. This reflects conservation at the level of OR structural evolution: many species appear to retain a relatively large set of ORs with larger binding pockets, which may provide a structural basis for structure-derived volatile organic compound binding potential before species-specific OR diversification."

The following revision has also been added (lines 408-419): "The long-IL3 OR orthologous group includes Drosophila melanogaster DmOr13a and Bactrocera dorsalis BdorOR13a, both of which respond to 1-octen-3-ol and belong to a group of dipteran fly OR homologs associated with oviposition behavior. Modeling of these homologous ORs showed that they also contain a long IL3 region. We therefore propose that long-IL3 ORs correspond to these fly homologs that recognize 1-octen-3-ol. However, 1-octen-3-ol has different behavioral functions in different species: it acts as an attractant in blood-feeding mosquitoes and tsetse flies, whereas in flies it is usually associated with oviposition regulation. Mosquito receptors known to detect 1-octen-3-ol, such as AgOR8, AaOR8, TaOR8, and CquiOR118b, do not contain a long IL3 region. This suggests that the long IL3 structure in fly homologs may be related to the oviposition-associated role of 1-octen-3-ol in flies, which requires future experimental validation."

The analysis of structural diversity then leads the authors to focus on the Orco co-receptors, which are characterized by modifications of the binding pocket and the emergence of an extracellular loop that could explain the loss of the ability to bind odorant molecules. This part, which relies on in vitro experiments, is interesting and constitutes the most striking result of the study, which could in itself have been the subject of a separate manuscript. However, the molecular dynamics modelling does not add anything in the way it is conducted (5 ns is too short).

We thank the reviewer for the positive assessment of our Orco results and for raising this important concern. We agree that 5 ns is insufficient to evaluate the conformational stability or long‑term dynamics of the receptor complex. However, our simulations were designed specifically to examine whether the EL2 β‑sheet could sterically affect the early movement of VOCs near the extracellular entrance of the binding pocket. Importantly, rather than relying on a small number of long trajectories, we performed 60 independent short simulations with the explicit goal of capturing a statistical trend in how VOCs respond to this local steric constraint. Within 5 ns, VOC trajectories were visibly altered by the EL2 β‑sheet. Across the 60 replicates, VOCs reached the binding pocket less frequently in Orco proteins containing this structure, revealing a consistent trend (Fig. 4C and Movie S1). We therefore consider this timescale sufficient to detect the early steric response of small molecules, and the large number of repeats provides statistical confidence in this conclusion.

The following revision has also been added (lines 688–691): “It should be noted that the molecular dynamics simulations were designed to examine the early trajectories of VOCs near the extracellular entrance of the binding pocket, rather than to assess the conformational stability of the receptor complex. A duration of 5 ns is sufficient for small molecules to respond to local steric constraints.”

The rest of the manuscript is based on the prediction of OR response spectra using molecular docking. The work that has been carried out is extremely substantial, and the objective of linking clusters based on sequence similarity or 3D structural similarity with functional categories is entirely relevant. Nevertheless, I see two major problems with this in silico functional analysis:

(1) The docking score threshold used was chosen thoughtfully, which is very good, and according to the calculation performed, should ensure a true positive rate of more than 20%, which is excellent in such a docking analysis. But in the absence of functional validation, this 20% true positive rate is not sufficient to extrapolate OR function as the authors do in the remainder of the manuscript. The risk of error remains too high to compare in such detail the function of ORs from insects with different lifestyles or diets.

We thank the reviewer for pointing out this important limitation of our computational functional interpretation. We fully agree that, although the threshold calibrated from available experimental data improves enrichment, a true-positive rate of approximately 20% is not sufficient to make deterministic claims about individual OR-VOC pairs. We therefore do not equate docking predictions with true OR response spectra. To prevent any misinterpretation, we have undertaken a major revision of our terminology and framing throughout the manuscript:

The main purpose of docking in our study is not to predict the real ligand of every OR or to compare the exact functional properties of individual receptors across lifestyles. Because the study includes 115 insect species, thousands of ORs, and a large VOC set, systematic experimental validation of all ORVOC combinations is currently not feasible. Instead, we use docking as a high-throughput and internally comparable theoretical screening method to estimate the potential binding tendencies of OR repertoires toward different functional-group VOCs.

Under this framework, we focus on repertoire-level trends generated by a uniform structural modeling, docking, and scoring pipeline. The value of this analysis is to provide candidate receptors, VOC categories, and ecological associations for future experimental testing. In the revised manuscript, when the results are based solely on docking analyses, we avoid directly referring to them as OR functions. Instead, we add the prefix “docking-derived” to prevent potential misunderstanding. We also explicitly state that confirming specific OR ligands and behavioral functions will require heterologous expression, electrophysiology, calcium imaging, genetic perturbation, or behavioral assays.

The following limitation statement has been added to the revised manuscript (lines 285-288): "Although this threshold enriches experimentally responsive OR-VOC pairs, the resulting hit rate is not sufficient to support deterministic inference for individual receptor-ligand pairs. Therefore, subsequent analyses were interpreted at the OR repertoire level rather than as experimentally validated OR response spectra.”; (lines 292-293) “It should be noted that dFunCs represent docking-derived functional communities rather than experimentally confirmed functional classes.”

(2) The six functional clusters identified are only slightly different from one another, with similar detection of all chemical families except acids and amines (which was expected, given that these families are a priori detected by IRs rather than ORs). This shows that even though the approach is relevant and deserves to be tested, it cannot be used to establish a link between groups/lineages of ORs and response spectra at the scale of insects as a whole. This is reflected in the final analysis by the fact that there is no visible link between sequence or structural clusters and functional clusters. Given the uncertainty surrounding the docking results, the entire subsequent analysis of the relationship between the Binding Breadth Index and ecological variables is highly questionable.

We thank the reviewer for this key comment. We agree that the differences among dFunCs for any single VOC functional group are often modest. However, the dFunC classification was not based only on whether an OR strongly recognizes one specific functional-group category. Rather, it was based on each OR’s overall docking-derived binding profile across a large VOC set.

In other words, the differences among dFunCs mainly reflect combinations of relative binding tendencies across multiple VOC categories, rather than a strong difference for one chemical family alone. Therefore, even if the predicted binding level for a single functional group differs only slightly among dFunCs, their multivariate profiles can still form distinct dFunCs.

We also agree that SeqCs, StrCs, and dFunCs do not show a simple one-to-one correspondence. This is expected because the three classifications are based on different information: sequence similarity, overall structural similarity, and docking-derived binding-profile similarity. At present, there is no direct evidence that insect OR sequence, structure, and potential binding profile maintain a conserved one-to-one relationship across the insect class. We also cannot exclude the possibility that false positives in docking reduce the resolution of the functional classification.

To avoid overinterpretation, we now refer to FunCs as docking-derived functional communities (dFunC) and define BBI as docking-derived BBI. This metric is intended for macro-scale relative comparisons, not for assigning true ligands or specific ecological functions to individual ORs. Ecological interpretations based on dFunC or BBI have been substantially toned down and are presented as candidate trends for future functional validation.

The following limitation statement has been added to the revised manuscript: "Ecological associations based on BBI should be interpreted as repertoire-level predicted binding trends, not as direct evidence for lifestyle-specific OR function or behavioral olfactory capacity."

This reframing ensures our conclusions are appropriately supported by the computational evidence and aligns the manuscript's claims with its methodological strengths.

Finally, the evolutionary analysis proposed to conclude that the work suffers from an incorrect interpretation: ORs of non-holometabolous insects cannot be considered equivalent to those of species that existed before the Permian-Triassic extinction. The fact that a locust or a cockroach has more narrowly tuned ORs than holometabolous insects does not mean that this was also the case for ancestral insects. To advance this type of conclusion, it would be necessary to conduct a phylogenetic analysis and reconstruct ancestral states, which is not the case here.

In summary, despite the large number of analyses performed, the authors do not succeed in achieving the stated objective of reconstructing the evolutionary history of insect ORs, and the results obtained do not sufficiently support the conclusions regarding the links between OR repertoires and environment or lifestyle.

We thank the reviewer for this important point. We agree that a rigorous test of changes in ancestral OR functional spectra before and after the EPME would require ancestral reconstruction within a phylogenetic framework based on reliable orthologous groups. For insect ORs, this is currently limited by rapid sequence divergence, very restricted cross-order and even within-order orthology, and frequent gene duplication, loss, and lineage-specific diversification.

Therefore, our current data cannot reliably reconstruct complete ancestral OR repertoires for different insect-order nodes, nor can they accurately trace extant ORs back to ancestral OR states before and after the EPME. The analysis is better understood as a comparison of OR repertoire composition among extant insect lineages, rather than as a comparison of ancestral states.

If major environmental transitions influenced the long-term evolution of insect ORs, such effects may appear either as traceable orthologous changes or as broad differences in extant repertoire composition and binding-profile composition. However, differences observed among extant lineages cannot be directly attributed to the EPME because they may also reflect later lineage-specific expansions, gene losses, and ecological adaptations.

Accordingly, we have revised the conclusion. We no longer state that ancestral insect OR functional spectra changed before and after the EPME. Instead, we state that comparisons among extant insect repertoires reveal differences in docking-derived binding-profile composition among lineages. These patterns may provide hypotheses for exploring links between major geological/ecological transitions and long-term olfactory evolution, but they require broader phylogenetic sampling, ancestral reconstruction, and functional validation.

The following revision has been added to the revised manuscript (lines 427-430): "We also detected differences in dFunC composition between extant lineages whose order-level origins fall before and after the EPME. Lineages originating after the EPME showed a higher proportion of broad-tuned ORs, likely influenced by multiple factors."

We also added the following clarification (lines 444-450): “However, because insect ORs evolve rapidly and cross-order orthologous relationships are limited, the current dataset does not allow reliable reconstruction of complete ancestral OR repertoires at deep insect nodes. The EPME-related comparison should therefore be interpreted as a descriptive comparison among extant lineages, not as a direct test of ancestral OR functional changes before and after the EPME. This hypothesis will require further testing through broader taxon sampling, reliable orthology assignment, ancestral-state reconstruction, and functional assays.”

This revision directly addresses the reviewer's concern by removing the unsupported causal claim and reframing the analysis within the appropriate, evidence-based scope of our study, thereby strengthening the manuscript's contribution as a source of robust comparative patterns and testable macroevolutionary hypotheses.

Reviewer #2 (Public review):

The remarkable evolvability of the olfactory system enables animals to rapidly adapt to dynamic and chemically complex environments. Over the past two decades, substantial effort has been devoted to uncovering the evolutionary principles that drive the diversification of odorant receptors (ORs), yielding key insights into the forces shaping their striking variability in both vertebrates and insects. In this manuscript, Zhang and colleagues analyze the OR repertoires of over 100 insect species, leveraging sequence and structural similarity to infer patterns of gene family evolution within this diverse and ecologically important clade. By integrating sequence-based and structure-based comparisons, their study builds on a compelling and recently emerging line of research made possible by the advent of AlphaFold, which has previously clarified the phylogenetic relationship between insect Ors and the gustatory receptor gene family and revealed the unexpectedly deep evolutionary origins of this ancient structural fold.

Applying this approach to a large set of ORs derived from species throughout the insect phylogeny, the authors confirm many previously reported patterns of OR evolution. Unfortunately, the way these results are presented lacks clarity in what is already known from previous work in the field versus what is a novel finding based on the analysis of this dataset.

We thank the reviewer for pointing this out. We agree that the original manuscript did not distinguish previously established findings from the novel results of the present study with sufficient clarity. Rapid OR evolution, lineage-specific expansion, and associations between OR repertoires and ecological traits have been reported in several insect groups. We have revised the manuscript to acknowledge these studies more explicitly and to indicate which results confirm known patterns, which extend them across a broader phylogenetic scale, and which arise from our new analyses.

Our study contributes more than a confirmation of previous observations. To our knowledge, it provides the first integrated analysis of OR evolution across 115 insect species that combines sequence similarity, structural conservation, and docking-derived binding profiles within a unified macroevolutionary framework. This analysis identifies class-wide repertoire patterns and previously unrecognized structural features, including the conserved structural cluster StrC23, that could not be evaluated using narrower taxonomic datasets.

The study also provides a mechanistic insight into Orco specialization. We experimentally demonstrate that the distinctive EL2 β-sheet of Orco reduces ligand binding affinity, supporting its contribution to the reduced odorant binding capacity of this conserved coreceptor. In addition, because exhaustive experimental characterization of the large insect OR family is currently impractical, our experimentally calibrated computational pipeline provides a framework for identifying repertoire-level trends, generating testable hypotheses, and prioritizing receptors for future functional studies.

The reviewer raises several specific examples of the distinction between previous knowledge and novel findings in the comments below. We address each of these points in the corresponding responses. We believe these revisions more clearly position our contribution as testing, extending, and integrating previously reported patterns while also identifying new class-wide structural and functional features.

It is unclear how complete the odorant receptor sets are. I recommend benchmarking the pipeline by comparing its output to a gold standard and a frequently vetted complete OR set, such as that of Robertson and Wanner 2006 or similar.

We thank the reviewer for this suggestion. We agree that without benchmarking, readers cannot adequately evaluate the completeness and accuracy of the OR annotation workflow. Following the suggestion, we performed detailed comparisons using three relatively well-curated repertoires: Drosophila melanogaster, Anopheles gambiae, and Bombyx mori. Because the species included in our study did not include Apis mellifera from Robertson and Wanner 2006, we did not use that dataset as the benchmark.

The benchmark addresses three questions. First, how well does the automated genome-based OR annotation workflow recover full-length ORs reported in previous studies? Second, does the structural filtering step improve the structural quality and reliability of the OR dataset? Third, for candidates in tandem-repeat regions where exon-merging errors may occur, do the candidate ORs share genomic positions with the most similar literature ORs, or are they more likely to represent distinct candidates? For Drosophila melanogaster, the accepted repertoire contains 60 OR genes and 65 protein products including splice variants. At the gene level, the automated workflow recovered most benchmark ORs before structural filtering, but Or47b and Or98b were not annotated. After structural filtering, fragmented or structurally incomplete ORs were removed, leaving 56 ORs that otherwise correspond one-to-one with the benchmark ORs. Two ORs that had been annotated before filtering, DmOr59a and DmOr85e, were removed by the structural filter because DmOr59a was fragmented and DmOr85e lacked TM7 in the predicted structural model (Author response image 3).

For Anopheles gambiae, the published repertoire includes 79 OR genes. Our automated workflow initially annotated 77 ORs, including 72 published ORs. After removing fragments and five full-length literature ORs, the final structurally filtered set contained 67 ORs. We inspected the five removed full-length ORs and found that AgOr52, AgOr47, and AgOr64 lacked a complete transmembrane region, whereas AgOr58 and AgOr6 contained all transmembrane regions but showed local deletions in some regions (Author response image 3). This indicates that structural filtering removes clearly incomplete structures, but can also be conservative enough to exclude a small number of true ORs with atypical transmembrane-region length variation.

For Bombyx mori, the published repertoire contains 66 OR genes including two pseudogenes. Our automated workflow initially annotated 103 ORs, including 60 previously reported ORs. After removing fragments and seven literature ORs, the final set contained 64 ORs, of which 53 corresponded to literature annotations. For structurally retained candidates that did not map one-to-one to literature ORs, we checked genomic locations (Table R2). Most were not located on the same chromosome or in the same tandem region as their most similar literature ORs, making exon-merging artifacts less likely. They may therefore represent previously unannotated OR candidates, although transcriptomic evidence or manual annotation will be needed for confirmation.

Overall, the benchmark shows that the automated workflow recovers most known OR repertoires and that structural filtering produces a high-confidence set suitable for structural comparison and docking. We now explicitly describe the final dataset as a set of structurally intact ORs rather than a complete OR repertoire for every species, and we discuss how strict filtering may underestimate OR counts. We added a benchmark description to the main text (lines 100-104) and included the detailed comparison strategy and results in Supplementary Text S1: "Benchmarking against existing OR repertoires from Drosophila melanogaster, Anopheles gambiae, and Bombyx mori showed that the annotation workflow provides high-confidence annotations suitable for downstream structural comparison and docking (see Supplementary Text S1, Table S1-3)."

Using their structural clustering approach, the authors identify a structural feature mostly unique to the OR co-receptor ORco, a beta-sheet in EL2, which they functionally show reduces odorant binding affinity - a key aspect of ORco, which does not bind ligands in the ancestral ligand-binding site. This is a particularly strong part of the manuscript, since the authors support their in silico-derived hypothesis with functional data.

We thank the reviewer for the positive assessment of the Orco-related work.

Lastly, in an attempt to assess the relationship between sequence identity and structure on one hand and function on the other, the authors perform an in silico structure prediction and chemical docking analysis. As it stands, this part is on the more speculative side since the docking approach has not been verified with available functional datasets.

We thank the reviewer for this comment and agree that the docking analysis is predictive. Molecular docking cannot be treated as experimental evidence for OR response spectra or as a definitive assignment of ligand specificity for individual OR-VOC pairs.

In the revised manuscript, we clarify that docking is used as a unified, comparable theoretical framework for macro-scale OR repertoire comparison. It is meant to estimate docking-derived binding potential across thousands of ORs and many VOCs, not to replace heterologous expression, electrophysiology, calcium imaging, or behavioral validation.

We also clarify that the docking framework was not used without any empirical calibration. We used existing OR-VOC experimental response data to examine the relationship between docking score and experimental hit rate, and we evaluated the ability of docking-derived labels to recover functional tendencies in available datasets. These analyses support the use of docking for repertoire-level enrichment and hypothesis generation, but they do not justify deterministic conclusions for individual receptor-ligand pairs. The revised manuscript now states this limitation explicitly.

The following revision has been added to the revised manuscript (lines 285-288): "Although this threshold enriches experimentally responsive OR-VOC pairs, the resulting hit rate is not sufficient to support deterministic inference for individual receptor-ligand pairs. Therefore, subsequent analyses were interpreted at the OR repertoire level rather than as experimentally validated OR response spectra."

Summary of Major Revisions:

In direct response to the eLife Assessment and reviewer comments, our revisions have systematically strengthened the evidence and clarified the interpretation:

(1) Added critical validations: Benchmarking of the annotation pipeline and sensitivity analyses for ecological correlations.

(2) Deepened evolutionary analysis: Phylogenetic exploration of shared clusters and structural characterization of StrC23.

(3) Clarified the scope and limitations of the computational functional analysis: Adopted “docking-derived” terminology and specified that the results represent hypothesis-generating, repertoire-level comparative trends rather than experimentally validated receptor functions.

(4) Corrected overinterpretations: Revised conclusions regarding the EPME and holometabolous/non-holometabolous comparisons to be descriptive and hypothesis-generating. '

(5) Improved scholarly accuracy: Updated framing to properly acknowledge prior work and corrected minor points throughout.

We believe that these revisions address the concerns underlying the assessment of incomplete evidence and provide stronger, more appropriately qualified support for the manuscript’s main conclusions.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

The annotation pipeline produces a number of genes that is slightly lower than reality, which is expected given the presence of a fairly stringent filter. However, the number is far too low for certain species, for example, Harmonia or Diabrotica. Whenever possible (for example, for Drosophila, Bombyx, Anopheles, etc.), adding a benchmarking step by comparing the results obtained with already identified repertoires would be a real improvement. Moreover, the dataset contains sequences of gustatory receptors that must be removed in order not to bias the analysis (CO₂ receptors in Drosophila melanogaster and Bombyx mori, for example, based on what I was able to verify). It might therefore be necessary to add a specific step in the pipeline to check for the presence of GRs in the data, since the sequences and 3D structures can be quite similar to those of ORs and therefore difficult to separate.

We thank the reviewer for this careful observation. To evaluate OR annotation counts, we compared our results with published OR annotations for previously annotated species. Except for Diabrotica virgifera virgifera, most species differed from published counts by fewer than ten ORs, supporting the general reliability of our annotation counts. For D. virgifera virgifera, the very low original count was likely caused by a combination of insufficient query representation and the difficulty of annotating ORs in a large genome. The previous D. virgifera virgifera annotation was published in 2025, whereas our original annotation was performed in 2022, so our query set likely did not adequately represent this lineage. When we re-annotated this species using new queries, 78 complete ORs were retained after structural filtering. We have updated the D. virgifera virgifera annotation in the revised dataset. Large-genome insects over 2 Gb are rare in our dataset, and the sensitivity analysis indicates that this issue does not substantially affect the main ecological associations based on OR count. Regarding gustatory receptor contamination, we found that a small number of sequences labeled as olfactory receptors but actually corresponding to GRs had been retained as queries because of an operational error during OR annotation. To remove potential GRs systematically, we collected 1,418 insect GR genes from the literature and databases and combined them with the OR library to create a receptor reference set. We then compared all annotated OR candidates against this reference set using BLASTP and removed candidates whose best match was a GR. In total, we identified and removed 76 potential GRs distributed across 38 species. These included the Drosophila CO2 receptor genes Gr63a and Gr21a, as well as Bombyx mori BmGr10 and BmGr9, which the reviewer specifically noted. All downstream analyses have been repeated using the corrected dataset. We are grateful for this careful observation, which substantially improved the rigor of the dataset.

Both Reviewer 1 and Reviewer 2 requested benchmarking against known OR repertoires, and we agree that this is necessary. We therefore benchmarked the workflow against Drosophila, Anopheles, and Bombyx repertoires, evaluating recovery of known ORs, the effect of structural filtering, and potential exon-merging artifacts in tandem regions. We further added sensitivity analyses using the larger of structurally intact OR counts and published OR counts for species with available literature data. These analyses are now included in the revised Results and supplementary materials.

The revised text is as follows (lines 100-104): "Benchmarking against existing OR repertoires from Drosophila melanogaster, Anopheles gambiae, and Bombyx mori showed that the annotation workflow provides high-confidence annotations suitable for downstream structural comparison and docking (see Supplementary Text S1, Table S1-3)."

For the SSN approach, the validation by comparison with the phylogeny of hymenopterans is a good idea, but the result is not as convincing as presented in the manuscript since the cluster SeqC25 does not correspond to a monophyletic group. This, therefore, raises the question of the evolutionary information that can be drawn from this clustering approach. The validation with lepidopteran PRs is also a good idea, but it would be necessary to verify exactly to which PR clades the sequences used for the analysis belong, in order to determine which clade corresponds to each of the identified SeqC clusters.

We thank the reviewer for this important point. For the Hymenoptera comparison, we agree that SeqC25 does not correspond to a strict monophyletic clade in existing Hymenoptera OR phylogenies. We have clarified that the SSN approach is not intended to replace traditional phylogenetic trees or to reconstruct ancestor-descendant relationships among OR communities.

SeqC25 should be interpreted as a sequence-similarity community that is strongly expanded in Hymenoptera and partially corresponds to previously defined Hymenoptera OR subfamilies, but not as a strict phylogenetic clade. SeqCs describe modular boundaries in OR sequence-similarity space, whereas phylogenetic trees infer branching relationships under a specific evolutionary model. For insect ORs, where sequence similarity is often very low, duplication and loss are frequent, and tree support is often limited, the two approaches need not produce identical groupings.

We now state that the evolutionary information provided by SSNs mainly concerns sequence-similarity diversification patterns: highly expanded modules, order-specific or cross-order communities, community boundaries, and differences in repertoire composition among lineages. These results complement phylogenetic analysis, but detailed relationships within close taxa or OR subfamilies still require traditional phylogenetic methods.

For the Lepidoptera pheromone receptor (PR) validation, we agree that the previous analysis mixed experimentally validated PRs and candidate PRs, making it unclear which PR branch each SeqC corresponded to. We reorganized the PR dataset by separating classic PRs from recently reported new PR branches and mapped these sequences to SeqCs using BLASTP. Classic PRs were assigned almost entirely to SeqC6 (93.9%), whereas the new PR branch mapped to SeqC35 (Fig. S2f). Both SeqC6 and SeqC35 are shared by Trichoptera and Lepidoptera, suggesting that they may correspond to distinct PR-related sequence communities.

The following revision has been added to the revised manuscript (lines 144-150): "This comparison indicates that SSN-based communities are broadly consistent with established phylogenetic subfamilies, while capturing sequence-similarity modules rather than strictly monophyletic clades. We further evaluated the SSN using curated lepidopteran pheromone receptors (PRs). To distinguish different PR lineages, experimentally validated classic PRs and the recently reported new PR lineage were analyzed separately. Classic PRs were predominantly assigned to SeqC6 (93.9%), whereas the new PR lineage was assigned to SeqC35 (fig. S2F and Data S5), supporting the hypothesis of multiple independent origins of PRs.”

(3) In Figure 3, the cluster StrC17 appears to be present in both Lepidoptera and Diptera, whereas when looking at Figure S3, it seems to be present only in Diptera. There may be an error somewhere.

We thank the reviewer for identifying this inconsistency. After rechecking the data, we confirmed that StrC17 contains 47 ORs, of which 46 are from Diptera and one is from Lepidoptera. We have corrected Figure S3. This correction does not affect the main conclusions regarding the long-IL3 ORs, because all downstream analyses were conducted using the correct StrC17 composition.

(4) The phylogeny presented in Figure S4 does not help clarify the phylogenetic context of the Orco study. The main reason is that it is not correctly rooted: the ORs of M. rhabei should have been used as an outgroup, or the two GRs of D. melanogaster that are present (perhaps by mistake) in this analysis and that clearly diverge strongly from the rest of the sequences. In any case, this phylogenetic analysis does not seem to add much compared with what was presented in the article by Thoma et al. in 2018.

We thank the reviewer for pointing this out. The original purpose of the tree was to show that early-diverging insect ORs lack clear orthologous relationships, making it difficult to reconstruct ancestral OR states before and after Orco emergence. However, because the figure contributes little to the core conclusions and could introduce confusion, we have removed it from the revised manuscript. The revised text instead directly cites previous phylogenetic studies for the early OR/Orco evolutionary background.

(5) In Figure 5A, many VOCs appear unclassified when using molecular descriptors. Have you tried classifying VOCs using molecular fingerprints instead of selected molecular descriptors? And can you explain how those 32 descriptors were chosen? Furthermore, it is surprising not to see a "terpenoids" category given the importance of these molecules in insect chemical ecology. One would expect them to cluster together in a chemical space based on molecular descriptors; where are they in Figure 5A?

We thank the reviewer for this question. We first clarify that unclassified in Figure 5A does not mean that these VOCs were excluded from the analysis. It only means that they were not assigned to one of the nine predefined major functional-group categories. All VOCs were included in molecular descriptor calculation, chemical-space visualization, and downstream docking.

The 32 molecular descriptors were not arbitrarily selected by us. They come from the optimized odorant metric proposed by Haddad et al. 2008. That study began with 1,664 Dragon molecular descriptors, represented odorants as multidimensional physicochemical vectors, and selected 32 descriptors that best explained similarity in odor-induced neural responses across multiple published datasets. Haddad et al. further showed that this optimized descriptor set performed well across different animals, recording methods, and levels of olfactory-system organization. We therefore adopted these 32 descriptors to represent odor physicochemical space.

We agree that molecular fingerprints are another useful representation, especially for capturing substructures and scaffold information. We chose molecular descriptors rather than fingerprints because the purpose of Figure 5A is to visualize VOC distribution in a continuous physicochemical odor space, and because the Haddad descriptor set was optimized using neural response data. We have added this rationale to the revised manuscript.

We also agree that terpenoids are important in insect chemical ecology. In the original classification, VOC categories were based mainly on functional groups, whereas terpenoids are defined by biosynthetic origin and carbon skeleton rather than by a single functional group. Therefore, terpenoid VOCs are distributed across several functional-group categories, such as terpene alcohols, terpene aldehydes, and terpene ketones, and are not expected to form one independent cluster in a two-dimensional odour space.

Following the reviewer’s suggestion, we added an extra annotation for common terpenoid-related compounds and mapped them onto the odor space shown in Figure 5A. The result shows that terpenoids are distributed across multiple regions and functional-group categories rather than forming a single compact cluster (Fig. S5 c). We therefore retain the major functional-group classification for the main analyses and add terpenoids as an additional annotation, while explaining why they are not treated as a main category parallel to alcohols, aldehydes, ketones, and esters.

In addition, because the downstream analyses in this study rely on large-scale docking predictions, and because docking itself contains inherent uncertainty, we chose to use relatively basic and chemically explicit major functional-group categories. This classification reduces interpretive instability that could arise from excessive subdivision of VOC classes. By contrast, treating terpenoids as an independent primary category would group together molecules with substantially different functional-group properties, thereby increasing the complexity of interpreting docking-derived binding profiles. Therefore, in the revised manuscript, we retain the major functional groups as the core VOC classification scheme and add terpenoids as an additional annotation.

The specific revision in the revised manuscript is as follows (lines 268-271): "Although terpenoids are important in insect chemical ecology, they are distributed across multiple regions of odor space. To reduce the docking instability that could result from excessive VOC subdivision, we did not treat terpenoids as a category parallel to the nine basic functional-group categories."

(6) The methodology used to calculate hit rates with respect to functional data on known ORs is not sufficiently explained. The result could be shown for the different species (drosophila, mosquito, butterfly). In addition, ORs from the locust should also have been used for this comparison.

We thank the reviewer for this suggestion. In the revised Methods, we now describe the workflow in detail. We first compiled receptor-odorant combinations with clear response relationships from published functional experiments. For each OR-VOC pair, we performed docking using the same pipeline as in the global analysis and extracted the corresponding docking score. We then labeled each pair as response or non-response according to the original experimental data. Next, we grouped pairs into docking-score intervals and calculated, within each interval, the proportion of experimentally positive pairs, which we define as the hit rate. To reduce fluctuation from random sampling, each score interval was sampled three times. Finally, following the hit-rate modeling strategy of Lyu et al. 2019, we fitted the relationship between docking score and hit rate with a Bayesian curve-fitting approach and used the posterior expectation of the parameters to draw the hit-rate curve. The threshold was determined from the docking score at which the hit rate reached a plateau.

Following the reviewer’s request, we calculated hit-rate curves separately for Drosophila melanogaster, Anopheles gambiae, Lepidoptera, and Locusta migratoria OR data (Fig. S5). Drosophila, Anopheles, and Lepidoptera showed similar trends: the hit-rate plateau was approximately 20%, and the corresponding docking-score threshold was around -8 kcal/mol. This is consistent with the original combined analysis and indicates that the threshold was not driven by a single species dataset. The locust dataset behaved differently. Its experimental positive rate was low, approximately 5.2%, and most reported locust ORs are narrowly tuned. In such a highly imbalanced dataset, docking score was less able to enrich experimental positives to the same level as in the other datasets; most score intervals had hit rates below 10%. When the locust data were combined with Drosophila, Anopheles, and Lepidoptera, the overall plateau decreased to about 18%, and the corresponding threshold shifted to about -11 kcal/mol.

We consider this difference informative because it shows that OR tuning properties and experimental positive rates can affect the relationship between docking score and hit rate. The locust result suggests that stricter thresholds may be required for systems dominated by narrowly tuned ORs and low positive rates. However, our goal is to establish a unified empirical threshold for large-scale repertoire comparison, not to optimize a separate threshold for every lineage. We therefore retain -8 kcal/mol from the Drosophila, Anopheles, and Lepidoptera datasets as the main threshold, present the locust analysis as a sensitivity test, and note that this threshold may have lower positive-enrichment ability in narrowly tuned OR systems.

The specific revision in the revised manuscript is as follows (lines 281-285): "In addition, our singletaxon hit-rate sensitivity analysis showed similar trends for Drosophila, Anopheles, and moths. Because the locust OR repertoire had an extremely low positive rate, we did not include the locust functional data in the final threshold assessment (see Supplementary Text S2 for details)."

Methods section (lines 748-755): "We first compiled published functional datasets containing explicit OR-VOC response relationships. These datasets included ORs from Drosophila melanogaster, Anopheles gambiae, Helicoverpa armigera, Spodoptera littoralis, and Locusta migratoria. For each dataset, OR-VOC combinations were assigned response or non-response labels according to the original experimental results. Each OR-VOC pair was then docked using the same structural modeling, docking, and scoring workflow used in the global OR-VOC analysis. Pairs were grouped into docking-score intervals, and the hit rate for each interval was defined as the proportion of experimentally positive pairs among all pairs in that interval. Each score interval was sampled three times."

(7) It is not clear how the ROC curves were constructed; they appear to be drawn with very few points. The methodology needs to be explained in more detail here, since this is an important point.

We thank the reviewer for this suggestion. We have added a detailed description of ROC construction. The ROC analysis was designed to evaluate whether the dFunC labels are directionally consistent with available experimental functional data.

Specifically, a positive dFunC label for a functional-group VOC indicates that ORs in that docking-derived community are predicted, relative to the all-insect OR background, to bind more VOCs from that functional group. A negative label indicates a lower predicted binding tendency. We mapped Drosophila ORs to the dFunC classification and assigned each OR a positive or negative label for each functional group according to its dFunC. We then used the Drosophila experimental functional matrix to count measured responses of each OR to different functional-group VOCs. For each functional group, we plotted ROC curves and calculated AUC values by comparing the dFunC-predicted labels with experimental response counts. We agree that the curves are based on limited data points for some functional groups. This is because comprehensive OR-VOC functional matrices are still sparse, and many insect ORs have been deorphanized using limited odor panels that do not cover many functional groups.

The specific revision in the revised manuscript is as follows (lines 794-798): "Drosophila melanogaster ORs were mapped to dFunCs to evaluate the consistency between dFunC labels and experimental response data. For each VOC functional group, ORs were assigned positive or negative labels according to their dFunC labels, and these labels were compared with response counts from the Drosophila functional matrix to generate ROC curves and AUC values."

(8) What the BBI represents is not very clear, even though the calculation method is presented in detail. This should be clarified in the main text and/or in the figure legends.

We thank the reviewer for this suggestion. We now define BBI more explicitly in the main text. BBI, or binding breadth index, is a relative repertoire-level metric calculated from docking-derived dFunC labels. A dFunC label indicates whether ORs in a docking-derived binding-profile community show higher or lower predicted binding tendency toward a functional-group VOC category relative to the all-insect OR background. When we calculate BBI for a set of ORs, we summarize the predicted binding tendencies of the dFunCs to which those ORs belong, thereby estimating the relative potential binding breadth of that OR repertoire for the VOC category.

Thus, BBI should be interpreted only as a relative docking-derived metric. When comparing two OR groups for the same VOC functional group, a higher BBI means that the group is predicted, under our docking-derived framework, to have broader potential binding breadth. It does not directly represent experimentally confirmed olfactory breadth, odor perception, or behavioral function. We have revised the terminology accordingly to docking-derived potential binding breadth.

The following revision has been added to the revised manuscript (lines 319-321): "BBI is a docking-based, repertoire-level relative metric used to summarize the potential binding breadth of a set of ORs toward a given functional-group VOC category."

(9) Insects with saprophagous larvae in your dataset are almost exclusively flies. Therefore, the conclusions concerning the saprophagous diet could just as well result from phylogenetic constraints rather than from an adaptation to the diet.

We thank the reviewer for pointing this out. We agree that, because the currently available saprophagous larval species are mainly concentrated in Diptera, this result may be influenced by lineage effects. We have therefore toned down the interpretation and present it as a candidate association. Future inclusion of more saprophagous species from non-dipteran lineages will allow this relationship to be tested more rigorously.

The following revision has been added to the revised manuscript (lines 493-496): "This interpretation should be treated with caution, as the saprophagous larval species currently available in our dataset are mainly concentrated in Diptera and may therefore reflect lineage effects. Future inclusion of saprophagous species from broader insect lineages will help test this association more rigorously."

Reviewer #2 (Recommendations for the authors):

Main feedback

(1) This study presents an interesting approach to unify the exceptionally large OR multigene family in a single macroevolutionary framework. While this is a non-trivial task, given the low sequence similarity of ORs, the inferences broadly overlap with previous findings, suggesting that the approach works. Unfortunately, the line between previously identified patterns of OR biology and evolution and new insights derived from this study is somewhat blurry throughout the text at the moment and needs to be sharpened. As it reads, the manuscript is prone to overselling the results.

A few examples:

(a) Lines 37-41 - contrary to the claim in this sentence, previous work has described the principles of OR function, structure, and evolution to an extent that would warrant concluding that the fundamental logic of insect olfaction is actually comparatively well understood.

We thank the reviewer for this comment. We agree that previous studies have already established important principles of insect OR biology, including OR-Orco complex structure, ion-channel function, ligand recognition, and rapid OR evolution. Our intention was not to imply that the basic logic of insect olfaction remains unknown, but to emphasize that an insect-class-scale framework integrating OR sequence divergence, structural variation, and docking-derived binding-profile diversity has been lacking. We have revised the Introduction to clarify this point and to better acknowledge the contribution of previous studies.

We therefore revised this sentence in the revised manuscript as follows (lines 38-42): "However, a classwide integrative framework linking OR sequence divergence, structural variation, and docking-derived binding-profile diversity is still lacking. Such a framework is needed to distinguish broadly conserved patterns from lineage-specific features and to place species-level OR diversification in a broader macroevolutionary context."

(b) Lines 47-48 - work in several systems has emphasized the "evolutionary interplay between receptor family diversification and macroecological adaptation", both in species with highly specialized ecologies (plant host specialists, slave making ants -> these are even cited in the manuscript) and in a broader macroevolutionary context (e.g. bee diet breadth, Singh et al. 2025), which does not really justify the claim that OR evolution with respect to ecology is a profound enigma. The present study indeed presents the first one combining data throughout the entire insect tree of life, which is an impressive feat. The fact that inferences drawn from smaller datasets with smaller phylogenetic breadth are supported is a useful finding. I recommend emphasizing this.

We agree that the relationship between OR repertoire evolution and ecological adaptation has been investigated in multiple insect systems and should not be described as an unresolved “profound enigma.” The novelty of our study is to test, integrate, and extend these previously reported patterns across 115 insect species under a unified sequence-structure-docking profile framework. We have revised the text to emphasize this class-wide extension rather than overstating the unknowns in the field. The following revision has been added to the revised manuscript (lines 46-47): "Although several studies have linked OR repertoire evolution to ecological adaptation in specific insect lineages, a unified class-wide comparison remains lacking."

(c) Lines 347 - 348 - what is meant by the statement that the manuscript 'revealed the mechanism of insect olfactory perception underlying macroenvironmental influences' is unclear to me. What mechanism is referred to here? What do the authors mean by 'macroenvironmental influences'? It reads as if the authors claim to have shown how olfactory perception works in insects, which is not an accurate statement.

We thank the reviewer for pointing out that this statement was unclear. We did not intend to claim that our study reveals the mechanism of insect olfactory perception itself. Rather, our results identify class-scale associations between OR repertoire variation, ecological traits, and docking-derived binding profile composition. We have revised the Discussion.

In the revised manuscript, this sentence has been changed to the following more accurate wording (lines 399-402): "Our results reveal class-level associations between OR repertoire variation and ecological traits and identify differences in docking-derived binding-profile composition among extant insect lineages that may be relevant to long-term macroevolutionary transitions."

(d) Lines 348-350: I do not agree with the statement that the authors "describe the evolutionary process through which Orco emerged". The data that the beta-sheet in EL2 impedes ligand binding in non-orco ORs is compelling and suggests an intriguing contribution to why Orco lack ligand-binding function, but is not sufficient to comprehensively describe the evolutionary process of Orco evolution from ancestral ORs. This would require more sequences of early-branching insect species. Accordingly, this statement should be toned down.

We agree that the original wording overstated the extent to which our data describe Orco emergence. Our results identify structural features, including the EL2 beta-sheet and specialized binding-pocket properties, that may have contributed to the loss of ligand-binding function during Orco specialization. However, a complete reconstruction of Orco origin will require broader sampling of early-diverging insect lineages and more reliable ancestral-state reconstruction. We have revised the relevant statements accordingly.

We therefore toned down this statement in the revised manuscript and described it more accurately as follows (lines 402-405): "We further examined Orco specialization within the insect 'conserved chassisdiversified sensors' model and identified structural features, including the EL2 beta-sheet and specialized binding-pocket properties, that may have contributed to the loss of ligand-binding function during Orco evolution." We also revised another sentence as follows (lines 522-525): "Overall, we systematically explored the relationships among sequence, structure, and docking-derived function in insect ORs, identified structural features potentially associated with Orco specialization, and provided a framework for testing how OR repertoire evolution may relate to macroenvironmental and lifestyle variation."

It is understandable that the authors want to emphasize the value of their work, but this can't be achieved by inaccurately portraying the state of the field. The framing should be adjusted throughout the manuscript.

We thank the reviewer for this important suggestion. We agree that the manuscript should emphasize its contribution without overstating gaps in the field. We have therefore revised the framing throughout the manuscript to better acknowledge previous advances in insect OR structure, function, evolution, and ecological adaptation. The revised text presents our main contribution as an insect-classscale integration and extension of these findings using a unified sequence, structure, and docking-derived binding-profile framework.

(2) It is unclear how complete the odorant receptor sets are, because the annotation pipeline used is not benchmarked. The high similarity of ORs often located in clusters of up to 50 genes on the genome represents a major obstacle for purely automated annotation, leading to artifacts such as the erroneous joining of exons across genes. The lack of manual verification in this study, combined with stringent filtering of annotations that do not meet structural criteria, does likely lead to a set of sequences that present a correct fold, but at the same time, the approach may discard misannotated genes and cannot distinguish between correct gene models and artifacts where exons of more than one gene could have been merged. Please benchmark the pipeline by comparing its output to a gold standard and frequently vetted complete OR set, such as that of Robertson and Wanner 2006 or similar. That would allow assessing the presented work better and reveal how complete the OR counts are. This is important since the authors use OR counts to derive conclusions about the evolutionary dynamics of the gene family.

We thank the reviewer for raising this concern. Because this point overlaps with the benchmark issue raised in the public review, we provide the detailed benchmark results in our response above and in the revised supplementary text. Briefly, we compared the annotation and structural-filtering pipeline with curated OR repertoires from Drosophila melanogaster, Anopheles gambiae, and Bombyx mori. The results show that the workflow recovers most known ORs, while the structural filter removes fragmented or structurally incomplete models. We also added sensitivity analyses using published OR counts where available, which supported the main OR-count ecological associations.

(3) The in silico functional docking analysis is intriguing and bears the potential to produce testable hypotheses on the functional evolution of ORs. Docking can produce a wide range of results, including artifacts, and thus should be benchmarked with existing datasets derived from functional experiments (there are several large datasets available). While the authors have used previous functional data to constrain their model, they should benchmark it by running it on a set of ORs with known ligand-binding profiles to convincingly show that their predictions are an accurate assessment of the actual functional properties of ORs.

We thank the reviewer for this suggestion. We agree that docking-derived results should be evaluated using ORs with known ligand-response data. In this study, however, the key prediction used for downstream analyses is the dFunC functional label derived from each OR’s overall docking-derived binding profile, rather than the exact ligand assignment of each OR-VOC pair. Therefore, we evaluated the dFunC labels using the Drosophila melanogaster experimental OR response matrix.

Drosophila ORs were mapped to dFunC communities, and for each VOC functional group, ORs were assigned positive or negative labels according to the predicted binding tendency of their dFunC. These labels were compared with experimentally measured response counts from the Drosophila OR functional matrix. ROC curves and AUC values were calculated for each VOC functional group. Most functional-group labels achieved AUC values above 0.75, indicating that the dFunC classification is broadly consistent with the Drosophila experimental response matrix.

We did not use individual OR-VOC pair accuracy as the primary validation criterion because the hit rate analysis showed that the experimental positive rate in the enriched docking-score range is approximately 23%. This indicates that docking is more appropriate for evaluating dFunC-level binding profile tendencies than for exact ligand assignment. We have revised the manuscript to clarify this validation logic. The functional labels of dFunCs should be further tested with experimental response data from more insect species and can be updated as prediction methods improve.

The following details have been added to the revised manuscript (lines 794-798): " Drosophila melanogaster ORs were mapped to dFunCs to evaluate the consistency between dFunC labels and experimental response data. For each VOC functional group, ORs were assigned positive or negative labels according to their dFunC labels, and these labels were compared with response counts from the Drosophila functional matrix to generate ROC curves and AUC values."

(4) The authors conclude that holometabolous and non-holometabolous insects exhibit distinct OR differentiation patterns, stating that sequence diversity "increases progressively with increasing degree of insect order divergence, suggesting a gradual accumulation of sequence variation during insect evolution" (Lines 332-333). If the pattern is indeed gradual, is the phylogenetic splitting of holometabolous vs. non-holometabolous insects not arbitrary? Couldn't it be split at any point along the phylogeny and lead to a similar difference?

We thank the reviewer for pointing out this important logic issue. We agree that, if OR sequence variation accumulates gradually with phylogenetic distance, the difference between holometabolous and non-holometabolous insects should not be interpreted as two completely separate OR differentiation modes.

Our result is better understood as a comparison of extant deep insect lineages. In this comparison, holometabolous insects show higher OR sequence-community diversity than non-holometabolous insects. This pattern is still informative because it suggests that, during long-term evolution, OR repertoires in holometabolous lineages have accumulated broader sequence-community diversity. This may be related to lineage-specific OR duplication and loss, ecological diversification, life-history differences, and differences in repertoire size.

We have revised the manuscript accordingly. We no longer describe the result as two distinct OR differentiation patterns. Instead, we describe it as a difference in OR sequence-community diversity between extant holometabolous and non-holometabolous lineages.

The following revision has been added to the revised manuscript (lines 370-371): "Additionally, extant holometabolous insects showed higher OR sequence-community diversity than non-holometabolous

insects."

(5) Further, if I understand correctly, the assessment of the differentiation patterns seems to rest on total counts of ORs in 'SeqCs', quantified in Figure S9. These counts, however, are not normalized and thus seem influenced by the total number of sequenced genomes belonging to a specific clade in the dataset. These have a heavy bias towards holometabolous species. If this is true, the analyses should be repeated based on normalized counts. OR is this assessment based on the number of 'SeqCs' across the phylogeny? If yes, does this correlate with OR numbers? Please clarify.

We thank the reviewer for pointing this out. We agree that unnormalized SeqC counts can be affected by unequal sampling, because holometabolous insects include more sampled species and more ORs in our dataset.

To address this issue, we added normalized analyses. First, we calculated SeqC diversity at the species level (Fig. S9a). Second, because OR number was significantly correlated with SeqC count (pGLS p-value = 0.001035), we performed equal-OR random sampling (Fig. S9 b). Specifically, we randomly sampled the same number of ORs from holometabolous and non-holometabolous insects, 100 ORs per group, calculated how many SeqCs were covered by the sampled ORs, and repeated this procedure 100 times to obtain the distribution of SeqC diversity under equal OR sampling depth. These analyses showed that holometabolous insects still exhibited higher OR sequence-community diversity after controlling for OR sampling depth. We have revised the manuscript to clarify the calculation and now interpret the result as a normalized difference in sequence-community diversity, rather than as an uncorrected difference in total SeqC number.

We added the following normalized analysis to the revised manuscript (lines 371-379): "Because total SeqC counts may be affected by unequal species sampling and OR numbers, we added normalized analyses. First, we calculated SeqC numbers at the species level and found that holometabolous insects contained more SeqCs on average. Second, because OR number was significantly correlated with SeqC count (P=0.001035), we performed equal-OR random sampling between holometabolous and nonholometabolous insects. Under the same OR sampling depth, holometabolous insects still covered more SeqCs, indicating higher OR sequence-community diversity after controlling for OR number. Therefore, this result supports a genuine difference in SeqC diversity rather than a difference driven by unequal species sampling or OR number. "

(6) Phylogenetic differences in OR evolutionary patterns have been described previously for Paleoptera compared to Neoptera. Are the authors simply picking up on these patterns? Is the holometabolous vs non-holometabolous hypothesis supported when only Neoptera are taken into account?

We thank the reviewer for raising this point. Differences in OR evolutionary patterns between Paleoptera and Neoptera have already been described in previous studies, so it was necessary to test whether our pattern simply reflected this older phylogenetic division.

Following the suggestion, we restricted the comparison to Neoptera and compared Holometabola with non-holometabolous Neoptera. After excluding Paleoptera and earlier-diverging lineages, Holometabola still showed higher OR sequence-community diversity (Fig. S10). Additional normalized analyses indicate that this trend is not explained only by unequal species sampling or OR counts (Fig S9a-b).

Therefore, our result is not simply a rediscovery of the known Paleoptera versus Neoptera difference. Instead, within Neoptera we observe an additional difference in SeqC diversity between Holometabola and non-holometabolous Neoptera.

We added the following comparison of neopteran insects to the revised manuscript (lines 379-382): "The higher SeqC diversity observed in Holometabola was retained when the comparison was restricted to Neoptera, indicating that this pattern is not simply a restatement of previously reported Paleoptera-Neoptera differences."

Minor comments

(1) Line 10: "all the insect orders were found to contain fully functional OR repertoires". It is unclear what the authors mean by this. What is a fully functional OR repertoire? Is it the number of receptors? The number of functional receptors? Given this lack of clarity and the fact that the manuscript does not present functional data on Or binding profiles, this part should be altered.

We have revised this sentence. The intended meaning was that all sampled insect orders include ORs assigned to all six docking-derived functional communities.

(2) Lines 17-18: I am not convinced the presented data explain the adaptive relationship between insect olfactory potential and diverse ecological environments. It suggests that few aspects of insect ecology (larval diet and terrestriality) correlate with receptor number and predicted ligand binding breadth. This should be toned down.

We have toned down the statement. The sentence has been revised as follows: " Our findings provide a class-level framework for investigating insect OR evolution and generate testable hypotheses about how receptor repertoire diversification may relate to ecological adaptation."

(3) Line 44: the low amino acid sequence identity among ORs has been described before. Please add appropriate references.

We have added the requested references.

(4) Lines 52 - 55: The 1:3 stoichiometry is not yet confirmed in vivo - indeed, other work suggests a 2:2 stoichiometry as a possibility. Please adjust this sentence to reflect this uncertainty.

We have revised the sentence to reflect this uncertainty (lines 51-56): Unlike vertebrate G protein-coupled ORs, insect ORs function as ligand-gated ion channels by assembling with the conserved co-receptor Orco, with recent cryo-EM structures supporting a 1:3 OR–Orco heterotetrameric model. However, the exact in vivo stoichiometry remains to be fully resolved, and alternative arrangements such as 2:2 may also be possible.

(5) Lines 94-95: Please cite studies that previously annotated ORs.

We have added the relevant citations.

(6) Lines 99-100: It is well established that hymenopterans have the largest sets of ORs among insects. Please add appropriate references.

We have added the relevant citations.

(7) Lines 101-102: It has been shown before that Odonata have very few ORs. However, there are other insects with even smaller sets of ORs. Please add appropriate references.

We have clarified that the statement refers only to our intact OR dataset. The sentence has been revised as follows: "In our OR dataset, Odonata had the fewest intact ORs, with a mean of 4 (N = 3)."

(8) Lines 109-111: This sentence appears to be illogical. Please revise.

We re-evaluated analyses that depend on OR counts. Because the association between genome size and OR count was sensitive to annotation completeness, structural filtering, and a few extreme species, we removed this result and the related discussion from the revised manuscript.

(9) Line 113: What is ecological behavior? Do you mean ecology?

We have replaced ecological behavior with ecological traits.

(10) Lines 148 - 150: This has been described previously. Please add appropriate references.

We have added the appropriate citation.

(11) Lines 161-162: This has been previously described. Please add appropriate references.

We have added the appropriate citation.

(12) Lines 199-201: Could it be that a lack of orthologs is a result of sampling bias? Only very few genomes of early-branching lineages have been sequenced.

We thank the reviewer for this helpful comment. We agree that the apparent lack of direct orthologs among early-diverging insect ORs may be influenced by limited genome sampling from these lineages. We have revised the manuscript to clarify that, with the currently available data, we cannot reliably determine direct orthologous relationships among early-diverging insect ORs or reconstruct ancestral OR states. We now state that broader sampling of high-quality genomes and transcriptomes from early-diverging insect lineages will be needed to test this question more rigorously.

The revised text is as follows (lines 218-220): "With the currently available early-diverging insect genomes, we could not reliably identify direct orthologous relationships among early ORs, which limits the reconstruction of ancestral OR states."

(13) Line 204-205: Do you mean TdomOR1-8 may be ancestral to Orco? This is not possible, given that Orco and TdomOR1-8 co-exist in one genome. But previous phylogenetic analyses of multiple silverfish genomes suggest that Orco is closely related to TdomOR1-8 and similar receptors in other species. Please revise.

We thank the reviewer for pointing this out. We did not mean that TdomOR1-8 are the ancestors of TdomOrco. Because currently available OR data from early-diverging insects are limited, direct orthologous relationships cannot be reliably identified and the ancestral OR state of Zygentoma cannot be reconstructed with confidence. Therefore, we can only search among extant zygentoman ORs for comparators that are closely related to the Orco clade.

The current gene tree shows that TdomOR1-8 are closely related to TdomOrco. Thus, they can serve as structural references for comparing Orco-related receptors and for understanding structural changes that may have been involved in Orco specialization.

In the revised manuscript, we have changed the relevant sentence to (lines 223-225): “Therefore, we speculate that Orco recruitment in the ancestral zygentoman lineage may have been accompanied by expansion of a set of candidate ORs, and that TdomOR1-8 represent extant retained members of this receptor set.”

(14) Lines 387-388, this was already known. Please cite accordingly

We have added the appropriate citation.

(15) Line 389: Please include a reference for the finding that early ORs have a homologous tetrameric configuration.

We have added the appropriate citation.

(16) Lines 403-407: This hypothesis has been previously put forward by del Marmol and colleagues. Add reference.

We have added the appropriate citation.

Structurally erroneous ORs in Propsilocerus akamusi.

Sensitivity analysis of OR counts.

(a-b) Correlations between updated species-level OR counts and genome size.

OR protein structures removed by structural filtering from the automated annotations of Drosophila melanogaster and Anopheles gambiae.

Normalized analysis of SeqC diversity between holometabolous and non-holometabolous insects within Neoptera at the species level (a) and OR-count level (b).

Welch's t-test was used. ***p < 0.001.

Reference

Bobkov YV, Walker Iii WB, Cattaneo AM. 2021. Altered functional properties of the codling moth Orco mutagenized in the intracellular loop-3. Sci Rep 11: 3893.

Carey AF, Wang G, Su CY, Zwiebel LJ, Carlson JR. 2010. Odorant reception in the malaria mosquito Anopheles gambiae. Nature 464: 66-71.

Chang H, Unni AP, Tom MT, Cao Q, Liu Y, Wang G, Llorca LC, Brase S, Bucks S, Weniger K et al. 2023. Odorant detection in a locust exhibits unusually low redundancy. Curr Biol 33: 5427-5438 e5425.

Dekel A, Pitts RJ, Yakir E, Bohbot JD. 2016. Evolutionarily conserved odorant receptor function questions ecological context of octenol role in mosquitoes. Sci Rep 6: 37330.

Frunze O, Lee D, Lee S, Kwon HW. 2024. A single mutation in the mosquito (Aedes aegypti) olfactory receptor 8 causes loss of function to 1-octen-3-ol. Insect Biochem Mol Biol 167: 104069.

Guo S, Kim J. 2007. Molecular evolution of Drosophila odorant receptor genes. Mol Biol Evol 24: 11981207.

Hall DR, Beevor PS, Cork A, Nesbitt BF, Vale GA. 1984. 1-Octen-3-ol. International Journal of Tropical Insect Science 5: 335-339.

Kline DL, Allan SA, Bernier UR, Welch CH. 2007. Evaluation of the enantiomers of 1-octen-3-ol and 1octyn-3-ol as attractants for mosquitoes associated with a freshwater swamp in Florida, U.S.A. Med Vet Entomol 21: 323-331.

Kreher SA, Mathew D, Kim J, Carlson JR. 2008. Translation of sensory input into behavioral output via an olfactory system. Neuron 59: 110-124.

Liu WB, Li HM, Wang GR, Cao HQ, Wang B. 2023. Conserved Odorant Receptor, EcorOR4, Mediates Attraction of Mated Female Eupeodes corollae to 1-Octen-3-ol. J Agric Food Chem 71: 1837-1844.

Lu T, Qiu YT, Wang G, Kwon JY, Rutzler M, Kwon HW, Pitts RJ, van Loon JJ, Takken W, Carlson JR et al. 2007. Odor coding in the maxillary palp of the malaria vector mosquito Anopheles gambiae. Curr Biol 17: 1533-1544.

Morawo T, Fadamiro H. 2016. Identification of Key Plant-Associated Volatiles Emitted by Heliothis virescens Larvae that Attract the Parasitoid, Microplitis croceipes: Implications for Parasitoid Perception of Odor Blends. J Chem Ecol 42: 1112-1121.

Paddock KJ, Corcoran JA. 2025. Life-stage dependent behavior mimics chemosensory repertoire diversity in a belowground, specialist herbivore. G3 (Bethesda) 15.

Pierce A, Pierce H, Borden J, Oehlschlager C. 1989. Production Dynamics of Cucujolide Pheromones and Identification of I-Octen-3-o1 as a New Aggregation Pheromone for Oryzaephilus surinamensis and O. mercator (Coleoptera: Cucujidae). Environmental Entomology 18: 747-755.

Qiu L, Tao S, He H, Ding W, Li Y. 2018. Transcriptomics reveal the molecular underpinnings of chemosensory proteins in Chlorops oryzae. BMC Genomics 19: 890.

Rondoni G, Roman A, Meslin C, Montagne N, Conti E, Jacquin-Joly E. 2021. Antennal Transcriptome Analysis and Identification of Candidate Chemosensory Genes of the Harlequin Ladybird Beetle, Harmonia axyridis (Pallas) (Coleoptera: Coccinellidae). Insects 12.

Tanaka K, Uda Y, Ono Y, Nakagawa T, Suwa M, Yamaoka R, Touhara K. 2009. Highly selective tuning of a silkworm olfactory receptor to a key mulberry leaf volatile. Curr Biol 19: 881-890.

Tian Z, Sun L, Li Y, Quan L, Zhang H, Yan W, Yue Q, Qiu G. 2018. Antennal transcriptome analysis of the chemosensory gene families in Carposina sasakii (Lepidoptera: Carposinidae). BMC Genomics 19: 544.

Turner RM, Derryberry SL, Kumar BN, Brittain T, Zwiebel LJ, Newcomb RD, Christie DL. 2014. Mutational analysis of cysteine residues of the insect odorant co-receptor (Orco) from Drosophila melanogaster reveals differential effects on agonist- and odorant-tuning receptor-dependent activation. J Biol Chem 289: 31837-31845.

Wang Q, Smid HM, Dicke M, Haverkamp A. 2024. The olfactory system of Pieris brassicae caterpillars: from receptors to glomeruli. Insect Sci 31: 469-488.

Xu L, Jiang HB, Yu JL, Pan D, Tao Y, Lei Q, Chen Y, Liu Z, Wang JJ. 2023. Two odorant receptors regulate 1-octen-3-ol induced oviposition behavior in the oriental fruit fly. Commun Biol 6: 176.

Xu P, Zhu F, Buss GK, Leal WS. 2015. 1-Octen-3-ol - the attractant that repels. F1000Res 4: 156.

Xu Q, Wu Z, Zeng X, An X. 2020. Identification and Expression Profiling of Chemosensory Genes in Hermetia illucens via a Transcriptomic Analysis. Front Physiol 11: 720.

Yan C, Sun X, Cao W, Li R, Zhao C, Sun Z, Liu W, Pan L. 2020. Identification and expression pattern of chemosensory genes in the transcriptome of Propsilocerus akamusi. PeerJ 8: e9584.

Zhang S, Zhang Z, Wang H, Kong X. 2014. Antennal transcriptome analysis and comparison of olfactory genes in two sympatric defoliators, Dendrolimus houi and Dendrolimus kikuchii (Lepidoptera: Lasiocampidae). Insect Biochem Mol Biol 52: 69-81.

Zhang Y, Wang B, Zhou Y, Liao M, Sheng C, Cao H, Gao Q. 2023. Identification and characterization of odorant receptors in Plutella xylostella antenna response to 2,3-dimethyl-6-(1-hydroxy)-pyrazine. Pestic Biochem Physiol 194: 105523.

Version 2

Abstract

Insect olfaction is facilitated by a heterotetrameric odorant receptor–odorant receptor co-receptor (OR-Orco) complex, which is distinct from that of vertebrate ORs. However, extreme sequence divergence among insect ORs has hindered a unified understanding of their evolutionary history and ecological importance. In this study, we present a multiscale analysis of OR genes across 115 insect species. We overcome the limitations of traditional phylogenetic approaches by applying a protein similarity network-based strategy and introduce a "trunk–branch" framework to systematically describe the evolutionary trajectories of insect ORs across sequence, structural, and predicted olfactory potential levels. Although they possess different sequences and structural communities, all sampled insect orders contained ORs assigned to all six docking-derived functional communities. Notably, extant insect lineages with order-level origins spanning the end-Permian mass extinction (EPME) differed in docking-derived functional OR repertoires, and early-and late-diverging lineages differed in OR sequence-community diversity. The emergence of Orco represents a key evolutionary transition point, marking the shift from a homomeric to a heteromeric complex accompanied by specialization of the extracellular domain and binding pocket. Furthermore, we established robust associations between potential olfactory recognition breadth and ecological variables, including diet, circadian rhythm, and habitat. Our findings provide a class-level framework for investigating insect OR evolution and generate testable hypotheses about how receptor repertoire diversification may relate to ecological adaptation.

Introduction

Olfactory systems play a critical role in animal survival strategies, governing critical behaviors from nutrient acquisition to mate selection(Bargmann 2006; Nei et al. 2008; Kaupp 2010). Within the kingdom Animalia, insects—the most speciose metazoan group—have evolved exceptionally sophisticated chemosensory systems to interpret complex volatile landscapes(Nei et al. 2008; Kaupp 2010; Mora et al. 2011; Robertson 2019). The extraordinary sensitivity of insect olfaction to hydrophobic volatile organic compounds (VOCs) is facilitated by a heterotetrameric channel complex consisting of odorant receptors (ORs) and odorant receptor co-receptors (Orco)(Clyne et al. 1999; Vosshall et al. 1999; Sato et al. 2008). Insect ORs represent a distinct protein superfamily that originated independently of vertebrate ORs, with no shared homology between the two(Sato et al. 2008; Hansson and Stensmyr 2011). Studies have shown that this ancient structural fold has a deeper phylogenetic origin(Benton and Himmel 2023; Himmel et al. 2023). While vertebrate ORs have been extensively studied(Niimura and Nei 2005; Hughes et al. 2018; Pacalon et al. 2023; Policarpo et al. 2024), the extremely low sequence similarity among insect ORs has made resolving their internal sequence, structural, and functional differences challenging(Robertson and Wanner 2006). Extensive studies have reported advances in the olfactory recognition mechanism and OR function(Guo et al. 2021; Chang et al. 2023; Depetris-Chauvin et al. 2023; Jiang et al. 2024), which have greatly facilitated a deeper understanding of insect olfactory perception. However, a class-wide integrative framework linking OR sequence divergence, structural variation, and docking-derived binding-profile diversity is still lacking. Such a framework is needed to distinguish broadly conserved patterns from lineage-specific features and to place species-level OR diversification in a broader macroevolutionary context.

Within the class Insecta, Orcos exhibit exceptional conservation (>400 Mya evolutionary stability), whereas lineage-specific ORs display unprecedented diversity, far exceeding that of vertebrate ORs(Larsson et al. 2004; Nei et al. 2008; Brand et al. 2018; Robertson 2019; Wicher and Miazzi 2021). Although several studies have linked OR repertoire evolution to ecological adaptation in specific insect lineages, a unified class-wide comparison remains lacking(Jongepier et al. 2022; Yin et al. 2022; Singh et al. 2025). Traditional phylogeny-driven subfamily categorization, constrained by extreme cross-order sequence divergence, fails to capture the continuum of OR evolutionary innovation(Zhou et al. 2012; Zhou et al. 2015; Mitchell et al. 2020; Legan et al. 2021; Tian et al. 2022; Yin et al. 2022). Unlike vertebrate G protein-coupled ORs, insect ORs function as ligand-gated ion channels by assembling with the conserved co-receptor Orco, with recent cryo-EM structures supporting a 1:3 OR–Orco heterotetrameric model. (Jones et al. 2005; Nei et al. 2008; Kaupp 2010; Jones et al. 2011; Wang et al. 2024; Zhao et al. 2024). However, the exact in vivo stoichiometry remains to be fully resolved, and alternative arrangements such as 2:2 may also be possible(Wang et al. 2024). The mechanistic drivers of this functional divergence raise key evolutionary questions: What were the key factors that drove the emergence of Orco and the formation of the insect-specific OR–Orco olfactory perception model, and what advantages did it provide?

Identifying the binding profiles of ORs for VOCs is fundamental to understanding how insect olfactory systems perceive and adapt to their environments(McBride et al. 2014; Auer et al. 2020). To date, numerous studies have characterized the ligand-binding profiles of various insect ORs using heterologous expression systems(Hallem and Carlson 2006; Carey et al. 2010; de Fouchier et al. 2017; Guo et al. 2021; Chang et al. 2023; Jiang et al. 2024). However, when attempting to investigate the functional relevance of tens of thousands of ORs across the class Insecta at a macroevolutionary scale, experimental methods are no longer feasible due to their long timeframes and low throughput. Consequently, theoretical prediction has become the only viable approach. Machine learning approaches are limited by the size of available training datasets and exhibit poor scalability when applied across species on a large scale(Boyle et al. 2013; Chepurwar et al. 2019; Caballero-Vidal et al. 2020). Therefore, molecular docking remains the mainstream approach for ligand screening(Lyu et al. 2019; Comte et al. 2025). This strategy has already proven successful in mammals: by employing large-scale molecular docking and OR binding profile prediction, researchers have revealed differences in olfactory perception among bats with different diets, as well as dynamic trade-offs between vision and olfaction in primates(Zhang et al. 2024; Chi et al. 2025). These results suggest that molecular docking can facilitate the exploration of macroevolutionary patterns.

In this study, we reconstructed the evolutionary history of 450 million years of OR through a multilevel analysis of the OR superfamily of 115 insect species spanning 16 orders. Using a newly developed pipeline combining network analysis and protein structure and function prediction, we constructed a sequence similarity network (SSN), a structural similarity network (StSN), and a docking-derived functional similarity network (dFSN) for insect ORs and performed similarity community classification to investigate OR relationships from three distinct perspectives. The results revealed distinct OR repertoire patterns between holometabolous and non-holometabolous insects and EPME-spanning differences in dFunC composition among extant insect lineages. Moreover, the β-sheet structure of the extracellular loop 2 (EL2) region in Orco, along with its highly specialized ligand-binding pocket, plays a critical role in the loss of ligand-binding function and the evolutionary origin of Orco. In addition, we systematically revealed broad evidence of adaptive evolution linking the potential olfactory repertoire to different lifestyles at the scale of the class Insecta.

Results

Characterization of the insect odorant receptor gene repertoire

We compiled all published insect genomes (as of December 2022), selecting the highest-quality genome assembly (contig N50 > 1 Mb) from each family as the representative species. Our study includes 115 species, featuring 8 representative species from different insect orders with previously annotated ORs(Engsontia et al. 2008; Wang et al. 2015; Zhou et al. 2015; Brand et al. 2018; Robertson et al. 2019), spanning 16 orders and 111 families. This approach ensured a comprehensive analysis across the class Insecta (Fig. 1A). To increase annotation reliability, we implemented the most stringent protein superfamily member screening strategy to date (fig. S1A). A total of 8,905 structurally intact ORs were retained for downstream analyses. Benchmarking against existing OR repertoires from Drosophila melanogaster(Ozturk-Colak et al. 2024), Anopheles gambiae(Giraldo-Calderon et al. 2015), and Bombyx mori(Tanaka et al. 2009) showed that the annotation workflow provides high-confidence annotations suitable for downstream structural comparison and docking (see Supplementary Text S1, Table S1-3). Our analysis revealed substantial variation in OR counts across insect species. Hymenoptera exhibited the highest number of ORs and accounted for 8 of the top 10 species with the largest OR repertoires(Bonasio et al. 2010). In our OR dataset, Odonata had the fewest intact ORs, with a mean of 4 (N = 3). Hymenoptera also showed the widest range in OR counts (min: 38, max: 473, N = 23), followed by Coleoptera (min: 10, max: 271, N = 15) (Fig. 1B).

Evolutionary ecology of insect OR repertoires.

(A). Phylogeny of 115 insect species (see Methods and Data S4 for details). Genome size and the number of OR genes for each species are shown as bar plots. Ecological parameters, including larval and adult diets, habitat, and circadian rhythm, are displayed as labels with color coding corresponding to the legend below. Different geological periods are separated by shaded intervals. The tree was visualized using the iTOL(Letunic and Bork 2024). (B). Boxplot showing the number of OR genes for each insect order. (C-F). Results of two-sided pGLS tests assessing the relationship between intact OR gene numbers and various ecological parameters: adult diet (C), larval diet (D), circadian rhythm (E), and habitat (F). Dots represent OR counts of each species. N represents the counts of species. Animal silhouettes were obtained from PhyloPic.org.

Previous studies have shown that in certain insect orders and specific lineages, OR repertoire characteristics are associated with various ecological traits (McBride 2007; Liu et al. 2021; Schrader et al. 2021; Jongepier et al. 2022; Singh et al. 2025). To determine whether these relationships extend to the class Insecta, we analyzed four ecological factors: adult diet, larval diet, habitat, and circadian rhythm. Adult and larval diets represent stage-specific ecological traits, and our analysis of OR gene numbers at the genomic level aimed to reveal their relationship with long-term evolutionary pressures shaping OR gene family size. Our results revealed a significant correlation between insect OR counts and dietary preferences, both in adults (Fig. 1D, P = 0.61 × 10⁻⁶) and larvae (Fig. 1E, P = 0.03143). Habitat was also significantly associated with OR counts (P = 0.0354), with terrestrial insects exhibiting more ORs than aquatic insects do (Fig. 1F). However, there was no significant correlation between circadian rhythm and OR count (P = 0.7188, Fig. 1G). Overall, our findings suggest that insect OR counts correlate with dietary specialization (in both larvae and adults) and habitat type but are not associated with circadian rhythm. We then repeated the phylogenetic generalized least squares (pGLS) analyses using the larger of our structurally intact OR count and the published OR count, and the main ecological associations were retained (fig. S1 B-E). This indicates that the main OR-count ecological analyses are reasonably robust to moderate underestimation caused by missed ORs or strict structural filtering.

OR sequence similarity network

This study includes species representing the full phylogenetic diversity of the class Insecta, and their OR sequences exhibit substantial variation in amino acid similarity (min = 13.7%, max = 97.6%, and mean = 29.6%) (Fig. 2B). The traditional method of classifying subfamilies on the basis of bootstrap values of phylogenetic tree nodes suffers from reduced precision and accuracy. Therefore, we constructed an SSN on the basis of an "ALL vs. ALL" sequence alignment. In the SSN, each node represents a cluster of ORs sharing at least 60% sequence similarity, and edges are drawn between nodes if the average alignment E value < 1e-25 (fig. S2A). Nodes with a greater number of connections indicate closer evolutionary relationships.

Subfamily classification of ORs based on SSN.

(A). Schematic representation of the “trunk-branch” strategy pipeline for constructing the similarity network and classifying subfamilies. First, all ORs were subjected to greedy clustering based on sequence similarity to generate nodes. The average similarity between nodes was then calculated, and nodes were treated as units to identify core regions (trunks) using a strict similarity threshold. By gradually relaxing the threshold, new core regions and additional nodes (branches) emerged. After a final merging step based on defined criteria, the optimal community structure was determined using modularity as an index (see Methods). (B). Distribution of pairwise sequence similarity among all ORs. The red line indicates the mean similarity. (C). SSN of ORs constructed using the “trunk-branch” strategy. Each node represents ORs sharing >60% sequence identity. Edges represent node pairs with an average sequence alignment E-value of < 1 × 10⁻²⁵. Node colors indicate SeqC classifications. (D). Heatmap depicting the number of ORs within different SeqCs across various insect orders. Due to substantial differences in OR numbers among orders, the values are ln-transformed. The bar plot above the heatmap shows the number of SeqCs present in each order, while the bar plot on the right indicates the proportion of ORs within each SeqC generated by tandem duplication. Two ORs located within 5 kb of each other on the genome were defined as tandem duplicates. TD: Tandem Duplication.

Using the "trunk–branch" strategy (Fig. 2A), we identified 38 OR sequence communities (SeqCs), which together account for 94.9% of all ORs. Among them, 22 SeqCs demonstrated clear interaction relationships (Fig. 2C). The sequence similarity within each SeqC was predominantly between 20% and 30% (fig. S2B). By mapping previously classified Acromyrmex echinatior sequences using phylogenetic methods(Zhou et al. 2015), we found that most subfamilies—such as 9-exon, Orco, F, H, B, and A—were independently delineated (fig. S2C) and exhibited well-defined boundaries (fig. S2E). Our classification did not disrupt any known subfamilies and instead merged some families with minor differences (fig. S2, C and D). This comparison indicates that SSN-based communities are broadly consistent with established phylogenetic subfamilies, while capturing sequence-similarity modules rather than strictly monophyletic clades. We further evaluated the SSN using curated lepidopteran pheromone receptors (PRs). To distinguish different PR lineages, experimentally validated classic PRs and the recently reported new PR lineage were analyzed separately(Bastin-Heline et al. 2019). Classic PRs were predominantly assigned to SeqC6 (93.9%), whereas the new PR lineage was assigned to SeqC35 (fig. S2F), supporting the hypothesis of multiple independent origins of PRs(Bastin-Heline et al. 2019). These results further support the applicability and robustness of SSN-based community partitioning.

SeqC21 includes the Orco group as well as all ORs from Thermobia domestica and Machilis hrabei (fig. S2G), suggesting that ORs in early wingless insects share high sequence similarity with Orco(Brand et al. 2018). In addition to SeqC21, most SeqCs are order specific, indicating substantial sequence divergence among ORs from different insect orders (Fig. 2D). SeqC26 and SeqC28 are large shared clusters that may contain cross-order orthologous relationships. In SeqC26, we identified 56 orthologous groups, four of which spanned multiple insect orders. OGG54 included ORs from Orthoptera, Blattodea, Trichoptera, and Hymenoptera. This group contains the locust receptors LmigOR5 and LmigOR4. LmigOR5 has been reported to bind geranyl acetone and to be associated with avoidance behavior in the migratory locust(Chang et al. 2023). The functions of the corresponding receptors in other insect orders remain to be tested, and we now present these as candidate conserved modules rather than confirmed functional orthologs. In contrast, we did not identify clear cross-order orthologous groups in SeqC28.

Lepidoptera was found to contain the greatest number of SeqCs, reflecting the highest level of OR sequence divergence. Additionally, we identified SeqC25 and SeqC27 as two exceptionally large, Hymenoptera-specific groups, comprising 39.8% and 31.9% of all Hymenopteran ORs, respectively. SeqC27 corresponds to the previously identified Hymenopteran 9-exon family (fig. S2, C and D), which originated through tandem duplication events(Engsontia et al. 2015; Zhou et al. 2015; McKenzie et al. 2016; Legan et al. 2021). Interestingly, we found that SeqC25 exhibited an even higher rate of tandem duplication than did SeqC27, suggesting that SeqC25 also expanded through extensive tandem duplication. Notably, among SeqCs with tandem duplication rates exceeding 60%, 5 of 7 are exclusive to Hymenoptera, indicating that OR lineages in this order have undergone particularly intense tandem duplication expansion compared with other insect orders(Legan et al. 2021).

Structural group diversity within the ORs

Insect ORs fall within the "twilight zone" (10%–40% sequence identity) of sequence alignment. However, nearly all ORs remain within the "safe zone" in structural alignment (Fig. 3A).

StSN and structural community diversity of ORs.

(A). Sequence similarity versus structural similarity of insect ORs, assessed using pairwise comparisons with Dali. Each dot represents comparison between a pair of ORs, with color indicating point density. (B). StSN of ORs constructed using the “trunk-branch” strategy. Nodes represent ORs with pairwise Dali Z-scores >50. Edges connect nodes with an average pairwise Z-score of at least 42.9. Node colors denote StrCs classifications. (C). Structural comparison of ORs and Orco. The structures were aligned using US-align, with multiple structural alignments performed for annotated Orco and OR structures (visualized in PyMol). Colors represent pLDDT scores provided by AlphaFold2, indicating the confidence of structural modeling. (D). Visualization of differences among StrCs. The heatmap shows the number of ORs within different StrC across insect orders, with OR counts ln-transformed due to large inter-order variations. The bar plot above the heatmap indicates the number of StrC in each order. On the right, a ridge plot illustrates the flexibility rates of different structural regions across StrCs. Flexibility rate = 1 - conservation rate. Light gray represents transmembrane regions, white denotes loop regions, and dark gray marks the anchor domain of ORs. Colors correspond to StrCs, with gray indicating StrCs without interactions. (E). Two types of ORs in StrC17 — long IL3 ORs and general ORs — with the IL3 regions highlighted in color. (F). Phylogenetic analysis of StrC17 ORs based on amino acid sequences. Red circles indicate bootstrap values, with most ORs collapsed based on bootstrap support. Gray branches represent the clade of long IL3 ORs, and the dN/dS values are shown in blue on the corresponding evolutionary branches.

To investigate the structural diversity of insect ORs, we constructed a StSN, where each node represents ORs with Z score > 50 (from Dali(Holm 2022)). Edges were drawn between nodes if their average Z score was > 42.9 (fig. S3A). Similar to the SSN, we classified 28 structural communities (StrCs), which together accounted for 87.0% of all ORs. Among these, 13 StrCs exhibited distinct interaction relationships (Fig. 3B). The boundaries between StrCs were well defined (fig. S3C).

Most StrCs were also order specific (Fig. 3D). StrC1 represents the Orco group, which includes 8 T. domestica ORs. These ORs were previously identified as Orco-like genes through sequence analysis(Brand et al. 2018). Compared to ORs, Orco is characterized by two distinct structural features: a β-sheet in EL2 and a long loop in intracellular loop 2 (IL2) (Fig. 3C and fig. S3B). Lepidoptera presented the highest diversity of StrC types (Fig. 3D). We found that the transmembrane regions in different StrCs presented low variability, whereas the loop regions presented distinct variable patterns (Fig. 3D). In both ORs and Orcos, the IL2 region was identified as the most variable region and the only variable segment within the anchor domain. This high degree of variability may be associated with its role in stabilizing and regulating the anchoring domain(Bahk and Jones 2016; Jain et al. 2021; Wang et al. 2024; Zhao et al. 2024).

StrC23 is the only structural community present in 12 insect orders and accounts for 46.3% of all structurally intact ORs in the dataset. We found that StrC23 has a much larger binding-pocket volume than other ORs and Orco. From a structural perspective, this suggests that StrC23 receptors may have greater potential to accommodate diverse VOCs (fig. S3 D). This reflects conservation at the level of OR structural evolution: many species appear to retain a relatively large set of ORs with larger binding pockets, which may provide a structural basis for structure-derived volatile organic compound binding potential before species-specific OR diversification.

Notably, StrC17 exhibited a high degree of flexibility in the IL3 region. We observed that a subset of ORs within this group possessed unusually long IL3 regions (Fig. 3E). These ORs were distributed across 15 families from two suborders of Diptera, but they were absent from the earliest-diverging families in the Nematocera suborder, including Culicidae, Chironomidae, and Tipulidae (fig. S3, G and H). After manual inspection, annotation errors were ruled out. The construction of a maximum likelihood tree for StrC17 ORs revealed that ORs with long IL3 regions formed a distinct clade, with one subclade identified as orthologous (fig. S3F). To investigate the evolutionary pressures acting on these ORs, we calculated the dN/dS ratios for different branches (Fig. 3F). The results revealed that all StrC17 ORs were subject to strong purifying selection (dN/dS = 0.17). The amino acid substitution rate of ORs with long IL3 regions was two orders of magnitude lower than that of other ORs (dN/dS = 0.001–0.004), indicating intensified purifying selection.

Origin and driving factors of Orco

Orco originated in the ancestor of Zygentoma and evolved from ORs(Brand et al. 2018; Thoma et al. 2019). The evolutionary drivers underlying Orco specialization remain unresolved. With the currently available early-diverging insect genomes, we could not reliably identify direct orthologous relationships among early ORs, which limits the reconstruction of ancestral OR states. The T. domestica ORs TdomOR1-8 are classified within Orco communities (SeqC21 and StrC1) and occupy a transitional phylogenetic position between ORs and characterized Orcos, representing the closest ORs related to Orcos(Brand et al. 2018). Therefore, we speculate that Orco recruitment in the ancestral zygentoman lineage may have been accompanied by expansion of a set of candidate ORs, and that TdomOR1-8 represent extant retained members of this receptor set. We found that TdomOrcoFJ has the smallest, least polar, and most hydrophobic binding pocket among TdomOR1-8 (Fig. 4A). Molecular docking analysis revealed that TdomOrcoFJ is less suited for VOC binding than other candidates are (fig. S4A). These findings suggest that the extreme characteristics of the binding pocket are key features that distinguish TdomOrcoFJ from other candidate Orcos, with a reduced VOC-binding capacity serving as the primary driving force in Orco determination.

Drivers promoting the early origin of Orco.

(A). Differences in EL2, IL2, and binding pocket properties between M. hrabei and T. domestica ORs. Only ORs with closed binding pockets were analyzed, and all scores were normalized. Circles and triangles represent the minimum and maximum scores after column normalization, respectively. The presence of an EL2 β-sheet is indicated with a check mark. (B). Differences in amino acid properties of the EL2 region between OR and Orco. Hydrophilic amino acids are shown in blue, neutral amino acids in green, and hydrophobic amino acids in black (visualized by WebLogo(Crooks et al. 2004)). The x-axis numbers represent amino acid positions from the multiple sequence alignment, with highly gapped positions removed. (C). MD simulation-based analysis of the movement trajectory of VOC (geranyl acetate) near Orco and OR. The red protein represents ApisOR5, and the blue protein represents ApisOrco. The green area indicates the probability of VOC remaining above the binding pocket, while the pink represents its probability of occupying other regions of the protein (see Methods). Statistical significance was evaluated using a Chi-Squared Test. The red pentagram indicates the initial position of geranyl acetate in the simulation. (D). Changes in relative calcium fluorescence intensity (ΔF/F0) in cells. Cells were stimulated with 100 μM of the OR47a agonist pentyl acetate (PA) at 50 s and 100 μM the Orco agonist VUAA1 at 180 s. el2.DmOR47a: OR47a mutant in which the EL2 region of OR47a was replaced with that of Orco. mut.DmOr47a: the EL2 region of wild-type DmOr47a contains an inserted short peptide. The total number of cells (n = 30) was derived from three independent replicate experiments. The shading indicates the 95% confidence interval. (E). Relative calcium fluorescence intensity in cells following stimulation with PA and VUAA1, respectively. n=30. Statistical significance was determined using Tukey’s HSD test. (F). Schematic illustration of the mechanism by which the β-sheet structure in the EL2 region of Orco influences VOC proximity to the binding pocket. (G). Comparative analysis of binding pocket parameters (volume, hydrophobicity score, and polarity score) between Orco and ORs across all studied insect species. Statistical significance was determined using a two-sided t-test. **** P < 0. 0001; n.s., no significant.

The reasons behind the loss of Orco function remain unclear, although its conserved binding pocket may act as a molecular sieve(Pacalon et al. 2023). We found that M. hrabei ORs lack the EL2 β-sheet but have a long IL2 loop region. Similarly, most T. domestica ORs also have a long IL2 loop region, whereas only TdomOR1-8 retains the EL2 β-sheet (Fig. 4A). This finding suggests that the long IL2 loop region was conserved from ancient ORs but gradually disappeared during subsequent evolution. In contrast, the EL2 β-sheet likely emerged during Orco evolution and was retained in all subsequent Orcos.

The formation of the EL2 β-sheet may be associated with the loss of Orco function. We observed that the EL2 region of Orco is distinctly divided into two parts: the β-sheet region and the transmembrane loop region (Fig. 4B). In the β-sheet region, the proportion of hydrophilic amino acids is 47.8%, whereas it is 14.2% in the transmembrane loop region and 18% in ORs. Overall, across all the insects, the β-sheet region of Orco EL2 presented a significantly greater proportion of hydrophilic amino acids than did the ORs (fig. S4B), facilitating its stable presence in the extracellular aqueous environment. The extracellular port is the main pathway through which ligands enter the binding pocket(Renthal and Chen 2022; Pacalon et al. 2023). Using molecular dynamics (MD) simulations to analyze VOC movement outside the membrane, we found that the presence of the β-sheet in EL2 region significantly reduced the likelihood of VOCs reaching the binding pocket (Fig. 4C, fig. S4C and Movie S1). Using Drosophila OR47a as an example, we demonstrated through calcium imaging experiments that replacing the EL2 region of OR47a with that of Orco resulted in a mutant OR47a (el2. DmOR47a) that exhibited a markedly reduced ability to bind pentyl acetate (PA) (Fig. 4, D and E and fig. S4D). Additionally, we inserted a non–β-sheet short peptide into the EL2 region of OR47a, and the results showed that its function remained unaffected (Fig. 4, D and E and fig. S4D). This finding suggests that the β-sheet acts as a spatial barrier, limiting VOC access to the binding pocket as a primary physical gate (Fig. 4F). Furthermore, we observed that Orco has a smaller and less hydrophobic binding pocket than ORs do (Fig. 4G). These characteristics are unfavorable for VOC entry into the binding pocket, representing the key chemical factors contributing to Orco’s loss of natural VOC-binding function.

Functional labels for ORs on the basis of VOC recognition similarity

Currently, molecular docking and virtual screening have been successfully applied to the high-throughput prediction of OR binding profiles in mammals, revealing the macroevolutionary trajectories of species’ olfactory recognition abilities(Zhang et al. 2024; Chi et al. 2025). To better compare the potential binding profiles of insect OR repertoires and classify them accordingly, we collected 14,499 VOCs from 4 molecular libraries (including pheromones(Li et al. 2023), fragrances (the thegoodscentscompany.com), food odors(Chi et al. 2025), and theoretical odors(Mayhew et al. 2022)). These VOCs were categorized into 9 functional groups (Fig. 5A). Although terpenoids are important in insect chemical ecology, they are distributed across multiple regions of odor space (fig. S5C). To reduce the docking instability that could result from excessive VOC subdivision, we did not treat terpenoids as a category parallel to the nine basic functional-group categories. We used a comprehensive VOC library to systematically evaluate the theoretical functions of ORs, aiming to assess the olfactory potential encoded by the genome, even though some of these compounds may never be encountered by insects in nature.

Classification of insect OR docking-derived functional communities and the relationship between ecological parameters and the VOC recognition capability of ORs.

(A). 2D odor space based on molecule physicochemical properties. Each dot represents an odor molecule, with different colors indicating different functional groups. (B). Relationship between hit rate and docking score, inferred using Bayesian statistics. The black curve shows the distribution of hit rates and docking scores across all OR-VOC docking results. The top plateau has a hit rate of 22%, while the bottom plateau has a hit rate of 0%. Cyan dots represent the mean hit rate ± s.e.m. for each docking score interval. The orange curve indicates the peak hit rate at a docking score of -8 kcal/mol. The gray curve represents posterior distributions from Bayesian inference, with n = 500. (C). FSN and dFunC of ORs. Nodes represent ORs with pairwise PCCs > 0.6 based on docking scores. Edges represent node pairs with an average PCC ≥ 0.48. (D). Functional labels for the recognition of different functional group small molecules by each dFunC. Red indicates a tendency to recognize a higher number of odor molecules from a particular functional group, while blue indicates to recognize fewer. Circle size represents the degree of this tendency. (E). Correlation between four ecological parameters and species’ BBI for different VOC functional groups. pGLS tests were used for significance assessment. Red circles represent significant correlations, with size indicating the degree of significance. Gray triangles denote non-significant results. (F). Distribution of BBI values for different VOC functional groups across species with different larval diets. (G). BBI values for amine VOCs recognized by species with different larval diets. (H). 2D distribution of BBI values across species with different life habit combinations. The density distributions of different clusters along PCA1 and PCA2 are shown at the top and right. (I). Summed BBI values for all functional groups across three life habit combinations. One-way ANOVA followed by Bonferroni post-hoc tests was used for significance analysis. (J). Binary logistic regression coefficients for different life habit combinations. ****P < 0.0001

We performed molecular docking for all insect ORs using these VOCs to predict their theoretical recognition profiles (fig. S5A). A long-standing challenge in molecular docking is how to predict binding likelihood on the basis of ranking scores. To address this issue, we modeled a theoretical "hit rate" curve using docking scores and experimental datasets from Drosophila(Hallem and Carlson 2006), mosquitoes(Carey et al. 2010), and moths(de Fouchier et al. 2017) (Fig. 5B). We used the prior probability distribution in Bayesian statistics and determined the effective docking score range for insect ORs. A threshold score of -8 kcal mol-1 was established to maximize data retention while maintaining a reliable hit rate (Fig. 5B). In addition, our single-taxon hit-rate sensitivity analysis showed similar trends for Drosophila, Anopheles, and moths (fig. S5D-F). Because the locust OR repertoire had an extremely low positive rate, we did not include the locust functional data in the final threshold assessment (fig. S5 G and H, see Supplementary Text S2 for details). Although this threshold enriches experimentally responsive OR-VOC pairs, the resulting hit rate is not sufficient to support deterministic inference for individual receptor-ligand pairs. Therefore, subsequent analyses were interpreted at the OR repertoire level rather than as experimentally validated OR response spectra.

We constructed an FSN for ORs (Fig. 5C). In this network, each node represents ORs with a recognition profile Pearson correlation coefficient (PCC) greater than 0.6, and nodes with an average PCC greater than 0.48 are connected by edges. We identified 6 docking-derived functional communities (dFunCs), representing 72.5% of the ORs (fig. S5B). It should be noted that dFunCs represent docking-derived functional communities rather than experimentally confirmed functional classes. We subsequently quantified the number of VOCs with different functional groups recognized by each dFunC (fig. S5I) and assigned theoretical functional labels on the basis of their significant recognition compared with the background (Fig. 5D). The labels represent the breadth of VOC functional group recognition by each dFunC. To evaluate the accuracy of these functional labels, we calculated the area under the curve (AUC) by comparing the OR experimental data with the theoretical functional labels. For most functional group predictions, an AUC exceeding 0.75 was achieved, suggesting that these labels capture broad directional tendencies in available experimental data (fig. S5J)

dFunC1 showed broad predicted binding potential of almost all functional group VOCs, whereas dFunC6 had a limited recognition range, potentially indicating the existence of broad-tuned and narrow-tuned ORs (Fig. 5D). Additionally, dFunC3 primarily recognized aromatics, whereas dFunC5 specifically recognized nitriles. Overall, ORs had the smallest recognition range for acids and amines. Previous studies have suggested that ionotropic receptors (IRs) primarily recognize acids and amines, reducing the selective pressure on ORs to detect these VOCs(Silbering et al. 2011).

Unlike SeqC and StrC, almost all insect orders contain all dFunCs without exhibiting specificity (fig. S5K). The broad-tuned dFunC1 dominates in most orders, particularly in Psocodea, Diptera, and Lepidoptera (fig. S5L). Additionally, all the insect species presented a minimal proportion of dFunC6, indicating that a small number of narrow-tuned ORs are present across all the insects. These ORs are likely specialized to recognize specific VOCs, increasing the ability of a species to adapt to unique environmental conditions(Hallem and Carlson 2006; Carey et al. 2010; Zhang et al. 2024; Chi et al. 2025) (fig. S5L).

Associations between docking-derived OR binding breadth and insect life habits

To quantify OR promiscuity in recognizing VOCs, we developed the binding breadth index (BBI), which is based on dFunC labels. BBI is a docking-based, repertoire-level relative metric used to summarize the potential binding breadth of a set of ORs toward a given functional-group VOC category. This metric assigns positive values to ORs that bind to a greater variety of VOCs (higher BBI = broader potential olfactory breadth) and negative values to those that bind to limited VOCs (lower BBI = narrower potential olfactory breadth). Using this index, we generated docking-derived olfactory functional barcodes for different life habits in insect orders (fig. S6A), facilitating the comparison of the breadth of VOC recognition across various life habits. It should be noted that the olfactory breadth referred to in the subsequent analyses represents the species’ docking-derived potential olfactory capacity, as reflected by the full repertoire of OR genes encoded in its genome, rather than the olfactory perception exhibited at a specific developmental stage of the insect.

pGLS analysis revealed differential correlations between ecological traits and VOC class BBI patterns. Larval diet was correlated with the BBI (all except nitriles), whereas adult diet was correlated exclusively with aromatics (P= 0.038) (Fig. 5E). Habitat type was associated with nitrile/ketone perception (P= 0.0017/P= 0.003), whereas circadian rhythm was not significantly linked (Fig. 5E). Building on earlier results that the OR number was associated with insect diet (Fig. 1, D and E), suggesting that larval diet is associated with both the number and the species’ potential olfactory capacity. In contrast, the circadian rhythm had no apparent effect on either OR number or potential olfactory capacity. Insects with saprophagous larvae ranked first in the BBI for most VOC functional groups, indicating a broader VOC potential recognition capability than that of insects with other larval diets. In contrast, in insects with hematophagous larvae, the BBIs for all VOCs except nitriles were negative, suggesting that the ORs of these insects tend to bind to fewer VOCs (Fig. 5F). Additionally, insects with saprophagous larvae were the only life habit type with a positive BBI for amines (Fig. 5F, G and fig. S6B). This divergence of potential olfactory capacity correlated with expanded broad-tuned ORs (dFunC1) in insects with saprophagous larvae (fig. S6C). Ecological associations based on BBI should be interpreted as repertoire-level predicted binding trends, not as direct evidence for lifestyle-specific OR function or behavioral olfactory capacity.

Although single-parameter ecological classification revealed broad correlations between niche variables and VOC-specific BBI patterns, it failed to resolve combinatorial trait interactions shaping species-level olfactory plasticity.

To resolve combinatorial ecological effects, we defined 3 categories of potential olfactory capacity based on life habit combinations: low (Cluster 0, n=5), moderate (Cluster 1, n=23), and high (Cluster 2, n=12) responsiveness (Fig. 5, H and I). Logistic regression identified 11 key types of life habits, including enhancers (terrestrial, xylophagous and saprovorous in larvae, omnivorous and nectarivorous in adults) and suppressors (nocturnal, hematophagic, larval herbivorous, adult predatory) of potential olfactory capacity (Fig. 5J). These results support the previously established close relationship between insect OR function and ecological niches(McBride et al. 2014; Auer et al. 2020), and further provide a systematic and expanded explanation of the potential association between the olfactory breadth of insects and diverse ecological lifestyles.

Different evolutionary patterns of insect ORs across geological periods

Although the sequence evolution of the insect OR gene family and functional diversity among different orders and species have been reported multiple times(Brand and Ramirez 2017; McKenzie and Kronauer 2018; Legan et al. 2021; Chang et al. 2023), these studies reveal only a partial view rather than the overall evolutionary landscape of insect OR sequences, structures, and docking-derived functions. To elucidate the macroevolutionary trajectories of insect ORs, we integrated SeqCs, StrCs, and dFunCs. The results revealed strong congruence between SeqCs and StrCs, with both communities including nearly all dFunCs (Fig. 6 and fig. S7). This finding indicates that the evolution of OR functions follows a fundamentally different pattern from that of OR sequences and structures.

Relationships between OR sequence structure and docking-derived functional communities across insect orders.

The phylogenetic tree follows the structure from Fig.1. On the right, the mapping relationships between all insect OR SeqCs StrCs and dFunCs are shown, considering only ORs classified into communities. Time nodes for different geological periods are based on previous study(Misof et al. 2014) and are separated by gray shading. The timeline beneath the tree marks the origin of various insect orders. Animal silhouettes were obtained from PhyloPic.org.

Additionally, extant holometabolous insects showed higher OR sequence-community diversity than non-holometabolous insects (fig. S8). Because total SeqC counts may be affected by unequal species sampling and OR numbers, we added normalized analyses. First, we calculated SeqC numbers at the species level and found that holometabolous insects contained more SeqCs on average (fig. S9A). Second, because OR number was significantly correlated with SeqC count (P=0.001035), we performed equal-OR random sampling between holometabolous and non-holometabolous insects. Under the same OR sampling depth, holometabolous insects still covered more SeqCs, indicating higher OR sequence-community diversity after controlling for OR number (fig. S9B). Therefore, this result supports a genuine difference in SeqC diversity rather than a difference driven by unequal species sampling or OR number. The higher SeqC diversity observed in Holometabola was retained when the comparison was restricted to Neoptera, indicating that this pattern is not simply a restatement of previously reported Paleoptera-Neoptera differences (fig. S10).

The number of SeqCs increased with deeper divergence among insect orders, suggesting gradual accumulation of OR sequence variation during insect evolution (fig. S9C). Despite the limited SeqC diversity in non-holometabolous insects, all six types of dFunC were retained, indicating that functional community types were established early in insect evolution.

The EPME was a significant geological event that resulted in the extinction of most species and caused drastic environmental changes(Close et al. 2020; Jouault et al. 2022). In the extant species sampled here, we detected differences in docking-derived OR dFunC composition between lineages whose order-level origins fall before and after the EPME (fig. S9D). Lineages originating after the EPME showed a greater proportion of dFunC1 (broad-tuned ORs) and a lower proportion of dFunC2 (fig. S9E). This finding was confirmed by data showing that the proportion of broad-tuned ORs in Locusta is much smaller than that in Drosophila and Anopheles (Chang et al. 2023). These findings suggest that extant insect lineages originating after the EPME are characterized by more broad-tuned ORs, which are predominantly transformed from other tuned ORs.

Discussion

In this study, we developed a unified sequence–structure–docking-derived function classification system to discover insect OR evolutionary trajectories. Our results reveal class-level associations between OR repertoire variation and ecological traits and identify differences in docking-derived binding-profile composition among extant insect lineages that may be relevant to long-term macroevolutionary transitions. We further examined Orco specialization within the insect ’conserved chassis-diversified sensors’ model and identified structural features, including the EL2 beta-sheet and specialized binding-pocket properties, that may have contributed to the loss of ligand-binding function during Orco evolution. Finally, we found broad evidence at the scale of the class Insecta that adaptive evolution links species’ lifestyles to the breadth of their potential olfactory capacity.

The long-IL3 OR orthologous group includes Drosophila melanogaster DmOr13a and Bactrocera dorsalis BdorOR13a, both of which respond to 1-octen-3-ol and belong to a group of dipteran fly OR homologs associated with oviposition behavior(Kreher et al. 2008; Liu et al. 2023; Xu et al. 2023). Modeling of these homologous ORs showed that they also contain a long IL3 region. We therefore propose that long-IL3 ORs correspond to these fly homologs that recognize 1-octen-3-ol. However, 1-octen-3-ol has different behavioral functions in different species: it acts as an attractant in blood-feeding mosquitoes and tsetse flies(Hall et al. 1984; Kline et al. 2007), whereas in flies it is usually associated with oviposition regulation(Kreher et al. 2008; Liu et al. 2023). Mosquito receptors known to detect 1-octen-3-ol, such as AgOR8, AaOR8, TaOR8, and CquiOR118b(Lu et al. 2007; Xu et al. 2015; Dekel et al. 2016; Frunze et al. 2024), do not contain a long IL3 region (fig. S3 E). This suggests that the long IL3 structure in fly homologs may be related to the oviposition-associated role of 1-octen-3-ol in flies, which requires future experimental validation.

Despite differences in OR sequences and structural communities, all sampled insect orders maintain a complete set of docking-derived functional clusters. The results demonstrate stronger evolutionary coordination between OR sequences and structural types than between these features and functional types. Specifically, structural variations closely reflect sequence divergence, whereas functional distinctions show limited correlation with structural configurations. This represents a distinctive evolutionary characteristic unique to insect OR families, which allows all insect orders to contain full docking-derived functional OR repertoires, despite possessing different sequences and structural communities. We also detected differences in dFunC composition between extant lineages whose order-level origins fall before and after the EPME. Lineages originating after the EPME showed a higher proportion of broad-tuned ORs, likely influenced by multiple factors. During the Permian–Triassic transition, gymnosperms (such as cycads, lycophytes, and conifers) gradually replaced the previously dominant ferns(Grauvogel-Stamm and Ash 2005; Vajda et al. 2020; Lu et al. 2021). Plant diversification played a key role in driving insect diversification, and herbivorous insects held a central position in the Permo–Triassic interaction network(Jouault et al. 2022). The shift in plant composition led to significant changes in insect food sources and host plants. Additionally, EPME led to the extinction of 90–95% of species(Benton 2016) and was accompanied by continental reconfiguration, such as the breakup of Pangaea into Laurasia and Gondwana(Zaffos et al. 2017), resulting in numerous vacant ecological niches and driving insects to undergo adaptive radiation into new habitats. During the Triassic period, the plant community shifted(Benton 2016; Lu et al. 2021; Jouault et al. 2022), and the origin and rise of mammals(Luo 2007) and dinosaurs(Marsicano et al. 2016) led to a more complex and diverse environment. Increasing the proportion of broad-tuned ORs may increase the ability of insects to recognize a wide range of plant and animal VOCs on the basis of a limited number of OR types, thus providing a greater survival advantage for insects foraging and searching for habitats to achieve rapid occupation of new ecological niches. However, because insect ORs evolve rapidly and cross-order orthologous relationships are limited, the current dataset does not allow reliable reconstruction of complete ancestral OR repertoires at deep insect nodes. The EPME-related comparison should therefore be interpreted as a descriptive comparison among extant lineages, not as a direct test of ancestral OR functional changes before and after the EPME. This hypothesis will require further testing through broader taxon sampling, reliable orthology assignment, ancestral-state reconstruction, and functional assays.

Here, we systematically outline the evolutionary origin process of the insect OR–Orco olfactory perception model(Brand et al. 2018; Thoma et al. 2019). The ancestral insect recruited GRs to perform olfactory functions, giving rise to the earliest ORs(Robertson et al. 2003; Frank et al. 2024). These early ORs retained the homologous tetrameric pattern of GRs(Del Marmol et al. 2021). However, the sensitivity requirements of olfaction and gustation differ(de Fouchier et al. 2017; Pask et al. 2017; Slone et al. 2017; Dweck and Carlson 2023; Wang et al. 2024; Zhao et al. 2024). This issue appears to have been first addressed in the ancestor of Zygentoma by reducing the number of functional subunits while maintaining the tetrameric structure, thereby increasing sensitivity(Brand et al. 2018; Thoma et al. 2019; Wang et al. 2024; Zhao et al. 2024). We propose that the transition from OR to Orco involved duplication and selection. The number of ORs in T. domestica is eight times greater than that in M. hrabei, with no direct orthologous relationships, suggesting that OR duplication occurred in the ancestor of T. domestica. During Orco selection, at least two essential features were preserved: the β-sheet structure in the EL2 region and the extreme-property binding pocket. The β-sheet physically prevents VOCs from accessing the binding pocket, whereas the chemically restrictive nature of the binding pocket prevents VOCs from penetrating deeply. Orco exhibits strong sequence conservation, ensuring that these chemical and physical barriers are effectively preserved. Sequence similarity is much lower in the insect OR family than in the GR family, which may be related to the emergence of Orco. The conserved and stable basal structure of Orco (as a conserved chassis) reduces structural constraints during OR (as a diversified sensor) evolution, allowing ORs to evolve at a faster rate(Zhao et al. 2024). This strategy represents a unique evolutionary mechanism in the olfactory perception model, where structural conservation of chassis on the basis of Orco enables OR diversification to adapt to complex environments. However, we also identified some ORs with β-sheets in the EL2 region, which may be remnants of early Orco candidates. As insects continued to evolve, the proportion of β-sheet-containing ORs within OR repertoires gradually decreased (fig. S4, E and F), suggesting that OR evolution is a process of continuous renewal and elimination.

Insect olfaction has undergone evident adaptive evolution in response to lifestyle demands. We found that only larval feeding habits were generally correlated with the theoretical breadth of insect olfactory recognition (Fig. 5E), suggesting that, at a macroevolutionary scale, the classification of larval diets plays an important role in shaping the evolutionary trajectory of potential olfactory capacity encoded in the genome. However, this does not imply that the olfactory ability or biological importance of larvae is greater than that of adults. The actual olfactory function at any developmental stage depends on the specific types and expression levels of OR genes, which constitute only a subset of the species’ overall olfactory potential. This is biologically plausible because the larval stage constitutes a substantial portion of the insect life cycle and is critical for nutrient accumulation (Boggs 2009; Truman 2019). For example, insects with saprophagous larvae exhibit broader recognition of amines. Since the odor profile of humus is highly complex, broad-tuned ORs are required to detect a wide range of VOCs, including amines that are rarely found in other food sources(Hamana and Matsuzaki 1984; Hussain et al. 2013). This indicates that larval saprophagy may have driven the evolutionary shift of species-level olfactory potential toward the recognition of a broader range of VOCs. However, since behaviors such as host recognition and oviposition site selection are carried out by adults, the functional expression of this olfactory potential is likely to occur primarily at the adult stage. This interpretation should be treated with caution, as the saprophagous larval species currently available in our dataset are mainly concentrated in Diptera and may therefore reflect lineage effects. Future inclusion of saprophagous species from broader insect lineages will help test this association more rigorously. Regardless of the developmental stage, we found that hematophagy negatively impacts olfactory potential, likely due to highly specialized diets. Additionally, predatory behavior in adults negatively affects olfactory potential, whereas nectar feeding has a positive effect. Owing to the trade-off between vision and olfaction, predatory insects, such as mantises and dragonflies, have developed excellent vision at the cost of reduced olfactory potential (Lancer et al. 2020; Yuan et al. 2022; Kittelmann and McGregor 2024). These insects rely on strong dynamic tracking abilities, which olfaction alone cannot effectively achieve. In contrast, flowers are stationary and emit distinct, easily identifiable scent profiles, allowing nectar-feeding insects to locate their food sources precisely via olfaction. In addition to diet, terrestriality has a positive effect on the olfactory breadth of insects. Aquatic insect larvae live in water, where they encounter far fewer VOCs than terrestrial larvae do. Moreover, many aquatic insects do not feed during their adult stage, suggesting that their olfactory function may be more related to reproduction rather than foraging. It is widely believed that diurnal species rely primarily on visual cues, whereas nocturnal species prefer to invest in olfaction(Kaskan et al. 2005; Balkenius et al. 2006; Campi and Krubitzer 2010; Montgomery and Ott 2015). However, we found that nocturnality has a slight negative effect on the olfactory breadth of insects. This finding does not contradict previous conclusions but rather relates to the trade-off between sensory sensitivity (information collection) and resolution (information processing)(Stockl et al. 2016a).

Although diurnal moths generally have fewer ommatidia than nocturnal moths do, they possess more visual neuropils, which enhances resolution at the cost of sensitivity(Stockl et al. 2016a; Stockl et al. 2016b). Similarly, compared with diurnal insects, nocturnal insects allocate a greater proportion of their brain to calyces for olfactory processing(Stockl et al. 2016a). As a corresponding trade-off, their olfactory breadth is slightly reduced. In addition, this phenomenon is related to the number of VOC types. High daytime temperatures promote VOC volatilization, and a wide range of VOCs needs to be recognized during olfaction, whereas the opposite is true at night(Menchaca-Torre et al. 2015; Monard et al. 2021).

Overall, we systematically explored the relationships among sequence, structure, and docking-derived function in insect ORs, identified structural features potentially associated with Orco specialization, and provided a framework for testing how OR repertoire evolution may relate to macroenvironmental and lifestyle variation. This study provides new insights into the ecological factors driving OR evolution at the class scale. With the improvement of genome assemblies, our dataset and similarity-based community classification strategy will serve as valuable resources for future insect OR studies.

Materials and Methods

Genome data

We compiled data from all published insect genomes available in the NCBI public database as of December 2022. Genomes with a contig N50 > 1 Mb were considered reliable for OR annotation. When multiple high-quality genomes were available within a single insect family, we selected the genome with the highest assembly quality as the representative. After filtering, 107 insect genomes were included: Odonata (2), Plecoptera (2), Blattodea (1), Hemiptera (6), Psocodea (1), Hymenoptera (22), Megaloptera (1), Neuroptera (1), Coleoptera (14), Trichoptera (7), Lepidoptera (30), and Diptera (20). Detailed genome information is provided in the Data S1. To maximize phylogenetic diversity, we added eight representative species from different orders with annotated ORs (Machilis hrabei(Brand et al. 2018), Thermobia domestica(Brand et al. 2018), Ladona fulva(Brand et al. 2018), Ephemera danica(Brand et al. 2018), Locusta migratoria(Wang et al. 2015), Acyrthosiphon pisum(Robertson et al. 2019), Acromyrmex echinatior(Zhou et al. 2015), and Tribolium castaneum(RRID:SCR_010440)(Engsontia et al. 2008)). In total, our study includes 115 insect species, covering 16 orders and 111 families.

Species Phylogenies

To construct the phylogenetic tree used in this study, we merged existing phylogenies across different insect orders(Misof et al. 2014). For relationships among orders within Insecta, we referred to previously published phylogenetic frameworks. To ensure broad taxonomic representation, we included 16 insect orders in our analysis. Intra-order relationships were obtained from insectphylo.org and other published species trees covering various insect taxa(Chesters 2017). For species with available genomes that were not included in existing phylogenies, we inferred their phylogenetic placements based on family-level information. Divergence times among orders were obtained from TimeTree.org(Kumar et al. 2022) and calibrated using MCMCtree(Rannala and Yang 2007). The resulting phylogeny includes 115 insect species spanning 16 orders.

Ecological Trait Data of Insect Species

Given the vast diversity and ecological complexity of insects, and the lack of a comprehensive authoritative database, we manually retrieved ecological trait data from published literature for each species. We characterized insect ecology based on four aspects: larval diet, adult diet, habitat, and circadian rhythm. For dietary traits, we classified diets into 12 types: herbivorous, xylophagous, hematophagy, mycophagous, non-feeding, omnivorous, predatory, saprovorous, frugivorous, nectarivorous, detritivorous, and parasitic. For habitat, following previous classification criteria, species were assigned as either terrestrial or aquatic(Dijkstra et al. 2014). Insects that live entirely or have at least one life stage in water were considered aquatic; those that are strictly land-dwelling throughout their life cycle were considered terrestrial. Circadian rhythm was classified based on adult activity periods as either diurnal or nocturnal. The ecological classification and corresponding literature sources for each species are provided in the Data S3.

OR gene mining and structural filtering

All OR and Orco protein sequences were downloaded from NCBI, supplemented with ORs reported in the paper, yielding a total of 16,176 ORs. Redundancy and data cleaning were then performed using CD-HIT(Li and Godzik 2006; Fu et al. 2012) v4.8.1 to remove highly similar sequences with >90% identity, resulting in 6,029 non-redundant ORs used as queries. We used Exonerate(Slater and Birney 2005) v2.4.0 to search whole-genome data and identify potential exon regions. To improve annotation efficiency, genome contigs were split into 20 Mb segments, with 20 kb flanks added at breakpoints to avoid truncating ORs. Next, InsectOR(Karpe et al. 2021) was used for OR gene annotation.

Given that multi-exon gene annotation from genome can be error-prone, we applied the strictest structural filtering strategy to enhance annotation reliability (Fig.S1d). All annotated ORs were modeled using AlphaFold2(Jumper et al. 2021). Structural alignment against the cryo-EM structure of MhOR5 (PDB: 7LIC) was performed using Dali(Holm 2022), and ORs with Z-scores <20 were excluded. We further manually removed sequences lacking at least seven transmembrane domains. In total, 8,905 ORs with complete and credible structural models were retained for downstream analyses. This rigorous approach eliminates structurally compromised ORs that may appear intact at the sequence level, thereby increasing the likelihood of identifying ORs with functional physiological and biochemical properties.

Phylogenetic comparative analyses

The function pGLS of the R package caper was used to perform all phylogenetic generalized linear models presented in this study. Phylogenetic manipulations were conducted using the R packages phytools (RRID:SCR_015502) and ape. Phylogenetic trees were visualized by iTOL(Letunic and Bork 2024) (https://itol.embl.de/). Animal silhouettes used in this study were retrieved from http://phylopic.org/.

"Trunk-Branch" strategy

Most existing insect ORs have arisen through gene duplication events(McKenzie and Kronauer 2018; Legan et al. 2021). Initially, these duplicated OR sequences were highly similar, forming the “trunk” of core communities. Over time, mutations led ORs to diverge from their original sequences and functional features, forming the “branches” of the community (Fig. 2A). Based on this “trunk-branch” concept, we developed a strategy to construct similarity networks and define communities and the specific implementation process is as follows:

Communities with more than 20 nodes in the initial-threshold network are defined as core communities (trunks). The initial thresholds were set at e-value = 1e-68 for the SSN and Z-score = 45.9 for the StSN.

Thresholds were gradually relaxed to extend the branches of each core community. Newly appearing nodes were assigned to the community with which they shared the highest number of edges. The SSN threshold was relaxed in logarithmic steps (e.g., 1e-68, 1e-67, …), while the StSN threshold was relaxed in steps of 0.1 Z-score (e.g., 45.9, 45.8, …).

Communities were merged based on their inter-community connectivity. If the proportion of edges connecting two communities exceeded 50% for either community, they were merged. This process was repeated until all inter-community connectivities fell below the 50% threshold.

New communities with more than 20 unassigned nodes were identified and included.

Steps 2–4 were repeated until the network reached a stage of edge explosion and massive community merging.

Previously, final visualization thresholds for sequence similarity networks were often chosen based on known functional annotations. However, such information is lacking for most ORs. We found that under different thresholds, the modularity of community structures defined by the “trunk-branch” strategy showed a sharp inflection point when compared with theoretical modularity derived from the Louvain algorithm. Here, we propose a quantifiable approach for determining the final community display threshold by identifying the inflection point. Validation against known data confirmed that the resulting community partitions are highly reliable.

Sequence similarity network

First, we performed pairwise sequence alignment of all ORs using MMseqs2(Steinegger and Soding 2017) to obtain sequence similarities and e-values. To improve network readability, we applied pre-clustering with CD-HIT(Li and Godzik 2006; Fu et al. 2012) v4.8.1 (‘-c 0.6 -n 4’) to reduce redundant edges between highly similar nodes, resulting in a total of 4,947 nodes. The edge weight between nodes was defined as the average e-value of all OR pairs they represent. We then constructed the sequence similarity network based on the “trunk-branch” strategy (see "Trunk-Branch Strategy"). An e-value threshold of 1e-25 was ultimately selected as the cutoff for defining the final sequence similarity network and OR subfamilies. At this threshold, 38 OR subfamilies were identified, comprising 8,457 ORs. The network was visualized using the organic layout algorithm in Cytoscape(Shannon et al. 2003).

Structure similarity network

We performed structural modeling of all ORs using AlphaFold2(Jumper et al. 2021). Pairwise structural comparisons were then conducted using Dali(Holm 2022) to obtain Z-scores. Following an approach similar to that used in SSN, we applied a greedy clustering algorithm to pre-cluster structurally similar ORs at a Z-score threshold of 50, resulting in 3,854 nodes. The edge weight between nodes was defined as the average Z-score of all OR pairs they represent. We then constructed the structure similarity network using the “trunk-branch” strategy. A Z-score threshold of 42.9 was selected as the final cutoff for defining the StSN and OR structural subfamilies. At this threshold, 28 OR structural communities were identified, comprising 7,750 ORs. The network was visualized using the organic layout algorithm in Cytoscape(Shannon et al. 2003).

Phylogenetic tree and selection pressure analysis

To infer the amino acid phylogeny of ORs within StrC17, we performed multiple sequence alignment using MAFFT(Katoh and Standley 2013) v7.505 with the parameters ‘--globalpair --maxiterate 1000’. The alignments were trimmed using trimAl(Capella-Gutierrez et al. 2009) v1.4.rev22 with the parameter ‘-automated1’. A maximum likelihood tree was then constructed using IQ-TREE(Minh et al. 2013; Nguyen et al. 2015) v 2.2.0 with the parameters ‘-m MFP -B 1000 --bnnì. To estimate dN/dS ratios, we used the codeml program from the PAML package (RRID:SCR_014932)(Yang 1997; Yang 2007). The codeml analysis employed the branch model with the control file parameters set to ‘model = 2’, ‘NSsites = 0’. We used a custom script to extract log-likelihood values (lnL) from the output files and perform likelihood ratio tests (LRTs).

Evaluation of OR binding pocket properties

We used fpocket to identify OR binding pockets and assess their properties, with parameters set to ‘-m 3.1 -i 10’. To locate the centroid region of the binding pocket, we aligned each OR structure to MhOR5 using USalign(Zhang et al. 2022) v 20220924. Among all candidate pockets identified by fpocket(Le Guilloux et al. 2009) 4.0, the one closest to the centroid was designated as the binding pocket. Due to the stochastic nature of OR modeling, some ORs exhibited open binding pockets, making it difficult to define cavity-related parameters. To ensure consistency in the quantitative analysis, we only included ORs with closed binding pockets. Specifically, we retained only binding pockets with Number of Alpha Spheres < 140 and Volume < 800, excluding any that were open to the exterior.

Molecular dynamics simulation of Orco β-Sheet structural function

We used GROMACS(Abraham et al. 2015) to simulate the interaction between the EL2 regions of OR and Orco with VOCs. The proteins used were the cryo-EM structures of ApOrco and ApOR5(Wang et al. 2024) (PDB: 8z9z), and the selected VOC was geranyl acetate (CID: 1549026). The membrane system was rapidly constructed using the QwikMD plugin in VMD(Humphrey et al. 1996). After default minimization and annealing steps, an optimized membrane-protein system was obtained. The Caver(Jurcik et al. 2018) tool was used to identify the pocket tunnels in the cryo-EM structures of ApisOR5 and ApisOrco. Surface residues were identified using FindSurfaceResidues(https://pymolwiki.org/index.php/FindSurfaceResidues), helping define the residues surrounding the binding pocket. Geranyl acetate was positioned above the pocket, and 5 ns MD simulations were performed to evaluate interactions.

A binding state was considered potential if the following criteria were met:

- The minimum distance between geranyl acetate atoms and pocket residue atoms was <12 Å (cutoff).

- The spatial distribution of the ligand satisfied the following exclusion conditions:

- Not located on the lateral side of the protein

- Not embedded in the phospholipid bilayer

- Not sterically blocked by the EL2 of ApisOrco

Each simulation was repeated 60 times for both ApOrco and ApOR5.

It should be noted that the molecular dynamics simulations were designed to examine the early trajectories of VOCs near the extracellular entrance of the binding pocket, rather than to assess the conformational stability of the receptor complex. A duration of 5 ns is sufficient for small molecules to respond to local steric constraints.

Heterologous expression and calcium imaging

pDmelOR-mRho.V5.mER.hOr47a (RRID:Addgene_126478), a high-copy number bidirectional expression vector by inserting D. melanogaster Or47a (DmOr47a) and Orco in the pCMV-BI vector (RRID:Addgene_126475) backbone(Miazzi et al. 2019), was a gift from Bill Hansson (Addgene plasmid # 126478), then V5 tag was replaced with Flag tag, named pDmelOR-OR47a. The EL2 of OR47a (GHAEPELPFPCLFPWNIHI) in pDmelOR-OR47a was replaced by the EL2 of Orco (GDSVKMVVDHETNSSIPVEIPRLPIKSFYPWNASHG) to construct the new plasmid of pDmelOR-el2.OR47a. The sequences were confirmed by Sanger sequencing (Sangon Biotech, China).

HEK293T cells (RRID:CVCL_0063) were purchased from Wuhan Procell Biotechnology and maintained in DMEM medium (BasalMedia, China) supplemented with 10% fetal bovine serum (ExPhos, China) and 1% penicillin−streptomycin mixture (Sangon Biotech, China) at 37 °C with 5% CO2 until reaching 70–80% confluency. Cells were transfected with plasmids using LipoFiter transfection reagent (Hanbio, China). 6-8 h after transfection, transfected cells were seeded at a density of 6 × 105 cells/well into a 35 mm glass-bottom culture dish (NEST, China). The HEK293T(RRID:CVCL_0063) cell line used in this study is derived from a male human fetus. HEK293T (RRID:CVCL_0063) cells were authenticated by STR profiling (Wuhan Procell) and tested negative for mycoplasma contamination every 3 months.

Cells were harvested for calcium imaging approximately 36 hours post-transfection. The culture medium was removed, and cells were rinsed with HBSS solution (BasalMedia, China). Subsequently, cells were incubated with Fluo-8 AM (5 μM, AAT Bioquest, USA) diluted in HBSS + 0.04% pluronic acid (F127, Sangon Biotech, China) in a CO2 cell culture incubator for 30 minutes at 37°C. Following three washes with HBSS solution, the Ca2+ dependent fluorescence was recorded with a fluorescence microscope (DeltaVision Ultra, GE) with the emission wavelength set at 525 nm and the excitation wavelength at 490 nm. Experiments lasted 5 min with a sampling interval of 5 s. The stimuli—pentyl acetate (PA, 100 μM; Aladdin, China), an OR47a agonist, and VUAA1 (100 μM; Sigma-Aldrich, USA), an Orco agonist—were gently added to the culture dish at 50 s and 180 s, respectively, after the start of the imaging protocol. The recorded signals were analyzed by Fiji (RRID:SCR_002285) ImageJ v.1.54 (RRID:SCR_003070) (Schindelin et al. 2012). Levels of Ca2+ were expressed as ΔF/F0 ratios. The average fluorescence intensity of the cells during the first 30 seconds was recorded as the baseline (F0). Each assay was conducted in triplicate to ensure reproducibility. For calcium imaging experiments, data acquisition and initial ΔF/F0 ratio calculations were performed by researchers who were blinded to the treatment groups during the analysis phase. Sample sizes for calcium imaging (n = 30) were determined based on common standards in the field and prior pilot experiments to ensure sufficient statistical power for detecting changes in fluorescence intensity

Virtual docking data of ORs and VOCs

We constructed an insect-relevant odorant molecule library from four sources: pheromones(Li et al. 2023), food odors(Chi et al. 2025), fragrances(thegoodscentscompany.com), and theoretical odors(Mayhew et al. 2022). This approach ensured ecological relevance and chemical diversity. After deduplication, a total of 14,499 VOCs were retained for downstream analysis. We used Dragon (v.5.4)(Haddad et al. 2008) to calculate 32 representative molecular descriptors for each VOC, and the ChemmineR(Cao et al. 2008) R package to identify functional groups. VOCs were assigned functional group labels based on group priority rules. Dimensionality reduction of the 32 descriptors was performed using t-SNE, and the resulting two-dimensional odor space was visualized using Matplotlib(Hunter 2007). All VOC SMILES were converted to db2 format using the online resource tldr.docking.org. OR structures were aligned to the cryo-EM structure of MhOR5 (PDB:7LIC) to estimate the approximate position of the binding pocket. Virtual docking was then performed using DOCK 3.8 (RRID:SCR_000128)(Coleman et al. 2013), with batch processing handled via custom scripts ‘prepare_dock_file.sh’ and ‘run_dock_paraFly.sh’. In total, we obtained 130 million docking records between ORs and VOCs. This dataset serves as a foundational resource for predicting potential OR functions.

Hit rate curve prediction and maximum hit estimation

We first compiled published functional datasets containing explicit OR-VOC response relationships. These datasets included ORs from Drosophila melanogaster, Anopheles gambiae, Helicoverpa armigera, Spodoptera littoralis and Locusta migratoria. For each dataset, OR-VOC combinations were assigned response or non-response labels according to the original experimental results. Each OR-VOC pair was then docked using the same structural modeling, docking, and scoring workflow used in the global OR-VOC analysis. Pairs were grouped into docking-score intervals, and the hit rate for each interval was defined as the proportion of experimentally positive pairs among all pairs in that interval. Each score interval was sampled three times. To determine the docking score range at which VOCs are likely to exhibit binding activity in electrophysiological assays, we applied Bayesian statistical modeling using prior probability distributions to define the top, midpoint, random hit rate, and slope of the hit rate curve. We assumed that the relationship between docking score (energy, ei) and hit rate follows a dose-response-like curve (fig. S11 A)(Lyu et al. 2019). HitRate(ei)=top−bottom1+e−S⋅(ei−dock50)+bottom This function is defined by four parameters: (1) top: the maximum hit rate; (2) dock50: the docking score at which the hit rate is half of the top value (in kcal mol⁻¹); (3) S: the slope of the curve, calculated as S = slope × 4 / top, where slope is the percentage change in hit rate at dock50 per kcal mol⁻¹; (4) bottom: the minimum hit rate, fixed at 0. To define prior distributions, we analyzed OR hit rates for Drosophila melanogaster(Hallem and Carlson 2006), Anopheles gambiae(Carey et al. 2010), and Spodoptera littoralis(de Fouchier et al. 2017) across eight docking energy intervals and used this empirical data to derive posterior distributions of the model parameters. We specified independent Bayesian priors for each parameter: P(top) ∼ beta (α=20, β= 80), P(dock50) ∼ normal (μ = -10, σ = 15), P(S) ∼ normal (μ = -0.6, σ = 0.1) (fig. S11, B to D). Sampling from the prior distributions was performed using Hamiltonian Monte Carlo implemented in Stan, with parameters: ‘num_samples = 100000’, ‘num_chains = 4’, ‘num_warmup = 50000’, ‘max_depth = 12’.

Docking-derived functional similarity network

We first calculated the Pearson correlation coefficients (PCCs) for docking energies between all OR pairs across the VOC library. ORs were then ranked by the number of small molecules they bind, and a greedy clustering algorithm was applied to pre-cluster ORs with high functional similarity (PCCs ≥ 0.6), resulting in 6,299 nodes. The edge weight between nodes was defined as the average PCC of all OR pairs. The docking-derived functional similarity network was then constructed using the “trunk-branch” strategy. Due to the complexity of OR functional divergence, our strategy yielded a modularity value that deviated by 0.1 from the theoretical Louvain modularity. To ensure accurate community assignment, we adopted the Louvain algorithm for final community detection. A PCC threshold of 0.48 was selected as the cutoff for defining the final docking-derived functional similarity network and OR docking-derived functional communities. The network was visualized using the organic layout algorithm in Cytoscape(Shannon et al. 2003).

OR binding breadth to VOCs

We assigned a VOC recognition breadth label to each functional community. First, we counted the number of VOCs with each functional group that were recognized by ORs in each functional community. These counts were then compared against the total number of ORs recognizing those VOCs across all ORs, using Bonferroni-corrected post hoc tests. If a community showed a significantly higher or lower recognition count, it was considered to have distinct recognition capacity for that functional group. The results were normalized by the average VOC count to derive a recognition breadth score for each subfamily. Drosophila melanogaster ORs were mapped to dFunCs to evaluate the consistency between dFunC labels and experimental response data. For each VOC functional group, ORs were assigned positive or negative labels according to their dFunC labels, and these labels were compared with response counts from the Drosophila functional matrix to generate ROC curves and AUC values.

Based on the functional community labels, we further proposed a metric called binding breadth index (BBI) to quantify the potential olfactory breadth of a group of ORs toward a particular class of VOCs. To correct for sampling bias caused by small community sizes, we applied the Wilson score interval. Wilsonscore(Wk)=pk+z22n−zpk(1−pk)n+z24n21+z2n The definitions are as follows: (1) n: total number of ORs in the community; (2) pk: proportion of ORs in the kth community; (3) z: Z-score for the 95% confidence level (z = 1.96)

The BBI for the ith functional group of VOCs is defined as: BBIi=∑k=1n(Wk⋅Fki) Where: (1) Wk: Wilson score for the kth community; (2) Fki: the functional label matrix element representing the recognition score of kth community for ith functional group of VOCs; (3) n: the total number of docking-derived functional communities (n=6)

This approach provides a statistically robust framework to characterize and compare the olfactory recognition breadth of ORs.

Association between life habits and VOC recognition breadth

We calculated the overall BBI for each species based on combinations of life habits. Using BBI scores across functional groups of VOCs as input features, we performed t-SNE dimensionality reduction, followed by K-means clustering, identifying three major life habit clusters corresponding to high, moderate, and low levels of small-molecule recognition breadth. To further explore the association, we performed binary logistic regression using the R package glmnet, focusing on species in the high and low breadth categories. All life habits were encoded using one-hot encoding, and the dataset was balanced using the ROSE (Random Over Sampling Examples) method to address class imbalance. We split the dataset into training (80%) and testing (20%) sets. The training data underwent 10-fold cross-validation to determine the optimal value of the regularization parameter lambda (fig. S12, A and B). The model’s performance on the testing set was evaluated using the AUC. Finally, we fit the model using the testing data and the selected optimal lambda value, and extracted the coefficients of the predictor variables, identifying key life habits associated with VOC recognition breadth.